---
title: "Adding a new mzMl file to the database"
format:
  html:
    toc: true
    self-contained: true
author: "Ahlam Mentag, Rebecca Dauwe, Johannes Rainer"
---

## Shiny Documents

This Quarto document presents the processing workflow for a new mzML (raw) file
corresponding to a herbacetin experiment. It describes the steps required to add
the experiment to the SQL database.


##Load required libraries

```{r}
#| message: false

#' Load required libraries
library(MsExperiment)
library(Spectra)
library(xcms)
library(readxl)
library(pander)
library(pheatmap)
library(S4Vectors)
library(devtools)
library(DBI)
library(RSQLite)
library(here)
library(dplyr)
library(tidyr)
library(IRanges)
library(MSnbase)
rm(list = c("conv.CSPP_SQL"))
#` Uploading RDynLib: Load all functions of the RDynlib directory
load_all()
register(SerialParam())
```

Import the mzml file and create a corresponding sample data 
```{r}
mzml_file <- here("data","BeTio_herbacetin1327_F2T3.mzML")

sample_data_ftms <- DataFrame(
    sample_id = "S1",
    sample_name = "Sample_1"
)
```


then, we perform the centroiding
```{r}
sp <- Spectra(mzml_file)
a_c <- Spectra::pickPeaks(sp, halfWindowSize = 10L, k = 2L, threshold = 0.33,
                          msLevel. = 1L)

```

Next we export the centroided file
```{r}

export(
  a_c,
  backend = MsBackendMzR(),
  file = here("data","BeTio_herbacetin1327_F2T3_centroided.mzML"),
  format = "mzML",
  copy = FALSE
)
```

Convert the mzML file to an MSExperiment object
```{r}
mzml_file_c <- here("data","BeTio_herbacetin1327_F2T3_centroided.mzML")
ftms <- readMsExperiment(
    mzml_file_c,
    sampleData = sample_data_ftms
)
ftms
```

colors assignment for the heatmap visualization
```{r}
sampleData(ftms)
col_sample_ftms <- "blue"
col_sample_ftms_id <- c(S1 = "blue")
col_sample_ftms <- "blue"
```

Base peak chromatogram
```{r}
bpc0 <- chromatogram(ftms, aggregationFun = "max")

plot(bpc0, col = col_sample_ftms)
grid()
```

Filter by mass to get the feature range of herbacetin
```{r}
ftms <- filterSpectra(ftms, filterMzRange, mz = c(625.0, 625.4), msLevel. = 1L)

table(msLevel(spectra(ftms)))
```

Filter by rt to get the retention time range of herbacetin
```{r}
ftms <- filterSpectra(ftms, filterRt, c(690, 790))

table(msLevel(spectra(ftms)))
```

Here we remove the empty spectra 
```{r}
ftms <- filterSpectra(ftms, filterEmptySpectra)

table(msLevel(spectra(ftms)))
```

Visualizing the base peak chromatogram again:
```{r}
bpc0 <- chromatogram(ftms, aggregationFun = "max")

plot(bpc0, col = col_sample_ftms)
grid()
```

```{r}
ftms <- findChromPeaks(ftms, CentWaveParam(peakwidth = c(20, 80),
                                            integrate = 2, snthresh = 2))
```
```{r}
pks <- chromPeaks(ftms)
quantile(unname(pks[, "rtmax"] - pks[, "rtmin"]))
quantile(unname(pks[, "mzmax"] - pks[, "mzmin"]))
```
We next perform the **chromatographic peak refinement** to reduce the
number of potential centWave-specific peak detection artifacts. We
choose settings that depend on the observed peak widths above (i.e., about half
of the observed median widths).
```{r}
mnpp <- MergeNeighboringPeaksParam(expandRt = 10, expandMz = 0.001,
                                   minProp = 0.66)
ftms <- refineChromPeaks(ftms, mnpp)
```

```{r}
rt_widths <- unname(pks[, "rtmax"] - pks[, "rtmin"])
# Compute 75th percentile and multiply by 3
peakwidth <- quantile(rt_widths, 0.75) * 3
cpp <- CleanPeaksParam(maxPeakwidth = peakwidth)

ftms <- refineChromPeaks(ftms, cpp)
```

Next, we group chromPeaks
```{r}
pdp <- PeakDensityParam(sampleGroups = sampleData(ftms)$sample_id,
                        bw = 6, minFraction = 0.3, binSize = 0.1,
                        ppm = 10)

ftms <- groupChromPeaks(ftms, param = pdp)
```


```{r}
ftms_sub <- ftms
## Change the backend to in-memory backend
ftms_sub@spectra <- setBackend(spectra(ftms_sub), MsBackendMemory())
## Apply any processing steps
ftms_sub@spectra <- applyProcessing(spectra(ftms_sub))

```


```{r}
sp <- spectra(ftms_sub)
spectraData(sp)
```


```{r}
cps <- chromPeakSpectra(ftms_sub, ppm = 10)
cps$precursorMz
```

Visualizing some peaks:
```{r}
peaksData(sp)[[2]]
```

```{r}
peaksData(sp)[[126]]
```

We remove the MS1, and keep juste the MS2, MS3, and MS4 spectra:
```{r}
sps <- spectra(ftms_sub)

frag_sps <-  chromPeakSpectra(ftms_sub, msLevel = 2, ppm = 10)

#spectraData(frag_sps)$feature_id <- "FT001"

spectraData(frag_sps)

msLevel(frag_sps) |>
  table()
```

Rename the chrom_peak_id column to feature_id for more compatibility with 
the workflow
```{r}
frag_sps$feature_id <- frag_sps$chrom_peak_id
frag_sps$feature_mzmed <- frag_sps$chrom_peak_mz
frag_sps$feature_rtmed <- frag_sps$chrom_peak_rt
```


We next extract the full MSn tree for the feature:
```{r}
ftms_msn_tree <- ftms_all_levels(ftms,frag_sps)
```
Asigning an id for each tree
```{r}
ftms_msn_tree_herba <- assign_msntree_id(ftms_msn_tree)
spectraVariables(ftms_msn_tree_herba)
```
```{r}
length(unique(ftms_msn_tree_herba$MSntreeID))
ftms_msn_tree_herba$MSntreeID |>
    table() |>
    quantile()

table(table(ftms_msn_tree_herba$MSntreeID))
```
```{r}
save(ftms_msn_tree_herba, file = here("data", "ftms_msn_tree_herba.RData"))
```


Create assembled tree (pseudo MS2) by calling the assembledSpectraTree() function
from RDynlib package:

```{r}

ftms_assembled_herba <- assembleSpectraTree(ftms_msn_tree_herba)

ftms_assembled_herba
unique(spectraData(ftms_assembled_herba)$msLevel)
```

```{r}
spectraData(ftms_assembled_herba)
```
Add "not_assembled" label to the initial object with separated spectrums 
```{r}
spectraData(ftms_one_merged_tree)$spectrum.type <- "not_assembled"
```

Concatenate both spectra objects with "assembled" and "not_assembled" object.
```{r}
ftms_one_merged_tree <- setBackend(ftms_one_merged_tree,  MsBackendMemory())
ftms_assembled_herba <- setBackend(ftms_assembled_herba,  MsBackendMemory())

# Concatenate
ftms_all_herba <- concatenateSpectra(ftms_one_merged_tree, ftms_assembled_herba)

peaksData(ftms_all_herba)[[1,]]            

head(spectraData(ftms_all_herba), 3)
```



Create sql database of this feature
```{r}
dbfile <- here("data", "herba_ftneg.sqlite")
createSpectraSQLite(
    sps = ftms_all_herba,
    dbfile = dbfile,
    date = "18/6/2025",
    user = "herbacytine",
    machine = "FTMS",
    mode = "neg",
    tissue = "flax",
    mstype = NULL,
    column = NULL,
    buffera = NULL,
    bufferb = NULL,
    gradient_time = NULL,
    source = NULL,
    species = NULL,
    ce = NULL,
    meta = NULL
)
```


Add herbacytine to dynlib sql
```{r}
dbf0 <- here("data","DYL_FLAX_FTMSneg.sqlite")
db <- dbConnect(SQLite(), dbf0)
```

```{r}
dbGetQuery(db, "
  SELECT MAX(CAST(expid AS INTEGER)) AS max_expid
  FROM experiment
")
```
```{r}
# Create new database
dbf <- here("data", "DYL_herba_FTMSneg.sqlite")
con_merged <- dbConnect(RSQLite::SQLite(), dbf)

# merge
merge_databases(
  main_db   = here("data","DYL_FLAX_FTMSneg.sqlite"),
  add_db    = here("data","herba_ftneg.sqlite"),
  output_db = here("data","DYL_herba_FTMSneg.sqlite")
)
```
```{r}
dbf <- here("data", "DYL_herba_FTMSneg.sqlite")
con_merged <- dbConnect(RSQLite::SQLite(), dbf)
```



```{r}
dbGetQuery(con_merged, "
  SELECT MAX(CAST(expid AS INTEGER)) AS max_expid
  FROM experiment
")
```
```{r}
dbGetQuery(con_merged, "Select * from experiment where expid = '266' or expid = '267' or expid = '268' or expid = '269' ")
```

```{r}
add_db <- here("data","herba_ftneg.sqlite")
con_herba <- dbConnect(RSQLite::SQLite(), add_db)
dbGetQuery(con_herba, "Select * from experiment ")
```


