---
title: "Create an SQL database with spectral data, feature data, and experiment metadata"
format:
  html:
    toc: true
    self-contained: true
author:
  - name: Ahlam Mentag
  - name: Johannes Rainer
  - name: Rebecca Dauwe
execute:
  warning: false
  error: false
  message: false
editor: visual
---

## Introduction

This Quarto document describes the workflow used to create an SQL database containing spectral data, feature data, and experiment metadata for the FTMS negative-mode flax dataset using RDynlib package functions.

# Required packages

```{r}
library(DBI)
library(RSQLite)
library(dplyr)
library(RDynlib)
library(here)
library(devtools)
load_all()
```

Merge the two spectra object with two spectrum types

```{r}
# Load Spectra objects
load("data/ftms_all.RData")
dbfile <- here("data", "ftms_flax_neg_acc.sqlite")
```

Create an sql database with accessors for flax ftms spectra:

```{r}
createSpectraSQLite(
    sps = ftms_all,
    dbfile = dbfile,
    user_initials = "RD",
    date = "01/06/2015",
    user = "ament",
    machine = "LTQ_Orbitrap",
    mode = "neg",
    tissue = "Flax",
    mstype = "FT",
    column = "Waters_Sunfire_C18",
    buffera = "water_AA",
    bufferb = "ACN",
    gradient_time = "65(95-95in2_to82at32_to76at60min__to0at65minat300_40gr) min",
    source = "ESI",
    species = "Linum usitatissimum",
    ce = 35,
    meta = "Linea_varieties_S1_S9"
)
```

```{r}
con <- DBI::dbConnect(RSQLite::SQLite(), dbfile)

DBI::dbListTables(con)

```

```{r}
DBI::dbReadTable(con, "experiment")
```

```{r}
DBI::dbGetQuery(con, "SELECT COUNT(*) FROM ms_compound")
```

```{r}
DBI::dbGetQuery(con, "SELECT * FROM ms_compound LIMIT 5")
```

```{r}
DBI::dbGetQuery(con, "SELECT COUNT(*) FROM msms_spectrum")
```

```{r}
DBI::dbGetQuery(con, "SELECT * FROM msms_spectrum LIMIT 5")
```

```{r}
DBI::dbGetQuery(con, "
SELECT expid, COUNT(*) 
FROM ms_compound 
GROUP BY expid
")
```

```{r}
DBI::dbGetQuery(con, "
SELECT COUNT(*)
FROM msms_spectrum
WHERE compound_id IS NULL
")
```

```{r}
DBI::dbGetQuery(con, "
SELECT spectrum_id, COUNT(*) AS n_peaks
FROM msms_spectrum_peak
GROUP BY spectrum_id
ORDER BY n_peaks DESC
LIMIT 10
")
```

```{r}
DBI::dbGetQuery(con, "SELECT * FROM msms_spectrum_peak LIMIT 5")

```

```{r}
DBI::dbGetQuery(con, "
SELECT MIN(precursor_mz), MAX(precursor_mz)
FROM msms_spectrum
")
```

```{r}
DBI::dbGetQuery(con, "
SELECT MIN(rtime), MAX(rtime)
FROM msms_spectrum
")
```

```{r}
DBI::dbGetQuery(con, "
SELECT 
  (SELECT COUNT(*) FROM ms_compound) AS compounds,
  (SELECT COUNT(*) FROM msms_spectrum) AS spectra,
  (SELECT COUNT(*) FROM msms_spectrum_peak) AS peaks
")
```

```{r}
DBI::dbGetQuery(con, "
SELECT COUNT(*) 
FROM msms_spectrum
WHERE compound_id IS NULL
")
```

Add Feature matrix to the database

```{r}
add_FeatureMatrix_SQLite(
  dbfile = dbfile,
  xcms_exp = ftms
)
```

Verification:

```{r}
DBI::dbGetQuery(con, "
SELECT * 
FROM feature_matrix
Limit 5
")
```

Add RILs to Dynlib sql database

```{r}
# Create new database
dbf <- here("data", "DYL_FLAX_FTMSneg.sqlite")
# merge
merge_databases(
  main_db   = here("data","DyLib_FTMS_neg_or.sqlite"),
  add_db    = here("data","ftms_flax_neg.sqlite"),
  output_db = dbf
)
```

Next, we verify if the experimznt has been well added to the global database

```{r}
con_merged <- dbConnect(RSQLite::SQLite(), dbf)
```

the new experiment number of RILs is 267

```{r}
dbGetQuery(con_merged, "
  SELECT MAX(CAST(expid AS INTEGER)) AS max_expid
  FROM experiment
")
```

```{r}
dbGetQuery(con_merged, "Select * from experiment where expid = 266")
```

```{r}
dbGetQuery(con_merged, "Select * from ms_compound where expid = 266 LIMIT 10")
```

```{r}
dbGetQuery(con_merged, "Select * from msms_spectrum where compound_id = 100912")
```

```{r}
dbGetQuery(con_merged, "Select * from msms_spectrum_peak where spectrum_id = 202203")
```
