---
title: "Selecting one representative tree per feature"
format:
  html:
    toc: true
    self-contained: true
author: "Ahlam Mentag, Johannes Rainer"
editor:
  markdown:
    wrap: 72
---

# Introduction

This Quarto document presents the analysis of FTMS flaxseed data for
selecting one representative tree per feature. Preprocessing and feature
definition was performed in [Preprocessing steps](ftms_xcms_short.qmd).

# Data import

Loading all required libraries and importing the results from the data
preprocessing.

```{r}
#| message: false
library(dplyr)
library(tidyr)
library(xcms)
library(MsExperiment)
library(Spectra)
library(readxl)
library(IRanges)
library(S4Vectors)
register(SerialParam())
pth <- getwd()
pth
```

First we load the `XcmsExperiment` ftms object with the preprocessing
results.

```{r}
load(file.path(pth, "data", "ftms.RData"))
class(ftms)
```

```{r}
ftms

#convert the ftms XcmsExperiment object to a spectra object
sp <- spectra(ftms)

```

# Associate MS2 spectra to features

Next, we use `xcms::featureSpectra()` to extract all MS2 spectra for
each feature from `ftms` object.

> The `xcms::featureSpectra()` function returns a `Spectra` object.
>
> When using `msLevel = 2`, the resulting `Spectra` object contains all
> MS2 spectra whose retention time and precursor m/z fall within the
> ranges of any chromatographic peak of a feature.

```{r}
ms2 <- featureSpectra(dropFilledChromPeaks(ftms), msLevel = 2, skipFilled = TRUE)

msLevel(ms2) |>
  table()
```

it returns `r length(ms2)` MS level 2 spectra. The total number of
features for which an MS2 spectrum was identified is:

```{r}
length(unique(ms2$feature_id))
length(ms2$feature_id)
```

Here we print the initial number of MS levels.

```{r}
s <- spectra(ftms)
table(msLevel(s))
```

In total, there were `r table(msLevel(s))[2]` MS2 spectra. This means
that
`r round((table(msLevel(s))[2]-length(ms2$feature_id))*100/table(msLevel(s))[2], 1)`
% of the MS2 spectra could not be assigned to any chromatographic peak.

The percentage of features with at least one MS2:

```{r}
# Total number of features in ftms
n_total_features <- nrow(featureDefinitions(ftms))

# Features with ≥1 MS2 (those returned by featureSpectra)
n_features_with_ms2 <- length(unique(ms2$feature_id))

# Percentage
perc_features_with_ms2 <- (n_features_with_ms2 / n_total_features) * 100
perc_features_with_ms2

```

The total number of features with at least one MS2 is
`r n_features_with_ms2`, which corresponds to
`r format(perc_features_with_ms2, digits = 3)`%.

A summary on the number of MS2 spectra assigned to a feature is shown
below:

```{r}
table(ms2$feature_id) |>
    quantile()
```

The median number of MS2 spectra per feature (across all data files) is
hence `r median(table(ms2$feature_id))`.

the percentage of MS2 spectra that are assigned to a feature

```{r}
# Total number of MS2 spectra in the dataset
n_total_ms2 <- sum(msLevel(s) == 2)
sp <- spectraData(ms2)
sp$groups <- paste0(sp$dataOrigin, "_", sp$scanIndex)
# Number of MS2 spectra assigned to features
n_assigned_ms2 <- length(unique(sp$groups))

# Percentage assigned
perc_ms2_assigned <- (n_assigned_ms2 / n_total_ms2) * 100
perc_ms2_assigned

```

The total number of assigned MS2 spectra is `r n_assigned_ms2`, which
corresponds to a percentage of
`r format(perc_ms2_assigned, digits = 3)`.

Note that an MS2 spectrum could also be assigned to more than one
feature. The total number of MS2 spectra assigned to more than one
feature is:

```{r}
ms2_id <- paste0(ms2$dataOrigin, "_", ms2$acquisitionNum)
ms2_table <- table(ms2_id)
ms2_mult <- names(ms2_table)[ms2_table > 1]

```

Of the `r length(unique(ms2_id))` MS2 spectra, `r length(ms2_mult)` are
assigned to more than one feature.

# Extract all MS2-MSn spectra associated to features

We next define a function to extract the full MSn tree for features.
This function takes an `XcmsExperiment` object as input, extracts the
MS2 spectra using `xcms::featureSpectra()` function, and then retrieves
all MS3 (and subsequently MS4 spectra) for all these MS2 spectra using
the *precScanNum* spectra variable. The `ftms_all_levels` function
returns a combined `Spectra` object containing all MS2, MS3, and MS4
spectra, with each spectrum linked to a feature via a propagated
`feature_id` from the originally matched MS2 spectrum.

```{r}
ftms_all_levels <- function(ftms, ms2) {

  #ms2 <- featureSpectra(ftms, msLevel = 2)
  cat("Number of MS2 spectra found:", length(ms2), "\n")

  res <- lapply(unique(ms2$dataOrigin), function(origin) {
    ms2_subset <- filterDataOrigin(ms2, origin)

    ms3_filtered <- filterDataOrigin(filterMsLevel(spectra(ftms), 3), origin)
    ## To support n:m matches
    m <- findMatches(ms2_subset$acquisitionNum, ms3_filtered$precScanNum)
    cat("Number of MS3 matched to MS2:", length(m), "\n")
    ms3_filtered <- ms3_filtered[to(m)]
    ms3_filtered$feature_id <- ms2_subset$feature_id[from(m)]

    ms4_filtered <- filterDataOrigin(filterMsLevel(spectra(ftms), 4), origin)
    m <- findMatches(ms3_filtered$acquisitionNum, ms4_filtered$precScanNum)
    cat("Number of MS4 matched to MS3:", length(m), "\n")
    ms4_filtered <- ms4_filtered[to(m)]
    ms4_filtered$feature_id <- ms3_filtered$feature_id[from(m)]

    # Combine spectra for this origin
    c(ms2_subset, ms3_filtered, ms4_filtered)
  })
    do.call(c, res)
}

ftms_msn_tree <- ftms_all_levels(ftms,ms2)

table(msLevel(ftms_msn_tree))
```

The number of spectra from different MS levels is shown below:

```{r}
table(msLevel(ftms_msn_tree))
```

# Grouping MSn spectra into fragmentation trees

We now group all spectra that belong to the same fragmentation tree and
record the precursor m/z of the originating MS2 spectrum for each
spectrum within that tree.

The function `assign_msntree_id()` will:

-   Take a `Spectra` object `x` containing MS2–MSn spectra\

-   Assign tree IDs to each spectrum in `x`, based on precursor
    relationships within each sample (`dataOrigin`)\

-   Ensure that MS2 spectra always start a new tree, and that MS3/MS4
    spectra inherit the tree ID from their parent\

-   Propagate the MS2-level precursor m/z to all downstream spectra in
    the same tree (MS3, MS4, …), with `NA` for MS2 spectra\

-   Return the updated `Spectra` object with two new spectra variables:

    -   **`MSntreeID`** — integer ID of the fragmentation tree each
        spectrum belongs to\
    -   **`precursorMz.MS2`** — MS2-level precursor m/z propagated to
        MS3/MS4 (and beyond), `NA` for MS2

```{r}
assign_msntree_id <- function(x) {
  ## Order x by MS level: ensures we first assign an ID to the MS2.
  o <- order(msLevel(x))
  x <- x[o]
  scan_index <- scanIndex(x)
  prec_scan_num <- precScanNum(x)
  ms_level <- msLevel(x)
  data_origin <- dataOrigin(x)
  
  ## store precursorMz values
  precursor_mz <- precursorMz(x)

  tree_id <- rep(NA_integer_, length(scan_index))
  ## initialize output vector for MS2-level precursor m/z
  precursor_mz_MS2 <- rep(NA_real_, length(scan_index))
  
  global_counter <- 0

  for (origin in unique(data_origin)) {
    idx <- which(data_origin == origin)
    origin_scan_index <- scan_index[idx]
    origin_prec_scan_num <- prec_scan_num[idx]
    origin_ms_level <- ms_level[idx]
    ## per-origin precursorMz vector
    origin_precursor_mz <- precursor_mz[idx]
    origin_tree_id <- rep(NA_integer_, length(idx))
    ## per-origin precursorMz.MS2 vector
    origin_precursor_mz_MS2 <- rep(NA_real_, length(idx))

    scan_idx_map <- setNames(seq_along(origin_scan_index), origin_scan_index)

    for (i in seq_along(idx)) {
      level <- origin_ms_level[i]
      if (level == 2) {
        ## If MS2 hasn't been assigned yet
        if (is.na(origin_tree_id[i])) {
          global_counter <- global_counter + 1
          origin_tree_id[i] <- global_counter
        }
        ## MS2 spectra have no MS2-level precursor
        origin_precursor_mz_MS2[i] <- NA_real_

      } else {
        ## Assign ID of the related precursor ID
        parent_scan <- origin_prec_scan_num[i]
                parent_pos <- which(origin_scan_index == parent_scan)
        if (length(parent_pos) == 1 && origin_ms_level[parent_pos] == (level - 1)) {
          origin_tree_id[i] <- origin_tree_id[parent_pos]

          ## assign precursorMz.MS2 depending on depth
          if (level == 3) {
            origin_precursor_mz_MS2[i] <- origin_precursor_mz[i]
          } else {
            ancestor_pos <- parent_pos
            while (!is.na(ancestor_pos) && origin_ms_level[ancestor_pos] > 3) {
              ancestor_scan <- origin_prec_scan_num[ancestor_pos]
              ancestor_pos <- scan_idx_map[as.character(ancestor_scan)]
            }
            if (!is.na(ancestor_pos) && origin_ms_level[ancestor_pos] == 3) {
              origin_precursor_mz_MS2[i] <- origin_precursor_mz[ancestor_pos]
            } else {
              origin_precursor_mz_MS2[i] <- NA_real_
            }
          }

        } else {
          stop("MS", level, " with ", length(parent_pos), " precursors")
        }
      }
    }
    tree_id[idx] <- origin_tree_id
    ## store per-origin results
    precursor_mz_MS2[idx] <- origin_precursor_mz_MS2
  }

  ## Return in original order
  tree_id <- tree_id[order(o)]
  ## reorder new variable to match
  precursor_mz_MS2 <- precursor_mz_MS2[order(o)]
  
  x <- x[order(o)]

  ## add both to Spectra object
  x$MSntreeID <- tree_id
  x$precursorMz.MS2 <- precursor_mz_MS2

  return(x)
}


ftms_msn_tree <- assign_msntree_id(ftms_msn_tree)

spectraVariables(ftms_msn_tree)

```

```{r}
length(unique(ftms_msn_tree$MSntreeID))
```

There are in total `r length(unique(ftms_msn_tree$MSntreeID))` spectral
(fragmentation) trees for in total
`r length(unique(ftms_msn_tree$feature_id))` features. The distribution
of the lengths of the fragmentation trees is

```{r}
ftms_msn_tree$MSntreeID |>
    table() |>
    quantile()

table(table(ftms_msn_tree$MSntreeID))
```

Fragmentation trees have maximally 4 spectra. This is in accordance with
the setup of the ion trap instrument: The two most intense first order
product ions were refragmented (if their intensity was above a
threshold), resulting in maximally two MS3 spectra, and the most intense
second order product ion of the most intense first order product ion was
also refragmented (if it's intensity was above a threshold), resulting
in maximally one MS4 spectrum. So, a fragmentation tree can maximally
consist of one MS2 spectrum + two MS3 spectra + one MS4 spectrum = four
spectra.

```{r}
## we stored the tree ids and the msLevels to a dataframe
df <- spectraData(ftms_msn_tree, c("MSntreeID", "msLevel")) |>
    as.data.frame()

## here we take count the last level of each tree
library(dplyr)
max_level_per_tree <- df %>%
  group_by(MSntreeID) %>%
  summarise(max_level = max(msLevel), .groups = "drop")


print(t <- table(max_level_per_tree$max_level))

```

The number of trees that go up to MS2 level is : `r t[1]`. The number of
trees that go up to MS3 level is : `r t[2]`. The number of trees that go
up to MS4 level is : `r t[3]`.

# Select single best MSn spectral tree per feature

Next we select one most representative tree per feature based on (1) the
intensity of the precursor, (2) the length of the tree, (3) the total
number of peaks in all spectra of the tree, and (4) in rare cases of
several trees with maximal number of peaks: break ties based on
precursor intensity (max)

## (1) Filter out weak precursor trees (lowest precursor intensity)

Keep only trees (`vMSntreeID`v) whose `msLevel == 2` spectrum precursor
intensity is in the top 75%.

```{r}
sd <- as.data.frame(spectraData(ftms_msn_tree))
ms2 <- subset(sd, msLevel == 2)
threshold <- quantile(ms2$precursorIntensity, 0.25, na.rm = TRUE)
strong_MSntrees <- unique(ms2$MSntreeID[ms2$precursorIntensity > threshold])
sd_filtered <- subset(sd, MSntreeID %in% strong_MSntrees)

length(unique(sd_filtered$feature_id))
length(unique(sd$feature_id)) 
removed.features <- length(unique(sd$feature_id)) - length(unique(sd_filtered$feature_id))
message(removed.features, " features were removed after selecting only spectra with top 75% precursor intensity.")
message(length(unique(sd_filtered$feature_id))," of the ", length(unique(sd$feature_id)), " features remain.")

```

## (2) Among 75% remaining MSn trees, select the longest tree per feature
### Remove duplicates after rounding of precursorMz
MS2, MS3, and MS4 spectra have unit resolution. This means that, within the same dataOrigin, MS3 spectra with the same round(precursorMz), and same precursorIntensity, are duplicates. Of these duplicates we only retain the MS3 spectrum with the highest value for peaksCount
```{r}
# Round precursor m/z for unit resolution matching
sd_filtered <- sd_filtered %>%
  mutate(precursorMz_round = round(precursorMz))

# Identify duplicates only among MS3 spectra
ms3_filtered <- sd_filtered %>%
  filter(msLevel == 3)

# Deduplicate: within each dataOrigin, precursorMz_round, precursorIntensity group
ms3_dedup <- ms3_filtered %>%
  group_by(dataOrigin, precursorMz_round, precursorIntensity) %>%
  slice_max(order_by = peaksCount, n = 1, with_ties = FALSE) %>%
  ungroup()

# Replace MS3 spectra in the full dataset with deduplicated ones
sd_filtered_dedup <- bind_rows(
  sd_filtered %>% filter(msLevel != 3),
  ms3_dedup
)

# message summary
removed <- nrow(sd_filtered) - nrow(sd_filtered_dedup)
message(removed, " duplicate MS3 spectra removed after rounding and deduplication.")

sd_filtered <- sd_filtered_dedup
```
### Select longest tree per feature

```{r}
# Get max msLevel per tree
tree_length <- aggregate(msLevel ~ MSntreeID + feature_id, data = sd_filtered, max)

# Get max depth per feature
max_len <- aggregate(msLevel ~ feature_id, data = tree_length, max)

# Keep only trees with the maximum length for their feature
longest_trees <- merge(tree_length, max_len, by = c("feature_id", "msLevel"))

```

## (3) select the tree with highest total number of peaks in all levels

total number of peaks: sum of peaksCount per tree_id in spectraData

```{r}
tree_peaks <- aggregate(peaksCount ~ MSntreeID, data = sd_filtered, sum)

candidate_trees <- merge(longest_trees, tree_peaks, by = "MSntreeID")
max_peaks <- aggregate(peaksCount ~ feature_id, data = candidate_trees, max)
candidate_trees <- merge(candidate_trees, max_peaks, by.x = c("feature_id", "peaksCount"), by.y = c("feature_id", "peaksCount"))

```

## (4) break ties by precursor intensity (highest)

```{r}
# Compute maximal precursor intensity of MS2 spectra per tree
prec_max <- aggregate(precursorIntensity ~ MSntreeID, data = ms2, max)
candidate_trees <- merge(candidate_trees, prec_max, by = "MSntreeID")

# select the tree with highest precursor intensity per feature
candidate_trees <- candidate_trees[order(candidate_trees$feature_id, -candidate_trees$precursorIntensity), ]
best_tree <- candidate_trees[!duplicated(candidate_trees$feature_id), c("feature_id", "MSntreeID")]
```

## (5) filter Spectra object

```{r}
best_MSntreeIDs <- best_tree$MSntreeID
ftms_msn_tree_best <- ftms_msn_tree[ which(spectraData(ftms_msn_tree)$MSntreeID %in% best_MSntreeIDs) ]
```


## (Optional) evaluate based on known spectra

```{r}
idx <- which(round(ftms_msn_tree_best$precursorMz) == 1333)
idx
peaksData(ftms_msn_tree_best)[idx][[1]]


idx <- which(round(ftms_msn_tree_best$precursorMz) == 1171)
idx
peaksData(ftms_msn_tree_best)[idx][[1]]
peaksData(ftms_msn_tree_best)[idx][[2]]
```

This looks like a really good selection !
## (6) Save
Now we save the resulting ftms spectra object to a local directory.

```{r}
ftms_one_tree <- ftms_msn_tree_best
save(ftms_one_tree, file = file.path(pth, "data", "ftms_one_tree.RData"))
```

# Add additional MS3 and MS4 spectra, with different precursorMz, to the selected trees

Here, the purpose is to enrich the selected “best” MSn trees
(`ftms_msn_tree_best`) with additional MS3-MS4 spectra from other MS2
and/or MS3 precursors belonging to the same feature, to build a more
complete merged fragmentation tree. Add, from the "strong" spectral
trees with MS2 precursor intensity in top 75 %, MS3 spectra with
different MS2 precursorMz, and MS4 spectra with different MS3 or MS4
precursorMz, to the selected `ftms_msn_tree_best`. Create new integer
variable: `ftms_msn_tree_best$MergedMSntreeID`. All spectra belonging to
the same merged tree (`ftms_msn_tree_best` + added MS3-MS4) will share
the same `MergedMSntreeID`. Each feature will have 1 merged tree ID.

## MS3-MS4 spectra from other MS2 precursor

Add within each MSntree of `ftms_msn_tree_best`, MS3 spectra, from the
75% "strong" trees `sd_filtered`, belonging to same feature but having
different `round(precurosrMz)` : Select longest tree (meaning MS3 with
depending MS4 in the same MSntree_id) from these, select highest total
peakCounts if ties, select highest precursor Intensity of MS3 spectrum

### Create selectbest.ms3ms4() function:

This function selects best MS3 spectra to be added to merged MSntree,
and appends any dependent MS4 spectra from the same feature

```{r}
# Input: a dataframe with spectraData of all MS3 spectra to be added to the MSntree of a specific feature.
# Computes total number of peaks (pc), optionally including MS4 peaks of dependent daughter MS4 spectrum.
# Selects the best MS3 candidate based on pc score and precursor intensity.
# Appends dependent MS4 spectrum row from sd_filtered.


  selectbest.ms3ms4 <- function(x) {
  # Filter MS4 if any exist, compute pc, sort, keep top MS3
  top_ms3 <- x %>%
    {
      if (any(!is.na(.$ms4_acq))) {
        dplyr::filter(., !is.na(ms4_acq))
      } else {
        .
      }
    } %>%
    dplyr::mutate(pc = rowSums(cbind(peaksCount, peaksCount.MS4), na.rm = TRUE)) %>%
    dplyr::arrange(desc(pc), desc(precursorIntensity)) %>%
    dplyr::slice_head(n = 1)

  # Add corresponding MS4 spectra, if ms4_acq exists
  if (!all(is.na(top_ms3$ms4_acq))) {
    ms4_row <- sd_filtered %>%
  filter(msLevel == 4, acquisitionNum %in% top_ms3$ms4_acq, feature_id == top_ms3$feature_id)
combined <- bind_rows(top_ms3, ms4_row)
    
    
  } else {
    combined <- top_ms3
  }

  combined
} 
```

### Prepare merged MSntree dataset

```{r}
# Extract spectraData from ftms_msn_tree_best
sd_best <- as.data.frame(spectraData(ftms_msn_tree_best))

sd_filtered$precMzRound <- round(sd_filtered$precursorMz)
sd_best$precMzRound <- round(sd_best$precursorMz)

# Extract MS4 spectra from sd_filtered
# ms3_acq: acquisition number of the parent MS3 spectrum
# ms4_acq: acquisition number of the MS4 spectrum
# peaksCount.MS4: peaksCOunt of MS4 spectrum
ms4_spectra <- sd_filtered %>%
  filter(msLevel == 4) %>%
  select(MSntreeID, precScanNum, acquisitionNum, peaksCount) %>%
  dplyr::rename(ms3_acq = precScanNum, ms4_acq = acquisitionNum, peaksCount.MS4 = peaksCount)


# Initialize empty list to store additional MS3
all_additional_ms3 <- list()

for (f in unique(sd_best$feature_id)) {
  
  # Best MSn tree for this feature
  best_tree <- sd_best %>% filter(feature_id == f)
  best_precMz <- unique(best_tree$precMzRound[best_tree$msLevel == 3])
  
  # Candidate MS3 spectra with different precursor m/z
  ms3_candidates <- sd_filtered %>%
    filter(feature_id == f,
           msLevel == 3,
           !(round(precursorMz) %in% best_precMz))
  
  # Only continue if there are candidates
  if (nrow(ms3_candidates) != 0) {
    print(paste(f,"has", nrow(ms3_candidates), "additional MS3 candidates.", sep=" " ))

    # Join MS3 candidates to their dependent MS4 (if exists)
    ms3_candidates <- ms3_candidates %>%
      left_join(ms4_spectra, by = c("MSntreeID" = "MSntreeID", "acquisitionNum" = "ms3_acq"))
    
  
    # Group by precMzRound, select best MS3, and append MS4

    ms3.p.precmz <- split(ms3_candidates, ms3_candidates$precMzRound)
    additional.ms3 <- lapply(ms3.p.precmz, selectbest.ms3ms4)
    all_additional_ms3[[f]] <- do.call(rbind, additional.ms3)
  }
}

# Bind all features into a single data frame
all_additional_ms3 <- do.call(rbind, all_additional_ms3)
row.names(all_additional_ms3) <- NULL



# merge the additional MS3 spectra to the selected best trees
common_cols <- intersect(names(sd_best), names(all_additional_ms3))
merged <- bind_rows(
  sd_best[, common_cols],
  all_additional_ms3[, common_cols]
)  

# create new unique merged tree ID 
merged <- merged %>%
  mutate(MergedMSntreeID = as.integer(factor(feature_id)))

# Extract and update the corresponding spectra from the original MSn tree object
merged_spectra <- ftms_msn_tree[
  paste(spectraData(ftms_msn_tree)$dataOrigin,
        spectraData(ftms_msn_tree)$scanIndex, sep = "_") %in%
    paste(merged$dataOrigin, merged$scanIndex, sep = "_")
]

spectraData(merged_spectra) <- S4Vectors::DataFrame(merged)

```
### Check for duplicates and post-processing if duplicates found

```{r}
# Extract metadata
merged_sd <- as.data.frame(spectraData(merged_spectra))

# Identify groups with same feature_id, msLevel, precursorMz, precursorMz.MS2
merged_sd_MS4 <- merged_sd[merged_sd$msLevel == 4,]
merged_sd_MS4$precMz.MS2.Round <- round(merged_sd_MS4$precursorMz.MS2)
dup_groups <- merged_sd_MS4 %>%
  group_by(feature_id, msLevel, precMzRound, precMz.MS2.Round) %>%
  mutate(n_in_group = n()) %>%
  ungroup() %>%
  filter(n_in_group > 1)

# Report duplicates
if (nrow(dup_groups) > 0) {
  message("Found ", nrow(dup_groups), " spectra in duplicate groups:")
dup_groups_ordered <- dup_groups %>%
    arrange(feature_id)
  
  print(dup_groups_ordered[, c("feature_id","msLevel","precMz.MS2.Round","precMzRound","peaksCount")])} else {
  message("No duplicate MS4 spectra found.")
}
```

## MS4 spectra from other `round(`precursor_mz_MS2)\`

Identify “new” MS4 spectra in sd_filtered that aren’t already in merged
trees, but are connected to existing MS3 spectra via matching precursor
m/z values, and also have precursor m/z matching actual fragment peaks
of those MS3 spectra.

###Create set of candidate new MS4 spectra

```{r}
# Prepare subsets of the merged spectra
merged_sd <- as.data.frame(spectraData(merged_spectra))

ms3_merged <- merged_sd %>% filter(msLevel == 3)
ms4_merged <- merged_sd %>% filter(msLevel == 4)

# Select candidate MS4 spectra from sd_filtered :
# their MS3 parent spectrum exists in merged_spectra (to verify based on presence of their precursorMz.MS2 in ms3_merged$precursorMz)
# precursorMz.MS2 is not already present among MS4 spectra of merged_spectra
ms4_candidates <- sd_filtered %>%
  filter(
    msLevel == 4,
    !(precursorMz.MS2 %in% ms4_merged$precursorMz.MS2),
    precursorMz.MS2 %in% ms3_merged$precursorMz
  )
```

### Retain only cadidate MS4 spectra for which the precursor m/z (precursorMz) is present among the peaks of its corresponding MS3 parent spectrum in the merged_spectra object.

```{r}
# Identify MS3 spectra in best or merged trees, that could be parents of the candidate new MS4 spectra
parent_ms3_precursors <- unique(ms4_candidates$precursorMz.MS2)
ms3_parents_sd <- ms3_merged %>%
  filter(precursorMz %in% parent_ms3_precursors)
ms3_parents <- paste(ms3_parents_sd$dataOrigin,ms3_parents_sd$scanIndex)

# Extract peaksData from merged_spectra only for relevant MS3 spectra
pd <- peaksData(merged_spectra)
sd <- spectraData(merged_spectra)

spectra_id <- paste(sd$dataOrigin,sd$scanIndex)

ms3_indices <- which(spectra_id %in% ms3_parents)
peaks.ms3parents.list <- as.list(pd[ms3_indices,])

### filter  ms4_candidates so that each candidate MS4 spectrum is retained only if its precursor m/z (precursorMz) is present among the peaks of its corresponding MS3 parent spectrum.

# Build lookup of rounded parent precursorMz → rounded peaks
parent_lookup <- lapply(peaks.ms3parents.list, function(x) round(x[, 1]))
names(parent_lookup) <- as.character(round(ms3_parents_sd$precursorMz))

# Round MS4 precursor columns for matching
ms4_candidates <- ms4_candidates %>%
  mutate(
    precursorMz_round = round(precursorMz),
    precursorMz_MS2_round = round(precursorMz.MS2)
  )

# For each candidate, check if its rounded precursor is in the parent’s rounded peaks
keep_idx <- logical(nrow(ms4_candidates))

for (i in seq_len(nrow(ms4_candidates))) {
  parent_key <- as.character(ms4_candidates$precursorMz_MS2_round[i])
  if (parent_key %in% names(parent_lookup)) {
    keep_idx[i] <- ms4_candidates$precursorMz_round[i] %in% parent_lookup[[parent_key]]
  } else {
    keep_idx[i] <- FALSE
  }
}

# Filter retained candidates
ms4_candidates_filtered <- ms4_candidates[keep_idx, ]
```

### Check for duplicate MS4 spectra and remove if necessary
```{r}
# Prepare merged_spectra metadata for comparison
merged_sd <- as.data.frame(spectraData(merged_spectra)) %>%
  mutate(
    precursorMz_round = round(precursorMz),
    precursorMz_MS2_round = round(precursorMz.MS2)
  )

# Build unique composite keys for existing merged spectra and MS4 candidates
existing_keys <- paste(
  merged_sd$feature_id,
  merged_sd$msLevel,
  merged_sd$precursorMz_round,
  merged_sd$precursorMz_MS2_round,
  sep = "_"
)

candidate_keys <- paste(
  ms4_candidates_filtered$feature_id,
  ms4_candidates_filtered$msLevel,
  ms4_candidates_filtered$precursorMz_round,
  ms4_candidates_filtered$precursorMz_MS2_round,
  sep = "_"
)

# Retain only candidates whose key is NOT already in merged_spectra
keep_new_idx <- !(candidate_keys %in% existing_keys)
ms4_candidates_filtered <- ms4_candidates_filtered[keep_new_idx, ]

message(sum(!keep_new_idx), " candidate MS4 spectra removed because they already exist in merged_spectra.")
message(nrow(ms4_candidates_filtered), " unique new MS4 spectra retained.")
```

### Select MS4 candidate spectrum with highest number of peaks

```{r}
# Create composite key precMS2.precMS3
ms4_candidates_filtered <- ms4_candidates_filtered %>%
  mutate(
    ft.precMS2.precMS3 = paste0(
      feature_id,
      "precMS2_", round(precursorMz.MS2),
      "_precMS3_", round(precursorMz)
    )
  )

group_sizes <- ms4_candidates_filtered %>%
  count(ft.precMS2.precMS3)


# Keep only MS4 spectrum with the highest number of peaks per unique ft.precMS2.precMS3
if (all(group_sizes$n == 1)) {
  message("Each ft.precMS2.precMS3 group has only one row — skipping slice_max() step.")
  newms4_best <- ms4_candidates_filtered
} else {
  newms4_best <- ms4_candidates_filtered %>%
    group_by(ft.precMS2.precMS3) %>%
    slice_max(order_by = peaksCount, n = 1, with_ties = FALSE) %>%
    ungroup()
}
```

### Append new MS4 spectra (newms4_best) to merged MSntree dataset

```{r}
# Prepare metadata for merging
sd_merged <- as.data.frame(spectraData(merged_spectra))

# Round precursor m/z for matching consistency
sd_merged$precMzRound <- round(sd_merged$precursorMz)
newms4_best$precMzRound <- round(newms4_best$precursorMz)

# Create merge keys (for later extraction from original dataset)
mergeKey_existing <- paste(sd_merged$dataOrigin, sd_merged$scanIndex, sep = "_")
mergeKey_newms4   <- paste(newms4_best$dataOrigin, newms4_best$scanIndex, sep = "_")

# Combine metadata tables
common_cols <- intersect(names(sd_merged), names(newms4_best))

merged_df <- bind_rows(
  sd_merged[, common_cols],
  newms4_best[, common_cols]
)

# Assign new merged tree IDs
merged_df <- merged_df %>%
  mutate(MergedMSntreeID = as.integer(factor(feature_id)))

# Update Spectra object
merged_spectra <- ftms_msn_tree[
  paste(spectraData(ftms_msn_tree)$dataOrigin,
        spectraData(ftms_msn_tree)$scanIndex, sep = "_") %in%
    c(mergeKey_existing, mergeKey_newms4)
]

spectraData(merged_spectra) <- S4Vectors::DataFrame(merged_df)

```

### Check for duplicates (and post-processing if duplicates found)

```{r}
# Extract metadata
merged_sd <- as.data.frame(spectraData(merged_spectra))

# Identify MS4 spectra with same feature_id, msLevel, precursorMz, precursorMz.MS2
merged_sd_MS4 <- merged_sd[merged_sd$msLevel == 4,]
merged_sd_MS4$precMz.MS2.Round <- round(merged_sd_MS4$precursorMz.MS2)
dup_groups <- merged_sd_MS4 %>%
  group_by(feature_id, msLevel, precMzRound, precMz.MS2.Round) %>%
  mutate(n_in_group = n()) %>%
  ungroup() %>%
  filter(n_in_group > 1)

# Report duplicates
if (nrow(dup_groups) > 0) {
  message("Found ", nrow(dup_groups), " spectra in duplicate groups:")
dup_groups_ordered <- dup_groups %>%
    arrange(feature_id)
  
  print(dup_groups_ordered[, c("feature_id","msLevel","precMzRound","precMz.MS2.Round","peaksCount")])} else {
  message("No duplicate spectra found.")
}
```

## MS4 spectra from same `round(`precursor_mz_MS2)`, but different`round(precursorMz)\`.

Add to merged_spectra the MS4 spectra from sd_filtered such that:

(1) Their parent MS3 already exists in merged_spectra (precursorMz.MS2 %in%
merged_spectra[msLevel == 3]$precursorMz)

(2) Their composite key (precMS2.precMS3) is not already present among
existing MS4 spectra in merged_spectra

(3) Their precursor m/z (round(precursorMz)) is found among the rounded
peaks of their corresponding MS3 parent spectrum in merged_spectra.

### Create set of candidate new MS4 spectra
```{r}
# Create candidate new MS4 spectra
merged_sd <- as.data.frame(spectraData(merged_spectra))

ms3_merged <- merged_sd %>% filter(msLevel == 3)
ms4_merged <- merged_sd %>% filter(msLevel == 4)

# Round precursor m/z for matching
ms4_merged$precursorMz_round <- round(ms4_merged$precursorMz)
ms4_merged$precursorMz_MS2_round <- round(ms4_merged$precursorMz.MS2)
ms4_merged$precMS2.precMS3 <- paste0(
  "precMS2_", ms4_merged$precursorMz_MS2_round,
  "_precMS3_", ms4_merged$precursorMz_round
)

# Select MS4 candidates from sd_filtered
ms4_candidates <- sd_filtered %>%
  filter(
    msLevel == 4,
    precursorMz.MS2 %in% ms3_merged$precursorMz,                                # parent MS3 exists
    !(paste0("precMS2_", round(precursorMz.MS2), "_precMS3_", round(precursorMz))
      %in% ms4_merged$precMS2.precMS3)                                          # composite key not already present
  )
```

### Retain only candidates whose precursor m/z is present among parent MS3 peaks
```{r}
# Identify parent MS3 spectra
parent_ms3_precursors <- unique(ms4_candidates$precursorMz.MS2)
ms3_parents_sd <- ms3_merged %>%
  filter(precursorMz %in% parent_ms3_precursors)
ms3_parents <- paste(ms3_parents_sd$dataOrigin, ms3_parents_sd$scanIndex)

# Extract peaksData for relevant MS3 spectra
pd <- peaksData(merged_spectra)
sd <- spectraData(merged_spectra)
spectra_id <- paste(sd$dataOrigin, sd$scanIndex)

ms3_indices <- which(spectra_id %in% ms3_parents)
peaks.ms3parents.list <- as.list(pd[ms3_indices, ])

# Build lookup of rounded parent precursorMz
parent_lookup <- lapply(peaks.ms3parents.list, function(x) round(x[, 1]))
names(parent_lookup) <- as.character(round(ms3_parents_sd$precursorMz))

# Round MS4 precursor columns for matching
ms4_candidates <- ms4_candidates %>%
  mutate(
    precursorMz_round = round(precursorMz),
    precursorMz_MS2_round = round(precursorMz.MS2)
  )

# Keep candidate if its precursor m/z is found among peaks of its corresponding MS3 parent
keep_idx <- logical(nrow(ms4_candidates))
for (i in seq_len(nrow(ms4_candidates))) {
  parent_key <- as.character(ms4_candidates$precursorMz_MS2_round[i])
  if (parent_key %in% names(parent_lookup)) {
    keep_idx[i] <- ms4_candidates$precursorMz_round[i] %in% parent_lookup[[parent_key]]
  } else {
    keep_idx[i] <- FALSE
  }
}
ms4_candidates_filtered <- ms4_candidates[keep_idx, ]
```

### Check for duplicate MS4 spectra and remove if necessary
```{r}
# Prepare merged_spectra metadata for comparison
merged_sd <- as.data.frame(spectraData(merged_spectra)) %>%
  mutate(
    precursorMz_round = round(precursorMz),
    precursorMz_MS2_round = round(precursorMz.MS2)
  )

# Build unique composite keys for existing merged spectra and MS4 candidates
existing_keys <- paste(
  merged_sd$feature_id,
  merged_sd$msLevel,
  merged_sd$precursorMz_round,
  merged_sd$precursorMz_MS2_round,
  sep = "_"
)

candidate_keys <- paste(
  ms4_candidates_filtered$feature_id,
  ms4_candidates_filtered$msLevel,
  ms4_candidates_filtered$precursorMz_round,
  ms4_candidates_filtered$precursorMz_MS2_round,
  sep = "_"
)

# Retain only candidates whose key is NOT already in merged_spectra
keep_new_idx <- !(candidate_keys %in% existing_keys)
ms4_candidates_filtered <- ms4_candidates_filtered[keep_new_idx, ]

message(sum(!keep_new_idx), " candidate MS4 spectra removed because they already exist in merged_spectra.")
message(nrow(ms4_candidates_filtered), " unique new MS4 spectra retained.")
```
### Select best MS4 per precMS2.precMS3
```{r}
ms4_candidates_filtered <- ms4_candidates_filtered %>%
  mutate(
    precMS2.precMS3 = paste0(
      "precMS2_", precursorMz_MS2_round,
      "_precMS3_", precursorMz_round
    )
  )

newms4_best <- ms4_candidates_filtered %>%
  group_by(precMS2.precMS3) %>%
  slice_max(order_by = peaksCount, n = 1, with_ties = FALSE) %>%
  ungroup()
```

### Append new MS4 spectra to merged_spectra
```{r}
sd_merged <- as.data.frame(spectraData(merged_spectra))

# Round precursor m/z for consistency
sd_merged$precMzRound <- round(sd_merged$precursorMz)
newms4_best$precMzRound <- round(newms4_best$precursorMz)

# Create merge keys
mergeKey_existing <- paste(sd_merged$dataOrigin, sd_merged$scanIndex, sep = "_")
mergeKey_newms4   <- paste(newms4_best$dataOrigin, newms4_best$scanIndex, sep = "_")

# Combine metadata tables
common_cols <- intersect(names(sd_merged), names(newms4_best))
merged_df <- bind_rows(
  sd_merged[, common_cols],
  newms4_best[, common_cols]
)

# Assign new merged tree IDs
merged_df <- merged_df %>%
  mutate(MergedMSntreeID = as.integer(factor(feature_id)))

# Update Spectra object
merged_spectra <- ftms_msn_tree[
  paste(spectraData(ftms_msn_tree)$dataOrigin,
        spectraData(ftms_msn_tree)$scanIndex, sep = "_") %in%
    c(mergeKey_existing, mergeKey_newms4)
]

spectraData(merged_spectra) <- S4Vectors::DataFrame(merged_df)

```

### Check for duplicates and post-processing if duplicates found

```{r}
# Extract metadata
merged_sd <- as.data.frame(spectraData(merged_spectra))

# Identify groups with same feature_id, msLevel, precursorMz, precursorMz.MS2
merged_sd_MS4 <- merged_sd[merged_sd$msLevel == 4,]
merged_sd_MS4$precMz.MS2.Round <- round(merged_sd_MS4$precursorMz.MS2)
dup_groups <- merged_sd_MS4 %>%
  group_by(feature_id, msLevel, precMzRound, precMz.MS2.Round) %>%
  mutate(n_in_group = n()) %>%
  ungroup() %>%
  filter(n_in_group > 1)

# Report duplicates
if (nrow(dup_groups) > 0) {
  message("Found ", nrow(dup_groups), " spectra in duplicate groups:")
dup_groups_ordered <- dup_groups %>%
    arrange(feature_id)
  
  print(dup_groups_ordered[, c("feature_id","msLevel","precMz.MS2.Round","precMzRound","peaksCount")])} else {
  message("No duplicate MS4 spectra found.")
}
```
# Check MS levels in merged MSn trees

```{r}
# Extract spectraData from merged_spectra
merged_sd <- as.data.frame(spectraData(merged_spectra))

# Count spectra per feature : MS level
feature_counts <- table(merged_sd$feature_id, merged_sd$msLevel)
feature_counts <- as.data.frame.matrix(feature_counts)  # columns = MS levels

# Define MS levels present
ms_levels <- colnames(feature_counts)

# Compute summary for each MS level with descriptive names
ms_summary <- lapply(ms_levels, function(ms) {
  counts <- feature_counts[[ms]]
  summary(counts)
})

# Set descriptive names
names(ms_summary) <- paste0("MS", ms_levels, " spectra per feature")

# Print summaries
ms_summary
```

## inspect merged MSn tree with 5 MS3 spectra
```{r}

merged_sd <- as.data.frame(spectraData(merged_spectra))

# Identify features with exactly three MS3 spectra
features_5MS3 <- merged_sd %>%
  filter(msLevel == 3) %>%
  group_by(feature_id) %>%
  summarise(n_MS3 = n(), .groups = "drop") %>%
  filter(n_MS3 == 5) %>%
  pull(feature_id)

# Subset merged_spectra for these feature(s)
merged_subset <- merged_spectra[merged_sd$feature_id == features_5MS3[[1]]]
as.data.frame(spectraData(merged_subset))
```


#Save

Finally we verify if the number of unique features corresponds to the
number of unique MergedMSntreeID to ensure that we have one tree per feature.
```{r}
length(unique(unlist(merged_spectra$feature_id)))
length(unique(unlist(merged_spectra$MergedMSntreeID)))
length(unique(unlist(merged_spectra$MSntreeID)))
```

we have `r length(unique(merged_spectra$MergedMSntreeID))` unique feature and
unique MSntreeID, That's mean that we have for each feature one
representative tree.

Now we save the resulting ftms spectra object to a local directory.

```{r}
ftms_one_merged_tree <- merged_spectra
save(ftms_one_merged_tree, file = file.path(pth, "data", "ftms_one_merged_tree.RData"))
```

```{r}
load("data/ftms_one_merged_tree.RData")
```


# Session information

```{r}
sessionInfo()
```
