---
title: "Selecting one representative tree per feature"
format:
  html:
    toc: true
    self-contained: true
author: "Ahlam Mentag, Johannes Rainer"
editor:
  markdown:
    wrap: 72
---

# Introduction

This Quarto document presents the analysis of FTMS flaxseed data for
selecting one representative tree per feature. Preprocessing and feature
definition was performed in
[ftms_preprocessing.qmd](ftms_preprocessing.qmd).

# Data import

Loading all required libraries and importing the results from the data
preprocessing.

```{r}
#| message: false
library(dplyr)
library(tidyr)
library(xcms)
library(MsExperiment)
library(Spectra)
library(readxl)
library(IRanges)
library(S4Vectors)
register(SerialParam())
pth <- getwd()
pth
```

First we load the `XcmsExperiment` ftms object with the preprocessing
results.

```{r}
load(file.path(pth, "data", "ftms_680.RData"))
class(ftms)
```

# Explore MS2 spectra associated to features

Next, we use `xcms::featureSpectra()` to extract all MS2 spectra for
each feature from `ftms` object.

> The `xcms::featureSpectra()` function returns a `Spectra` object.
>
> When using `msLevel = 2`, the resulting `Spectra` object contains all
> MS2 spectra whose retention time and precursor m/z fall within the
> ranges of any chromatographic peak of a feature.

```{r}
ms2 <- featureSpectra(ftms, msLevel = 2)

msLevel(ms2) |>
  table()
```

```{r}
spectraData(ms2)
```

it returns `r length(ms2)` MS level 2 spectra. The total number of features for
which an MS2 spectrum was identified is:

```{r}
length(unique(ms2$feature_id))
length(ms2$feature_id)
```

Here we print the initial number of MS levels.

```{r}
s <- spectra(ftms)
table(msLevel(s))
```

In total, there were 20904 MS2 spectra. This means that 89.88% of the MS2
spectra ((20904-2115)\*100/20904) could not be assigned to any
chromatographic peak.

The percentage of features with at least one MS2:

```{r}
# Total number of features in ftms
n_total_features <- nrow(featureDefinitions(ftms))

# Features with ≥1 MS2 (those returned by featureSpectra)
n_features_with_ms2 <- length(unique(ms2$feature_id))

# Percentage
perc_features_with_ms2 <- (n_features_with_ms2 / n_total_features) * 100
perc_features_with_ms2

```

The total number of features with at least one MS2 is `r n_features_with_ms2`,
which corresponds to `r format(perc_features_with_ms2, digits = 3)`%.

A summary on the number of MS2 spectra assigned to a feature is shown
below:

```{r}
table(ms2$feature_id) |>
    quantile()
```

The median number of MS2 spectra per feature (across all data files) is
hence `r median(table(ms2$feature_id))`.

the percentage of MS2 spectra that are assigned to a feature

```{r}
# Total number of MS2 spectra in the dataset
n_total_ms2 <- sum(msLevel(s) == 2)
sp <- spectraData(ms2)
sp$groups <- paste0(sp$dataOrigin, "_", sp$scanIndex)
# Number of MS2 spectra assigned to features
n_assigned_ms2 <- length(unique(sp$groups))

# Percentage assigned
perc_ms2_assigned <- (n_assigned_ms2 / n_total_ms2) * 100
perc_ms2_assigned

```

The total number of assigned MS2 spectra is `r n_assigned_ms2`, which
corresponds to a percentage of `r format(perc_ms2_assigned, digits = 3)`.

Note that an MS2 spectrum could also be assigned to more than one
feature. The total number of MS2 spectra assigned to more than one
feature is:

```{r}
ms2_id <- paste0(ms2$dataOrigin, "_", ms2$acquisitionNum)
ms2_table <- table(ms2_id)
ms2_mult <- names(ms2_table)[ms2_table > 1]

```

Of the `r length(unique(ms2_id))` MS2 spectra, `r length(ms2_mult)` are assigned
to more than one feature. Below we evaluate to how many features these MS2
spectra are assigned:

```{r}
quantile(ms2_table[ms2_table > 1])
```

All MS2 spectra that are assigned to more than one feature, are
assigned to two features.

We next identify the features to which these MS2 spectra are assigned.

```{r}
ms2_mult_features <- sp$feature_id[sp$groups %in% ms2_mult]
```

The `r length(ms2_mult)` ambiguous MS2 features are assigned in to in total
`r length(unique(ms2_mult_features))` features. These are listed, along with the
number of assigned ambiguous MS2 spectra in the table below.

```{r, results = "asis"}
tmp <- table(ms2_mult_features) |> as.data.frame()
colnames(tmp) <- c("feature_id", "freq")
tmp <- tmp[order(tmp$freq, decreasing = TRUE), ]
rownames(tmp) <- NULL
library(pander)
pandoc.table(tmp, style = "rmarkdown")
```


## Evaluation of ambiguous MS2 spectra assignment

We next evaluate some of the ambiguous assignment of the same MS2 spectrum to
multiple features.

### Example 1

```{r}
fid <- as.character(tmp$feature_id[1L])
```

Information on the feature is summarized below.

```{r, results = "asis"}
feature_summary_table <- function(x, id) {
    fd <- featureDefinitions(x)[id, ]
    fa <- featureArea(x, features = id)
    data.frame(
        mzmed = fd$mzmed,
        mzwidth = fd$mzmax - fd$mzmin,
        mzwidth_full = fa[1L, "mzmax"] - fa[1L, "mzmin"],
        rtmed = fd$rtmed,
        rtwidth = fd$rtmax - fd$rtmin,
        rtwidth_full = fa[1L, "rtmax"] - fa[1L, "rtmin"],
        chrom_peak_count = length(fd$peakidx[[1L]])
    )
}
tab <- feature_summary_table(ftms, fid)
pandoc.table(tab, style = "rmarkdown", split.table = Inf)
```

Next we evaluate whether the current feature shared chromatographic peaks with
any other feature:

```{r}
#' Function to get the index of chromatographic peaks shared with other
#' features for a feature with ID `id`.
#'
#' @noRd
shared_chrom_peaks <- function(x, id) {
    pids <- unlist(featureDefinitions(x)[id, "peakidx"])
    other_features <- lapply(featureDefinitions(x)$peakidx,
                             function(z) z[z %in% pids])
    names(other_features) <- rownames(featureDefinitions(x))
    other_features <- other_features[lengths(other_features) > 0]
    other_features[names(other_features) != id]
}

## are chrom peaks shared with other feature?
shared_cpks <- shared_chrom_peaks(ftms, fid)
shared_cpks
```

No chromatographic peaks are shared. Next we summarize the number of ambiguous
MS2 spectra and the feature(s) with which they are shared.

```{r, results = "asis"}
#' Get all MS2 for the current feature
fid_ms2 <- ms2[ms2$feature_id == fid]
#' Get the ambiguous MS2 for current feature
fid_ms2_mult <- intersect(ms2_mult, ms2_id[ms2$feature_id == fid])
#' Get the other features for the ambiguous MS2 of current feature
fid_other_features <- setdiff(ms2$feature_id[ms2_id %in% fid_ms2_mult], fid)

tab <- data.frame(ms2_count = length(fid_ms2),
                  ambiguous_ms2_count = length(fid_ms2_mult),
                  other_feature = paste0(fid_other_features, collapse = ","))
pandoc.table(tab, style = "rmarkdown", split.table = Inf)
```

A considerable number of MS2 spectra are shared with another feature. We next
visually inspect their data, i.e., their EIC and the retention time where the
unambiguous and ambiguous MS2 were detected.

```{r}
fid_eic <- featureChromatograms(ftms, features = fid)
other_eic <- featureChromatograms(ftms, features = fid_other_features)

par(mfrow = c(2, 1))
plot(fid_eic)
grid()
abline(v = rtime(fid_ms2), col = "blue")
abline(v = rtime(ms2)[ms2_id %in% fid_ms2_mult], col = "red")

plot(other_eic)
grid()
abline(v = rtime(fid_ms2), col = "blue")
abline(v = rtime(ms2)[ms2_id %in% fid_ms2_mult], col = "red")
```

So, the two features seem overlapping on RT and *m/z* - it seems very odd that
they were not combined in the same feature.

Checking another example.

### Example 2

```{r}
fid <- as.character(tmp$feature_id[3L])
tab <- feature_summary_table(ftms, fid)
pandoc.table(tab, style = "rmarkdown", split.table = Inf)
```

Next we evaluate whether the current feature shared chromatographic peaks with
any other feature:

```{r}
## are chrom peaks shared with other feature?
shared_cpks <- shared_chrom_peaks(ftms, fid)
shared_cpks
```

No chromatographic peaks are shared. Next we summarize the number of ambiguous
MS2 spectra and the feature(s) with which they are shared.

```{r, results = "asis"}
#' Get all MS2 for the current feature
fid_ms2 <- ms2[ms2$feature_id == fid]
#' Get the ambiguous MS2 for current feature
fid_ms2_mult <- intersect(ms2_mult, ms2_id[ms2$feature_id == fid])
#' Get the other features for the ambiguous MS2 of current feature
fid_other_features <- setdiff(ms2$feature_id[ms2_id %in% fid_ms2_mult], fid)

tab <- data.frame(ms2_count = length(fid_ms2),
                  ambiguous_ms2_count = length(fid_ms2_mult),
                  other_feature = paste0(fid_other_features, collapse = ","))
pandoc.table(tab, style = "rmarkdown", split.table = Inf)
```

A considerable number of MS2 spectra are shared with another feature. We next
visually inspect their data, i.e., their EIC and the retention time where the
unambiguous and ambiguous MS2 were detected.

```{r}
fid_eic <- featureChromatograms(ftms, features = fid)
other_eic <- featureChromatograms(ftms, features = fid_other_features)

par(mfrow = c(2, 1))
plot(fid_eic)
grid()
abline(v = rtime(fid_ms2), col = "blue")
abline(v = rtime(ms2)[ms2_id %in% fid_ms2_mult], col = "red")

plot(other_eic)
grid()
abline(v = rtime(fid_ms2), col = "blue")
abline(v = rtime(ms2)[ms2_id %in% fid_ms2_mult], col = "red")
```

There seems definitely to be some issue with peak detection and/or
correspondence analysis. The retention times seem to clearly overlap. Thus we
evaluate whether the m/z ranges of the associated chromatographic peaks are
similar or different maybe explaining the observed separation.

```{r}
#' Getting the chromatographic peaks assigned to the first and to the
#' second feature
a <- chromPeaks(ftms)[featureDefinitions(ftms)[fid, "peakidx"][[1L]], ]
b <- chromPeaks(ftms)[featureDefinitions(ftms)[
                   fid_other_features, "peakidx"][[1L]], ]

par(mfrow = c(1, 1))
plot(1, 1, pch = NA, xlim = range(rbind(a, b)[, c("rtmin", "rtmax")]),
     ylim = range(rbind(a, b)[, c("mzmin", "mzmax")]))
grid()
rect(xleft = a[, "rtmin"], xright = a[, "rtmax"],
     ybottom = a[, "mzmin"], ytop = a[, "mzmax"], border = "#ff000080")
rect(xleft = b[, "rtmin"], xright = b[, "rtmax"],
     ybottom = b[, "mzmin"], ytop = b[, "mzmax"], border = "#0000ff80", lty = 2)
```

There seems no explanation why these chrom peaks are not grouped into the same
feature. Thus, this seems to be a problem with the correspondence analysis.

Note: on second look.

```{r}
#' Just to check if these are gap-filled
chromPeakData(ftms)[featureDefinitions(ftms)[fid, "peakidx"][[1L]], "is_filled"]
#' so, yes, there are gap-filled peaks that are overlapping.
```

Need to understand why these chrom peaks were grouped into different
features. Extracting the EIC for this m/z slice.

```{r}
#' Define the m/z range of the present features
mzr <- range(c(a[, c("mzmin", "mzmax")], b[, c("mzmin", "mzmax")]))
#' Get rid of gap-filled chromatographic peaks - we don't have/want them
#' for correspondence analysis.
ftms_mod <- dropFilledChromPeaks(ftms)

#' Just to confirm - we extract again the chromatograophic peak areas, but
#' this time it's just detected, no gap-filled peaks.
a <- chromPeaks(ftms_mod)[featureDefinitions(ftms_mod)[fid, "peakidx"][[1L]], ]
b <- chromPeaks(ftms_mod)[featureDefinitions(ftms_mod)[
                   fid_other_features, "peakidx"][[1L]], ]
par(mfrow = c(1, 1))
plot(1, 1, pch = NA, xlim = range(rbind(a, b)[, c("rtmin", "rtmax")]),
     ylim = range(rbind(a, b)[, c("mzmin", "mzmax")]))
grid()
rect(xleft = a[, "rtmin"], xright = a[, "rtmax"],
     ybottom = a[, "mzmin"], ytop = a[, "mzmax"], border = "#ff000080")
rect(xleft = b[, "rtmin"], xright = b[, "rtmax"],
     ybottom = b[, "mzmin"], ytop = b[, "mzmax"], border = "#0000ff80", lty = 2)
```

So, indeed, the chromatographic peaks of the two features, even if we don't
consider gap-filled peaks are overlapping in m/z and RT. So, we would expect
that these get grouped into the same feature. We next show the results from the
correspondence analysis for the m/z slice:

```{r}
eic <- chromatogram(ftms_mod, mz = mzr, rt = c(430, 520))
plotChromPeakDensity(eic)
```

Indeed, two separate features were identified - apparently based on the RT of
their peak apex. In this case, we should increase the value for the `bw`
parameter of `PeakDensityParam`.

```{r}
#' Use the same setting as for the preprocessing, but increase `bw` from 2 to 3.
pdp <- PeakDensityParam(sampleGroups = sampleData(ftms)$sample_id,
                        bw = 4, minFraction = 0.3, binSize = 0.1,
                        ppm = 10)
plotChromPeakDensity(eic, simulate = TRUE, param = pdp)

```

Evaluating another feature.

### Example 3

```{r}
fid <- as.character(tmp$feature_id[5L])
tab <- feature_summary_table(ftms, fid)
pandoc.table(tab, style = "rmarkdown", split.table = Inf)
```

Next we evaluate whether the current feature shared chromatographic peaks with
any other feature:

```{r}
## are chrom peaks shared with other feature?
shared_cpks <- shared_chrom_peaks(ftms, fid)
shared_cpks
```

No chromatographic peaks are shared. Next we summarize the number of ambiguous
MS2 spectra and the feature(s) with which they are shared.

```{r, results = "asis"}
#' Get all MS2 for the current feature
fid_ms2 <- ms2[ms2$feature_id == fid]
#' Get the ambiguous MS2 for current feature
fid_ms2_mult <- intersect(ms2_mult, ms2_id[ms2$feature_id == fid])
#' Get the other features for the ambiguous MS2 of current feature
fid_other_features <- setdiff(ms2$feature_id[ms2_id %in% fid_ms2_mult], fid)

tab <- data.frame(ms2_count = length(fid_ms2),
                  ambiguous_ms2_count = length(fid_ms2_mult),
                  other_feature = paste0(fid_other_features, collapse = ","))
pandoc.table(tab, style = "rmarkdown", split.table = Inf)
```

A considerable number of MS2 spectra are shared with another feature. We next
visually inspect their data, i.e., their EIC and the retention time where the
unambiguous and ambiguous MS2 were detected.

```{r}
fid_eic <- featureChromatograms(ftms, features = fid)
other_eic <- featureChromatograms(ftms, features = fid_other_features)

par(mfrow = c(2, 1))
plot(fid_eic)
grid()
abline(v = rtime(fid_ms2), col = "blue")
abline(v = rtime(ms2)[ms2_id %in% fid_ms2_mult], col = "red")

plot(other_eic)
grid()
abline(v = rtime(fid_ms2), col = "blue")
abline(v = rtime(ms2)[ms2_id %in% fid_ms2_mult], col = "red")
```

There seems definitely to be some issue with peak detection and/or
correspondence analysis. The retention times seem to clearly overlap. Thus we
evaluate whether the m/z ranges of the associated chromatographic peaks are
similar or different maybe explaining the observed separation.

```{r}
a <- chromPeaks(ftms)[featureDefinitions(ftms)[fid, "peakidx"][[1L]], ]
b <- chromPeaks(ftms)[featureDefinitions(ftms)[
                   fid_other_features, "peakidx"][[1L]], ]

plot(1, 1, pch = NA, xlim = range(rbind(a, b)[, c("rtmin", "rtmax")]),
     ylim = range(rbind(a, b)[, c("mzmin", "mzmax")]))
grid()
rect(xleft = a[, "rtmin"], xright = a[, "rtmax"],
     ybottom = a[, "mzmin"], ytop = a[, "mzmax"], border = "#ff000080")
rect(xleft = b[, "rtmin"], xright = b[, "rtmax"],
     ybottom = b[, "mzmin"], ytop = b[, "mzmax"], border = "#0000ff80", lty = 2)
```

Similar to the previous example, there is absolutely no reason these chrom peaks
are not part of the same feature! Again, most of the peaks are gap-filled. We
plot the peak area again without gap-filled peaks.

```{r}
a <- chromPeaks(ftms_mod)[featureDefinitions(ftms_mod)[fid, "peakidx"][[1L]], ]
b <- chromPeaks(ftms_mod)[featureDefinitions(ftms_mod)[
                   fid_other_features, "peakidx"][[1L]], ]
par(mfrow = c(1, 1))
plot(1, 1, pch = NA, xlim = range(rbind(a, b)[, c("rtmin", "rtmax")]),
     ylim = range(rbind(a, b)[, c("mzmin", "mzmax")]))
grid()
rect(xleft = a[, "rtmin"], xright = a[, "rtmax"],
     ybottom = a[, "mzmin"], ytop = a[, "mzmax"], border = "#ff000080")
rect(xleft = b[, "rtmin"], xright = b[, "rtmax"],
     ybottom = b[, "mzmin"], ytop = b[, "mzmax"], border = "#0000ff80", lty = 2)

```

Here it becomes quite obvious that different features were defined, as there is
some difference in RT between the peak areas. We next evaluate the results from
the correspondence.

```{r}
mzr <- range(c(a[, c("mzmin", "mzmax")], b[, c("mzmin", "mzmax")]))
eic <- chromatogram(ftms_mod, mz = mzr, rt = c(2800, 2990))
plotChromPeakDensity(eic)
```

Indeed, different features were identified - apparently based on the RT of
their peak apex. In this case, we should increase the value for the `bw`
parameter of `PeakDensityParam`. Increasing the `bw` parameter will not help in
this case, because there are only few, sparse peaks detected in the individual
samples. Maybe increasing the `minFraction` might help?

```{r}
#' Use the same setting as for the preprocessing, but increase `bw` from 2 to 3.
pdp <- PeakDensityParam(sampleGroups = sampleData(ftms)$sample_id,
                        bw = 3, minFraction = 0.5, binSize = 0.1,
                        ppm = 10)
plotChromPeakDensity(eic, simulate = TRUE, param = pdp)

```


### Example 4

```{r}
fid <- as.character(tmp$feature_id[7L])
tab <- feature_summary_table(ftms, fid)
pandoc.table(tab, style = "rmarkdown", split.table = Inf)
```

Next we evaluate whether the current feature shared chromatographic peaks with
any other feature:

```{r}
## are chrom peaks shared with other feature?
shared_cpks <- shared_chrom_peaks(ftms, fid)
shared_cpks
```

No chromatographic peaks are shared. Next we summarize the number of ambiguous
MS2 spectra and the feature(s) with which they are shared.

```{r, results = "asis"}
#' Get all MS2 for the current feature
fid_ms2 <- ms2[ms2$feature_id == fid]
#' Get the ambiguous MS2 for current feature
fid_ms2_mult <- intersect(ms2_mult, ms2_id[ms2$feature_id == fid])
#' Get the other features for the ambiguous MS2 of current feature
fid_other_features <- setdiff(ms2$feature_id[ms2_id %in% fid_ms2_mult], fid)

tab <- data.frame(ms2_count = length(fid_ms2),
                  ambiguous_ms2_count = length(fid_ms2_mult),
                  other_feature = paste0(fid_other_features, collapse = ","))
pandoc.table(tab, style = "rmarkdown", split.table = Inf)
```

A considerable number of MS2 spectra are shared with another feature. We next
visually inspect their data, i.e., their EIC and the retention time where the
unambiguous and ambiguous MS2 were detected.

```{r}
fid_eic <- featureChromatograms(ftms, features = fid)
other_eic <- featureChromatograms(ftms, features = fid_other_features)

par(mfrow = c(2, 1))
plot(fid_eic)
grid()
abline(v = rtime(fid_ms2), col = "blue")
abline(v = rtime(ms2)[ms2_id %in% fid_ms2_mult], col = "red")

plot(other_eic)
grid()
abline(v = rtime(fid_ms2), col = "blue")
abline(v = rtime(ms2)[ms2_id %in% fid_ms2_mult], col = "red")
```

There seems definitely to be some issue with peak detection and/or
correspondence analysis. The retention times seem to clearly overlap. Thus we
evaluate whether the m/z ranges of the associated chromatographic peaks are
similar or different maybe explaining the observed separation.

```{r}
a <- chromPeaks(ftms)[featureDefinitions(ftms)[fid, "peakidx"][[1L]], ]
b <- chromPeaks(ftms)[featureDefinitions(ftms)[
                   fid_other_features, "peakidx"][[1L]], ]

plot(1, 1, pch = NA, xlim = range(rbind(a, b)[, c("rtmin", "rtmax")]),
     ylim = range(rbind(a, b)[, c("mzmin", "mzmax")]))
grid()
rect(xleft = a[, "rtmin"], xright = a[, "rtmax"],
     ybottom = a[, "mzmin"], ytop = a[, "mzmax"], border = "#ff000080")
rect(xleft = b[, "rtmin"], xright = b[, "rtmax"],
     ybottom = b[, "mzmin"], ytop = b[, "mzmax"], border = "#0000ff80")
```


```{r}
```


Now proceeding with the previous code.


Manually evaluate some of these cases:

```{r}
a <- ms2[ms2_id == ms2_mult[3L]]
print(spa <- spectraData(a))

a_chr <- featureChromatograms(ftms, features = a$feature_id)
par(mfrow = c(1, 2))
plot(a_chr[1])
abline(v = a$rtime, col = "red")
plot(a_chr[2])
abline(v = a$rtime, col = "red")

ms2_rt <- a$rtime
abs(ms2_rt - spa$feature_rtmed)
abs(ms2_rt[1] - spa$feature_rtmed[1])/abs(ms2_rt[2] - spa$feature_rtmed[2])
```

The two features to which MS2 spectrum "a" was assigned have exactly the
same feature_mzmed = 285.181. The features are also very close in the
retention time dimension, eventually even not well separated. The MS2
spectrum (retention time highlighted in red) is associated to a
chromatographic peak in one sample that was eventually wrongly assigned
to the earlier feature. In this case, the MS2 assignment to feature
FT2029 should be removed. This can be done based on the relative
retention time differences with respect to the two features.

For most other examples, the feature seem to represent noise or at least
no clear chromatographic signal (see for example below).

```{r}
a <- ms2[ms2_id == ms2_mult[200L]]
print(spa <- spectraData(a))

a_chr <- featureChromatograms(ftms, features = a$feature_id)
par(mfrow = c(1, 2))
plot(a_chr[1])
abline(v = a$rtime, col = "red")
plot(a_chr[2])
abline(v = a$rtime, col = "red")
ms2_rt <- a$rtime
abs(ms2_rt - spa$feature_rtmed)
abs(ms2_rt[1] - spa$feature_rtmed[1])/abs(ms2_rt[2] - spa$feature_rtmed[2])
```

Here we have a look at MS2 that were assigned to three different
features:

```{r}
a <- ms2[ms2_id == ms2_three[1L]]
print(spa <- spectraData(a))

a_chr <- featureChromatograms(ftms, features = a$feature_id)
par(mfrow = c(1, 3))
plot(a_chr[1])
abline(v = a$rtime, col = "red")
plot(a_chr[2])
abline(v = a$rtime, col = "red")
plot(a_chr[3])
abline(v = a$rtime, col = "red")
ms2_rt <- a$rtime

abs(ms2_rt - spa$feature_rtmed)
abs(ms2_rt[1] - spa$feature_rtmed[1])/abs(ms2_rt[2] - spa$feature_rtmed[2])
abs(ms2_rt[1] - spa$feature_rtmed[1])/abs(ms2_rt[3] - spa$feature_rtmed[3])
```

Also here, all three features have exactly the same mzmed. The first
feature (FT0170) has the retention time closest to the MS2 RT, and has
also the best peak shape. the two other features don't have a proper
peak shape and look mainly like noise.

```{r}
a <- ms2[ms2_id == ms2_three[2L]]
print(spa <- spectraData(a))

a_chr <- featureChromatograms(ftms, features = a$feature_id)
par(mfrow = c(1, 3))
plot(a_chr[1])
abline(v = a$rtime, col = "red")
plot(a_chr[2])
abline(v = a$rtime, col = "red")
plot(a_chr[3])
abline(v = a$rtime, col = "red")
ms2_rt <- a$rtime

abs(ms2_rt - spa$feature_rtmed)
abs(ms2_rt[1] - spa$feature_rtmed[1])/abs(ms2_rt[2] - spa$feature_rtmed[2])
abs(ms2_rt[1] - spa$feature_rtmed[1])/abs(ms2_rt[3] - spa$feature_rtmed[3])
```

Also here, all three features have exactly the same mzmed. The first
feature (FT2021) has the retention time closest to the MS2 RT, and
clearly has also the best peak shape. The two other features don't have
a proper peak shape, they look like multiple unseparated peaks.

**Question**: (maybe for the next meeting with Rebecca) should we maybe
consider keeping only *high quality* signals? trying to remove these
noisy features/chromatographic peaks? ==\> we could try gradually
increasing the sn ratio again

# Explore, for all MS2 spectra that were assigned to multiple features, the feature_rtmed and feature_mzmed of these features.

```{r}
for (id in ms2_mult) {
  a <- ms2[ms2_id == id]
  spa <- spectraData(a)

  mzmeds <- spa$feature_mzmed
  uniq_mzmeds <- unique(mzmeds)

  cat("Checking id:", id, "\n")
  cat("All mzmeds:", paste(round(mzmeds, 5), collapse = ", "), "\n")
  cat("Unique mzmeds:", paste(round(uniq_mzmeds, 5), collapse = ", "), "\n")

  print(spa[, c("feature_id", "feature_rtmed", "feature_mzmed")])

}

```



```{r}
# Example 1 : FT0413 et FT0416
ids1 <- c("FT3944", "FT3944")
a_chr1 <- featureChromatograms(ftms, features = ids1)

plot(a_chr1[1])
abline(v = 462.78, col = "red", lty = 2)

plot(a_chr1[2])
abline(v = 470.61, col = "red", lty = 2)

```

```{r}
# Example 2 : FT2594 et FT2595
ids2 <- c("FT4641", "FT4644")
a_chr2 <- featureChromatograms(ftms, features = ids2)

plot(a_chr2[1])
abline(v = 435.277, col = "red", lty = 2)

plot(a_chr2[2])
abline(v = 422.188, col = "red", lty = 2)

```



# Explore MS2 spectra that have been assigned to features with different mzmed
We have a look specifically at the MS2 spectra that have been
assigned to features with different mzmed. Can one of these assigned
features be unambiguously assigned to the MS2 based on the difference
between precursorMz and feature_mzmed? Is a single feature_mzmed much
closer to the precursorMz of the MS2?


First, for MS2 spectra that were assigned to multiple features, we
analyze the range distribution of feature_mzmed values of these multiple
features.

```{r}
#Create Spectra object a, with MS2 spectra that are assigned to multiple features and extract spectraData
a <- ms2[ms2_id %in% ms2_mult]
spa <- spectraData(a)
spa$ms2_id <- factor(paste0(spa$dataOrigin, spa$acquisitionNum))
ambiguous_assignments <- length(unique(spa$ms2_id))
cat(c("In total", ambiguous_assignments,"MS2 spectra are assigned to multiple features."))

#Explore range of feature_mzmed values for different features to which each MS2 is assigned
# Split feature_mzmed values by ms2_id
split_list <- split(spa$feature_mzmed, spa$ms2_id)

# Check for each ms2_id whether all values are identical
mzmed_nonidentical_logical <- lapply(split_list, function(x) length(unique(x)) > 1)
mzmed_nonidentical_total <- sum(unlist(mzmed_nonidentical_logical))
cat(c("Of all ambiguously assigned MS2 spectra,", mzmed_nonidentical_total*100/ambiguous_assignments,"% are assigned to features with NON-identical mzmed."))

#closer look at ranges of mzmed values for different features assigned to same MS2
# Function to compute range difference
diff_fun <- function(x) {
  max(x, na.rm = TRUE) - min(x, na.rm = TRUE)
}

# apply to all ms2_id levels
mzmed_diff <- sapply(split_list, diff_fun)

# Overview of distribution
summary(mzmed_diff)
opar <- par(mfrow = c(2, 1))
hist(mzmed_diff, breaks = 100, main = "Distribution of feature_mzmed range per ms2_id")
hist(mzmed_diff, xlim = c(0,0.01), breaks = 1000, main = "Distribution of feature_mzmed range per ms2_id")
par(opar)

```

The mzmed values of different features to which an MS2 was assigned was
**never exactly identical**, but the majority of features assigned to
the same MS2 has mzmed values falling in a **range \< 0.001**. We can
consider that **differences \> 0.001 are significant**.

Now we can identify the MS2 spectra that are assigned to features with
**significantly different feature_mzmed** values:

```{r}
# Identify ms2_id with feature_mzmed range > 0.002
significant_ms2 <- mzmed_diff[mzmed_diff > 0.002]
length(significant_ms2)
cat("Of all", length(ms2_mult), "MS2 spectra that were assigned to multiple features, ", length(significant_ms2)*100/length(ms2_mult), "% was assigned to features that differ significantly in mzmed.\n")

# Extract the corresponding ms2_id values
ms2_id_significant <- names(significant_ms2)

```

Now we explore the differences precursorMz−feature_mzmed observed for
those MS2 spectra that are assigned to features with **NO significantly
different feature_mzmed** values.

```{r}
# Define groups based on mzmed_diff
ms2_id_non_significant <- names(mzmed_diff[mzmed_diff <= 0.002])
ms2_id_significant <- names(mzmed_diff[mzmed_diff > 0.002])

# Compute precursorMz - feature_mzmed for all spectra
spa$delta <- spa$precursorMz - spa$feature_mzmed

# Extract deltas for the two groups
delta_non_significant <- spa$delta[spa$ms2_id %in% ms2_id_non_significant]
delta_significant     <- spa$delta[spa$ms2_id %in% ms2_id_significant]

# Explore distribution of deltas in non-significant group
summary(abs(delta_non_significant))

hist(delta_non_significant, breaks = 100, main = "Delta distribution (non-significant)")
```

Based on the distribution observed among spectra that were assigned to
features with the same mzmed (the "non-significant" group), we define a
threshold for "close to zero" difference precursorMz−feature_mzmed. For
example, we take a quantile of the absolute differences.

```{r}
abs_delta_non_significant <- abs(delta_non_significant)
threshold <- quantile(abs_delta_non_significant, 0.98, na.rm = TRUE)

threshold
cat("For MS2 spectra that are assigned to multiple features with the same featuremzmed, the difference between precursorMz and feature_mzmed is in 98% of the cases below", round(threshold, 3), ".", "A difference between precursorMz and feature_mzmed <", round(threshold, 3),"is the cutoff for close-to-zero.\n")
threshold <- round(threshold,3)

# Count per ms2_id_significant how many deltas are within threshold
#reduce dataset to ms2_id_significant
spar <- spa[spa$ms2_id %in% ms2_id_significant,]
spar$ms2_id <- factor(spar$ms2_id)
split_deltas <- split(abs(spar$delta),
                      spar$ms2_id)
close_counts <- sapply(split_deltas, function(x) sum(x <= threshold, na.rm = TRUE))

cat("How many ms2_id, assigned to features with different RT (rtmed), have 0, 1, 2, ... close-to-zero deltas between precursor-RT and feature-RT:\n")
table(close_counts)
cat("For",table(close_counts)[2]*100/length(ms2_id_significant), "% of the MS2 spectra that were assigned to features with different mzmed, one feature can be unambiguously assigned based on the deltas between precursorMz and feature_mzmed?\n" )

```

In *conclusion:*
```{r}
cat("Of the MS2 spectra that were assigned to at least one feature, ",length(ms2_mult)*100/minimumone,"% was assigned to multiple features.\n")

cat("Of all", length(ms2_mult), "MS2 spectra that were assigned to multiple features, ", length(significant_ms2)*100/length(ms2_mult), "% was assigned to features that differ significantly in mzmed.\n")

cat("For",table(close_counts)[2]*100/length(ms2_id_significant), "% of the MS2 spectra that were assigned to features with different mzmed, one feature can be unambiguously assigned based on the deltas between precursorMz and feature_mzmed.\n" )
```

# Resolving multiple assignments of MS2 spectra to features

We next evaluate the cases above to reduce/remove the multiple
assignments and keep only a single MS2-feature association. First, we
select the duplicated MS2 spectra by identifying those with the same
*dataOrigin* and *acquisitionNum* but different *feature_id* to detect
duplicates assigned to multiple features. From these duplicated MS2
spectra, we then check first (*case A*) if a feature has unique MS2
spectra elsewhere. If that is the case any ambiguous assignment is
removed and only the unambiguous mappings are kept for the feature.
Subsequently, the assignments are checked again and if there are still
duplicated assignments the distance of the MS2 to the respective
features is evaluated (*case B*): we keep the top 10% MS2-feature
assignments. All other assignments are dropped for the respective
feature.

```{r}
select_unique_MS2 <- function(ms2, ftms) {

  sp <- spectraData(ms2)

  # Create a unique identifier for each unique groupe of (dataOrigin, scanIndex)
  sp$groups <- paste0(sp$dataOrigin, "_", sp$scanIndex)

  # Create indices for duplicated groups
  group_idx <- split(seq_len(nrow(sp)), sp$groups)

  # define MS2 as ambiguous if they are duplicated in sp (i.e., are assigned
  # to >= 2 features. We use the unique ID for the MS2 spectra and their
  # indices in sp for that.
  sp$is_ambiguous <- FALSE
  sp$is_ambiguous[sp$groups %in% names(group_idx)[lengths(group_idx) > 1]] <- TRUE

  # Initialization
  final_feature <- vector("list", nrow(sp))  # For the chosen features
  amb_f <- vector("list", nrow(sp))          # For ambiguous features

  # Table of occurences for each feature in a dataOrigine
  # jo: I think it's not good to use this table. it can happen (and it's not
  # a problem) if we have 2 MS2 spectra for the same feature in the same
  # sample (dataOrigin). The problem is if the same MS2 is assigned to
  # two different features.
  # feature_table <- xtabs(~ feature_id + dataOrigin, data = sp)

  # Loop on each group
  for (gname in names(group_idx)) {
    # line indices of each group
    gidx <- group_idx[[gname]]
    # all features that were assigned to the current group
    feature_ids <- unique(unlist(sp$feature_id[gidx]))

    # if a scanIndex was assigned to a unique feature, we don't do anything
    if (length(feature_ids) == 1) {
      final_feature[[gidx[1]]] <- feature_ids
      amb_f[[gidx[1]]] <- character(0)
      next  # go to the next group...
    }

    #rt of the first group
    ms2_rt <- rtime(ms2)[gidx[1]]


    # rtmed and intensity of the feature
    rt_features <- featureDefinitions(ftms)[feature_ids, "rtmed"]

    # Initialisation
    keep_feat <- character(0) #features to keep
    amb_feat <- feature_ids #ambiguous features

    # Loop on each feature in the group
    for (i in seq_along(feature_ids)) {
      i_id <- feature_ids[i]

      # Case A
      # Verify if a feature was assigned at least once to a unique MS2 so there
      # is an(other) unambiguous MS2-feature assignment. We thus skip/drop
      # the current assignment and move to the next.
      # unamb_f <- any(feature_table[i_id, ] == 1)
      if (any(!sp$is_ambiguous[sp$feature_id == i_id])) {
        # If yes we remove it's assignation to this scanIndex and
        next  # go to the next feature...
      }

      # Case B
      # Calculate the RT ratio
      num <- abs(ms2_rt - rt_features[i])
      ratios_rt <- sapply(seq_along(feature_ids)[-i], function(j) {
        denom <- abs(ms2_rt - rt_features[j])
        if (denom == 0) return(Inf)
        num / denom
      })

      # Condition RT
      keep_rt <- which(ratios_rt < 1)  # keep assignment to features with ratio < 1
      if (length(keep_rt)) {
        thr <- quantile(ratios_rt[keep_rt], 0.1)  # top 10% of the smallest rt
        if (any(ratios_rt[keep_rt] <= thr)) keep_feat <- c(keep_feat, i_id)
      }
    }

    # Update the features to keep the final features and the ambiguous ones
    final_feature[[gidx[1]]] <- unique(keep_feat)
    amb_f[[gidx[1]]] <- setdiff(feature_ids, keep_feat)
  }

  # Create a new spectra object with the updated features_id and ambiguous column
  ms2$is_ambiguous <- sp$is_ambiguous
  ms2_res <- ms2[lengths(final_feature) > 0]
  ms2_res$feature_id <- unlist(final_feature, use.names = FALSE)
  ms2_res$ambiguous <- CharacterList(amb_f[lengths(final_feature) > 0])
  ms2_res
}

ms2 <- select_unique_MS2(ms2, ftms)
```

Removing duplicated assignment reduced the data set to `r length(ms2)`
spectra for `r length(unique(ms2$feature_id))` unique features.

### Validation

We next evaluate whether the assignment of single MS2 spectra to
multiple features has been resolved. First we extract the spectra that
have been declared as *ambiguous* (i.e., that were initially assigned to
more than one feature).

```{r}
sp <- spectraData(ms2[ms2$is_ambiguous], c("acquisitionNum", "dataOrigin",
                                           "feature_id", "ambiguous"))
as.data.frame(sp)
```

Next we evaluate if each MS2 spectrum is now assigned to a single
feature (even if it might be/was ambiguous).

```{r}
sp_id <- paste0(ms2$dataOrigin, ms2$acquisitionNum)
max(table(sp_id))
```

Each MS2 spectrum is thus assigned to a single feature.

# Extract all MS2-MSn spectra associated to features

We next define a function to extract the full MSn tree for features.
This function takes an `XcmsExperiment` object as input, extracts the
MS2 spectra using `xcms::featureSpectra()` function, and then retrieves
all MS3 (and subsequently MS4 spectra) for all these MS2 spectra using
the *precScanNum* spectra variable. The `ftms_all_levels` function
returns a combined `Spectra` object containing all MS2, MS3, and MS4
spectra, with each spectrum linked to a feature via a propagated
`feature_id` from the originally matched MS2 spectrum.

```{r}
ftms_all_levels <- function(ftms, ms2) {

  #ms2 <- featureSpectra(ftms, msLevel = 2)
  cat("Number of MS2 spectra found:", length(ms2), "\n")

  res <- lapply(unique(ms2$dataOrigin), function(origin) {
    ms2_subset <- filterDataOrigin(ms2, origin)

    ms3_filtered <- filterDataOrigin(filterMsLevel(spectra(ftms), 3), origin)
    ## To support n:m matches
    m <- findMatches(ms2_subset$acquisitionNum, ms3_filtered$precScanNum)
    cat("Number of MS3 matched to MS2:", length(m), "\n")
    ms3_filtered <- ms3_filtered[to(m)]
    ms3_filtered$feature_id <- ms2_subset$feature_id[from(m)]

    ms4_filtered <- filterDataOrigin(filterMsLevel(spectra(ftms), 4), origin)
    m <- findMatches(ms3_filtered$acquisitionNum, ms4_filtered$precScanNum)
    cat("Number of MS4 matched to MS3:", length(m), "\n")
    ms4_filtered <- ms4_filtered[to(m)]
    ms4_filtered$feature_id <- ms3_filtered$feature_id[from(m)]

    # Combine spectra for this origin
    c(ms2_subset, ms3_filtered, ms4_filtered)
  })
    do.call(c, res)
}

ftms_msn_tree <- ftms_all_levels(ftms,ms2)

table(msLevel(ftms_msn_tree))
```

The number of spectra from different MS levels is shown below:

```{r}
table(msLevel(ftms_msn_tree))
```

# Grouping MSn spectra into fragmentation trees

We next need to group all MSn spectra from one fragmentation tree
together.

The function `assign_msntree_id()` will:

-   Take a `Spectra` object `x` containing MS2–MSn spectra\
-   Assign tree IDs to each spectrum in `x`, based on precursor
    relationships within each sample (`dataOrigin`)\
-   Ensure that MS2 spectra always start a new tree, and that MS3/MS4
    spectra inherit the tree ID from their parent
-   Return a vector of integers, the same length as the number of
    spectra in `x`, where each integer represents the fragmentation tree
    ID the spectrum belongs to

The output of the `assign_msntree_id()` function is then added to the
`Spectra` object as a new spectra variable *MSntreeID*.

```{r}

assign_msntree_id <- function(x) {
  ## Order x by MS level: ensures we first assign an ID to the MS2.
  o <- order(msLevel(x))
  x <- x[o]
  scan_index <- scanIndex(x)
  prec_scan_num <- precScanNum(x)
  ms_level <- msLevel(x)
  data_origin <- dataOrigin(x)

  tree_id <- rep(NA_integer_, length(scan_index))
  global_counter <- 0

  for (origin in unique(data_origin)) {
    idx <- which(data_origin == origin)
    origin_scan_index <- scan_index[idx]
    origin_prec_scan_num <- prec_scan_num[idx]
    origin_ms_level <- ms_level[idx]
    origin_tree_id <- rep(NA_integer_, length(idx))

    scan_idx_map <- setNames(seq_along(origin_scan_index), origin_scan_index)

    for (i in seq_along(idx)) {
      level <- origin_ms_level[i]
      if (level == 2) {
        ## If MS2 hasn't been assigned yet
        if (is.na(origin_tree_id[i])) {
          global_counter <- global_counter + 1
          origin_tree_id[i] <- global_counter
        }
      } else {
        ## Assign ID of the related precursor ID
        parent_scan <- origin_prec_scan_num[i]
        parent_pos <- scan_idx_map[as.character(parent_scan)]
        if (length(parent_pos) == 1 && origin_ms_level[parent_pos] == (level - 1))
          origin_tree_id[i] <- origin_tree_id[parent_pos]
        else
          stop("MS", level, " with ", length(parent_pos), " precursors")
      }
    }
    tree_id[idx] <- origin_tree_id
  }
  ## Return in original order
  return(tree_id[order(o)])
}

msntree_ids <- assign_msntree_id(ftms_msn_tree)

ftms_msn_tree$MSntreeID <- msntree_ids
spectraVariables(ftms_msn_tree)

```

The total number of MSn trees is 3508:

```{r}
length(unique(ftms_msn_tree$MSntreeID))
```

for in total `r length(unique(ftms_msn_tree$feature_id))` features. The
distribution of the lengths of the fragmentation trees is

```{r}
ftms_msn_tree$MSntreeID |>
    table() |>
    quantile()
```

There is apparently (at least) one tree with 4 spectra. Extracting this
particular MSn tree.

```{r}
max_index <- ftms_msn_tree$MSntreeID |>
    table() |>
    as.integer() |>
    which.max()

max_tree <- ftms_msn_tree[ftms_msn_tree$MSntreeID == max_index]
max_tree$rtime
max_tree$msLevel
max_tree$precScanNum
max_tree$scanIndex
max_tree$acquisitionNum
max_tree$dataOrigin
max_tree$feature_id
max_tree$MSntreeID
```

The MSntree with maximum number of spectra corresponds to the tree,
within dataOrigin S1R1, depending on MS2 with ScanIndex 2182 which was
assigned to one FeatureID: FT2705 Within dataOrigin S1R1, scanIndex 2183
(MS3), 2184 (MS3), and 2185 (MS4) are assigned to a single feature:
FT2705.

```{r}
## we stored the tree ids and the msLevels to a dataframe
df <- spectraData(ftms_msn_tree, c("MSntreeID", "msLevel")) |>
    as.data.frame()

## here we take count the last level of each tree
library(dplyr)
max_level_per_tree <- df %>%
  group_by(MSntreeID) %>%
  summarise(max_level = max(msLevel), .groups = "drop")


table(max_level_per_tree$max_level)

```

The number of trees that go up to MS2 level is : 1808 The number of
trees that go up to MS3 level is : 404 The number of trees that go up to
MS4 level is : 1339

Next we select one most representative tree per feature based on the
longest one and if we have many, we select one tree with the highest MS2
precursor intensity

```{r}

unique_tree_feature <- function(db) {
  # Store MS2 indices
  ms2_idx <- which(msLevel(db) == 2)

  # Table of linked MS2 features only
  feat_ms2 <- data.frame(
    idx = ms2_idx,
    feature_id = sapply(db$feature_id[ms2_idx], function(x) x[1]),
    MSntreeID = db$MSntreeID[ms2_idx],
    precursorIntensity = db$precursorIntensity[ms2_idx]
  )

  # for each feature choose the longest tree and the one with the highest
  #precursorIntensity()  in case of tie
  selected_tree_ids <- tapply(seq_len(nrow(feat_ms2)), feat_ms2$feature_id, function(i) {
    #the longest tree
    tbl <- table(feat_ms2$MSntreeID[i])
    FTmax <- as.numeric(names(tbl)[tbl == max(tbl)])
    #the highest precursorIntensity()
    sub_idx <- i[feat_ms2$MSntreeID[i] %in% FTmax]
    sub_idx[which.max(feat_ms2$precursorIntensity[sub_idx])]
  })

  tree_ids_keep <- feat_ms2$MSntreeID[unlist(selected_tree_ids)]
  db[which(db$MSntreeID %in% tree_ids_keep)]
}

ftms_one_tree <- unique_tree_feature(ftms_msn_tree)

msLevel(ftms_one_tree) |> table()
```

after selecting one representative tree per feature we have now: MS2 :
715 and MS3 : 615 and MS4 : 292

Finally we verify if the number of unique features corresponds to the
number of unique MSntreeID to ensure that we have one tree per feature.

```{r}
length(unique(unlist(ftms_one_tree$feature_id)))
```

```{r}
length(unique(ftms_one_tree$MSntreeID))
```

we have 715 unique feature and unique MSntreeID, That's mean that we
have for each feature one representative tree.

Now we save the resulting ftms spectra object to a local directory.

```{r}
save(ftms_one_tree, file = file.path(pth, "data", "unique_tree_ftms_ns.RData"))
```

# Session information

```{r}
sessionInfo()
```
