---
title: "Create an assembled MS/MS data from Ftms MSn data"
format:
  html:
    toc: true
    self-contained: true
author: "Ahlam Mentag, Johannes Rainer"
editor:
  markdown:
    wrap: 72
---

# Introduction

In this quarto document we create an assembled MS/MS data from ftms
object that contains one tree per feature created in the
[ftms_filtering.qmd](https://github.com/rebdau/RDynLib_ftms/blob/ahlam/ftms_filtering.qmd)
file.

# Data import

Loading all required libraries and importing the results from the ftms
filtered object.

```{r}
#| message: false
library(dplyr)
library(tidyr)
library(xcms)
library(Spectra)

register(SerialParam())
pth <- getwd()
pth
```

First we load the `Spectra` ftms object with one representative tree per
feature.

```{r}
load(file.path(pth, "data", "ftms_unique_tree.RData"))
class(ftms_one_tree)
```

Next we round `mz` values to the first integer using `round_perl()`
function

```{r}
round_perl <- function(number) {
  floor(number + 0.5)
}

peaks_round_mz <- function(x, spectrumMsLevel, ...) {
    x[, 1L] <- round_perl(x[, 1L])
    x
}

ftms_one_tree <- addProcessing(ftms_one_tree, peaks_round_mz)


```

keep only the most intense peak if the same mz is repeated in the same
spectrum

```{r}
#'Clean a single spectrum
#'keep only the most intense peak if the same mz is repeated
clean_spectrum <- function(spectrum) {
  as.data.frame(spectrum) %>%
    group_by(mz) %>%
    slice_max(order_by = intensity, n = 1, with_ties = FALSE) %>%
    arrange(mz) %>%
    as.matrix()
}

```

Combine across levels (sum intensities if same mz)

```{r}
#'Combine spectra across a tree
#'sum intensities if the same mz appears in different spectra
combine_spectra <- function(spectra_list) {
  do.call(rbind, spectra_list) %>%
    as.data.frame() %>%
    group_by(mz) %>%
    summarise(intensity = sum(intensity), .groups = "drop") %>%
    arrange(mz) %>%
    as.matrix()
}
```

final function to create a new assembled spectra object

```{r}

create_assembled_spectra <- function(ftms) {
  #'Converting the backend to MsBackendMemory 
  #'so we could modify the peaksData
  ftms <- setBackend(ftms, MsBackendMemory())
  meta <- spectraData(ftms)
  pd_list <- peaksData(ftms)
  tree_ids <- unique(meta$MSntreeID)
  
  assembled_list <- list()
  
  for (tree_id in tree_ids) {
    idx <- which(meta$MSntreeID == tree_id)
    ms_levels <- meta$msLevel[idx]
    
    #'First clean each spectrum individually 
    #'(remove duplicate mz by keeping the max intensity)
    cleaned_spectra <- lapply(pd_list[idx], clean_spectrum)
    
    #'If a tree has only MS2 level, we copy the MS2
    #'row to the new spectra object as is it (after cleaning)
    if (max(ms_levels) <= 2L) {
      ms2_row <- idx[ms_levels == 2][1]
      ms2_meta <- meta[ms2_row, , drop = FALSE]
      df <- ms2_meta
      #'Mark it as assembled
      df$spectrum.type <- "assembled"
      df$MSntreeID <- tree_id
      #copy the same mz and intensity (already cleaned)
      df$mz <- list(cleaned_spectra[[which(ms_levels == 2)[1]]][, "mz"])
      df$intensity <- list(cleaned_spectra[[which(ms_levels == 2)[1]]][, "intensity"])
      #'Convert the modified metadata+peaks into a new 
      #'Spectra object and append it to assembled_list
      assembled_list[[length(assembled_list) + 1]] <- Spectra(df)
      next
    }
    
    #'If a tree has more than MS2 level
    #'Combine all cleaned spectra by summing intensities
    merged_spec <- combine_spectra(cleaned_spectra)
    
    #'Copy spectraData from the MS2 level 
    #'of each tree to the new spectra object
    ms2_row <- idx[ms_levels == 2][1]
    ms2_meta <- meta[ms2_row, , drop = FALSE]
    df <- ms2_meta
    df$spectrum.type <- "assembled"
    df$MSntreeID <- tree_id
    #'assign the assembled spectrum to the new spectra object
    df$mz <- list(merged_spec[, "mz"])
    df$intensity <- list(merged_spec[, "intensity"])
    
    assembled_list[[length(assembled_list) + 1]] <- Spectra(df)
  }
  
  if (length(assembled_list) == 0) {
    warning("No assembled spectra found!")
    return(NULL)
  }
  
  do.call(c, assembled_list)
}



```

Calling the main function

```{r}
ftms_assembled <- create_assembled_spectra(ftms_one_tree)

spectraData(ftms_assembled)
```

the `spectraData()` contains 511 rows which correspond to the number of
unique trees in the global object.

example for duplicated peaks in different MS levels of a tree, first we
print the `spectraData()` and `peaksData()` of the 1813 tree before
assembling

```{r}
#spectraData
spectraData(ftms_one_tree)[spectraData(ftms_one_tree)$MSntreeID == 1072,]

#peaksData
idx <- which(spectraData(ftms_one_tree)$MSntreeID == 1072)
lapply(peaksData(ftms_one_tree)[idx], head) 
```

This tree has 4 MS levels (1 MS2, 2 MS3, 1 MS4). One MS3 peak has the
same m/z as the MS2 peak.

Now we print the `spectraData()` and `peaksData()` of the 1813 tree
after assembling

```{r}
#spectraData
spectraData(ftms_assembled)[spectraData(ftms_assembled)$MSntreeID == 1072, ]

#peaksData
idx <- which(spectraData(ftms_assembled)$MSntreeID == 1072)
lapply(peaksData(ftms_assembled)[idx], head)
```

The `spectraData()` of this tree after assembling contains only the MS2
`spectraData()`, as the other levels were dropped. Moreover, the
assembled `peaksData()` of this tree contains 4 pairs of (mz, intensity)
instead of 5 in `ftms_one_tree`, because the m/z of the MS3 level is
equal to that of the MS4 level; as a result, they were combined into one
by summing their intensities.

Other example for duplicated peaks in the same MS level of a tree

```{r}
#spectraData
spectraData(ftms_one_tree)[spectraData(ftms_one_tree)$MSntreeID == 483,]

#peaksData
idx <- which(spectraData(ftms_one_tree)$MSntreeID == 483)
lapply(peaksData(ftms_one_tree)[idx], head) 
```

This tree has 4 MS levels (1 MS2, 2 MS3, 1 MS4). 2 MS3 peaks has the
same m/z .

Now we print the `spectraData()` and `peaksData()` of the 483 tree after
assembling

```{r}
#spectraData
spectraData(ftms_assembled)[spectraData(ftms_assembled)$MSntreeID == 483, ]

#peaksData
idx <- which(spectraData(ftms_assembled)$MSntreeID == 483)
lapply(peaksData(ftms_assembled)[idx], head)
```

so the results shows that for the duplicated mz in both MS3 levels of
this tree, the algorithm keep just one and sum their intensities.

Saving the result to a `RData` object

```{r}
save(ftms_assembled, file = file.path(pth, "data", "ftms_assembled.RData"))
```
