---
title: "Untitled"
format: html
server: shiny
editor_options:
  chunk_output_type: inline
---

## Shiny Documents

In this document, we present all the main steps for analyzing FTMS data, including data visualization, preprocessing, XCMS processing, and retention time alignment. We also plotted the extracted ion chromatograms of some known compounds.

```{r}
library(MsExperiment)
library(Spectra)
library(dplyr)
library(tidyr)
library(BiocParallel)
library(xcms)
library(MsExperiment)
library(Spectra)
library(readxl)
library(pander)
library(pheatmap)
register(SerialParam())
```

now we import our dataset

```{r}
sample_data_ftms <- read_xlsx("C:/Users/amentag/Desktop/Rcodes/RDynLib_ftms/data/ftms_data_files.xlsx") |>
    as.data.frame()
file <- "C:/Users/amentag/Desktop/Rcodes/RDynLib_ftms/data/mzML/"
ftms <- readMsExperiment(paste0(file, sample_data_ftms$data_file),
                         sampleData = sample_data_ftms)
ftms
```



```{r}
sample_data_ftms
```

```{r}
pandoc.table(as.data.frame(sampleData(ftms)), caption = "Data files",
             split.table = Inf, style = "rmarkdown")
```

```{r}
msLevel(spectra(ftms)) |>
  table()
```

Defining colors for each sample

```{r}
library(RColorBrewer)
# colors
col_sample_ftms_id <- brewer.pal(9, "Set1")[c(1, 2)]
names(col_sample_ftms_id) <- unique(sample_data_ftms$sample_id)
# Assigne colors for each single name in the sample_id
names(col_sample_ftms_id) <- c("S1", "S9")
col_sample_ftms <- col_sample_ftms_id[sample_data_ftms$sample_id]
```

BPC Chromatogram

```{r}
#| fig-width: 10
#| fig-height: 5
bpc0 <- chromatogram(ftms, aggregationFun = "max")
plot(bpc0, col = paste0(col_sample_ftms, 80))
grid()
```

This BPC shows that there is no signal detected after 1500s,

We next clean the data set filtering to a retention time range of 0-1500 seconds.

```{r}
ftms <- filterSpectra(ftms, filterRt, c(10, 3500))
msLevel(spectra(ftms)) |>
 table()

```

This data contains 4 MS levels but with empty spectra so we removed them using `filterEmptySpectra()` function

```{r}
ftms <- filterSpectra(ftms, filterEmptySpectra)
msLevel(spectra(ftms)) |>
  table()
```

```{r}
#| fig-width: 10
#| fig-height: 5
bpc0 <- chromatogram(ftms, aggregationFun = "max")
plot(bpc0, col = paste0(col_sample_ftms, 80))
grid()
```

We next extract a total ion chromatogram (TIC) and base peak chromatogram (BPC) and calculate the similarities between these to evaluate similarity in the performance of the liquid chromatography.

```{r}
#' Extract TIC
tic <- chromatogram(ftms, aggregationFun = "sum")
#' Bin the signal
ticb <- bin(tic, binSize = 150)

ticmap <- ticb |>
    lapply(intensity) |>
    do.call(what = rbind) |>
    log2()

colnames(ticmap) <- rtime(ticb[[1]])
rownames(ticmap) <- sampleData(ftms)$sample_name

#' Center the data per bin by subtracting the median
ticmap <- scale(ticmap, center = TRUE, scale = FALSE)

ann <- data.frame(sample_id = sampleData(ftms)[, "sample_id"])
rownames(ann) <- rownames(ticmap)

pheatmap(ticmap, cluster_cols = FALSE,
         annotation_colors = list(sample_id = col_sample_ftms_id),
         annotation_row = ann)
```

This heatmap shows that S9 samples are better grouped compared with the S1 samples?

Now we plot the tic

```{r}
#| fig-width: 10
#| fig-height: 5
plot(tic, col = paste0(col_sample_ftms, 80), main = "TIC")
grid()
```

We repeat this analysis also for the BPC.

```{r}
#' Extract BPC
bpc <- chromatogram(ftms, aggregationFun = "sum")
#' Bin the signal
bpcb <- bin(bpc, binSize = 150)

bpcmap <- bpcb |>
    lapply(intensity) |>
    do.call(what = rbind) |>
    log2()

colnames(bpcmap) <- rtime(bpcb[[1]])
rownames(bpcmap) <- sampleData(ftms)$sample_name

#' Center the data per bin by subtracting the median
bpcmap <- scale(bpcmap, center = TRUE, scale = FALSE)
pheatmap(bpcmap, cluster_cols = FALSE,
         annotation_colors = list(sample_ftms_id = col_sample_ftms_id),
         annotation_row = ann)
```

This bpc heatmap is like the tic heatmap, it shows that S9 samples are better grouped compared with the S1 samples?

Now we plot the BPC again

```{r}
#| fig-width: 10
#| fig-height: 5
plot(bpc, col = paste0(col_sample_ftms, 80), main = "BPC")
grid()
```

# Data preprocessing

here we take an mz range of an known compound

```{r}
mzr <- c(287.04, 287.06)

a_kn <- chromatogram(ftms, mz = mzr, aggregationFun = "max")
plot(a_kn, col = paste0(col_sample_ftms, 80))
grid()

```

we zoom into a small retention time range that contains the highest peak signal

```{r}
#' Focus on the retention time range with the highest signal
plot(a_kn, xlim = c(1300, 1500))
grid()

#' Plot it for a single sample
plot(a_kn[, 1], xlim = c(1300, 1500))
grid()
```

```{r}
tmp <- filterMz(ftms, mz = mzr) |>
    filterRt(rt = c(1400, 1440))
plot(tmp[1])
```

Here we do the same thing as qtof to choose the peakwidth parameter, but here we used the ion chromatogram of a known compound, its width is about 40 seconds so we take the half and the double that’s mean 20 and 80 as a parameter window.

Next we run the chromatographic peak detection on that extracted ion chromatogram

```{r}
a_kn2 <- findChromPeaks(a_kn, CentWaveParam(peakwidth = c(20, 80), integrate = 2))

```

```{r}
pks <- chromPeaks(a_kn2)
head(pks)
quantile(unname(pks[, "rtmax"] - pks[, "rtmin"]))
```

```{r}
#| fig-width: 10
#| fig-height: 5

par(mfrow = c(1, 2))
#' In one sample
plot(a_kn2[, 1], xlim = c(1390, 1450))
grid()
#' In all samples
plot(a_kn2[, 1:5], xlim = c(1390, 1450))
grid()
```

the same on another retention time range

```{r}
#| fig-width: 10
#| fig-height: 5

par(mfrow = c(1, 2))
plot(a_kn2[, 1], xlim = c(870, 1000))
grid()
plot(a_kn2[, 1:5], xlim = c(870, 1000))
grid()
```

```{r}
#| fig-width: 10
#| fig-height: 5

par(mfrow = c(1, 2))
plot(a_kn2[, 1], xlim = c(0, 500))
grid()
plot(a_kn2[, 1:5], xlim = c(0, 500))
grid()
```

Eventually we could try another m/z range...

```{r}
plot(bpc)
grid()
abline(v = c(1020, 1100), col = "grey")
plot(bpc[, 1], xlim = c(1020, 1100))
```

Below we extract the ion chromatogram for this m/z range.

```{r}
mzr <- c(287.04, 287.06)

a <- chromatogram(ftms, mz = mzr, aggregationFun = "max")
plot(a, col = paste0(col_sample_ftms, 80))
grid()
```

This m/z range seems to contain signal from several ions, eluting at different retention times. We next focus on a retention time range to inspect the signal from a single ion.

```{r}
plot(a, col = paste0(col_sample_ftms, 80), xlim = c(1400, 1500))
grid()
```

We next perform the chromatographic peak detection using the centWave algorithm. With integrate = 2 we use an alternative algorithm to correctly identify the boundaries of the identified chromatographic peaks. Parameter chunkSize is used to control the number of files from which the data should be loaded into memory at a time.

```{r}
ftms <- findChromPeaks(ftms, CentWaveParam(peakwidth = c(20, 80), integrate = 2),
                       chunkSize = 2)
```

With this setting we identified `r nrow(chromPeaks(ftms))` peaks in the full data set. The number of peaks per sample are:

```{r}
pks <- chromPeaks(ftms)
table(pks[, "sample"])
```

The distribution of the peak widths in retention time and in m/z dimensions are:

```{r}
quantile(unname(pks[, "rtmax"] - pks[, "rtmin"]))
quantile(unname(pks[, "mzmax"] - pks[, "mzmin"]))
```

We next perform the chromatographic peak refinement to reduce the number of potential centWave-specific peak detection artifacts. We choose settings that depend on the observed peak widths above (i.e. half of the observed average widths).

```{r}
mnpp <- MergeNeighboringPeaksParam(expandRt = 9.67525, expandMz = 0.0006)
ftms <- refineChromPeaks(ftms, mnpp)
```

We evaluate the number of peaks and the observed peak widths also after refinement.

```{r}
pks <- chromPeaks(ftms)
table(pks[, "sample"])
quantile(unname(pks[,"rtmax"] - pks[,"rtmin"]))
quantile(unname(pks[,"mzmax"] - pks[,"mzmin"]))
```

We next perform an initial correspondence analysis that is needed for the subsequent retention time alignment. We evaluate the settings on the example m/z range from above.

```{r}
#| fig-width: 10
#| fig-height: 5

#' Extract the ion chromatogram again.
a_2 <- chromatogram(ftms, aggregationFun = "max", mz = mzr)

#' Configure settings

pdp <- PeakDensityParam(sampleGroups = sampleData(ftms)$sample_id,
                        bw = 6, minFraction = 0.5)

#' Evaluate settings on the full range
plotChromPeakDensity(
    a_2, param = pdp, col = paste0(col_sample_ftms, 80),
    peakCol = paste0(col_sample_ftms[chromPeaks(a_2)[, "sample"]], 80),
    peakBg = paste0(col_sample_ftms[chromPeaks(a_2)[, "sample"]], 40),
    peakPch = 16)
grid()
```

now we will focus on a smaller retention time range

```{r}
a_sub <- filterRt(a_2, rt = c(1300, 1500))
plotChromPeakDensity(
    a_sub, param = pdp,
    peakCol = paste0(col_sample_ftms[chromPeaks(a_sub)[, "sample"]], 80),
    peakBg = paste0(col_sample_ftms[chromPeaks(a_sub)[, "sample"]], 40),
    peakPch = 16)
grid()
```

```{r}
ftms <- groupChromPeaks(ftms, param = pdp)
```

we next perform retention time alignment

```{r}
pgp <- PeakGroupsParam(
    minFraction = 0.9,
    extraPeaks = 100,
    smooth = "loess",
    span = 0.4,
    family = "gaussian"
)
ftms <- adjustRtime(ftms, param = pgp)
```

Below we plot the results from the retention time alignment, i.e., the difference between the original and adjusted retention times (y-axis) along the retention time axis (x-axis).

```{r}
plotAdjustedRtime(ftms, col = paste0(col_sample_ftms, 80))
```

we can see from the plot above that there is one sample that has been largely adjusted, so it goes up to 15 seconds.

```{r}
bpc_adj <- chromatogram(ftms, aggregationFun = "max")
#' Calculate similarity (Pearson correlation) between BPCs
#' Bin the signal
bpcb <- bin(bpc_adj, binSize = 150)

bpcmap <- bpcb |>
    lapply(intensity) |>
    do.call(what = rbind)

bpcmap[bpcmap <= 0] <- NA 

bpcmap <- log2(bpcmap)

colnames(bpcmap) <- rtime(bpcb[[1]])
rownames(bpcmap) <- sampleData(ftms)$sample_name

#' Center the data per bin by subtracting the median
bpcmap <- scale(bpcmap, center = TRUE, scale = FALSE)
pheatmap(bpcmap, cluster_cols = FALSE,
         annotation_colors = list(sample_ftms_id = col_sample_ftms_id),
         annotation_row = ann)
```

we see that S1 and S9 are well grouped after retention time alignment.

Now we can visualize the changes before and after retention time alignment

```{r}
bpc_adj <- chromatogram(ftms, aggregationFun = "max", chromPeaks = "none")
a_adj <- chromatogram(ftms, aggregationFun = "max", mz = mzr)
```

We first plot the BPC of the original data and then of the data after retention time alignment.

```{r}
#| fig-width: 10
#| fig-height: 10

par(mfrow = c(2, 1))
plot(bpc0, col = paste0(col_sample_ftms, 80), main = "raw")
grid()
plot(bpc_adj, col = paste0(col_sample_ftms, 80),
     main = "retention time adjusted")
grid()

```

And we plot them also separately

```{r}
par(mfrow = c(1, 1))
plot(bpc0, col = paste0(col_sample_ftms, 80))
grid()
```

```{r}
par(mfrow = c(1, 1))
plot(bpc_adj, col = paste0(col_sample_ftms, 80))
grid()
```

the data looks well aligned, next we need to visualize the alignment result on a smaller retention time range

```{r}
par(mfrow = c(1, 1))
plot(bpc0, col = paste0(col_sample_ftms, 80), xlim = c(400, 500))
grid()

```

```{r}
par(mfrow = c(1, 1))
plot(bpc_adj, col = paste0(col_sample_ftms, 80), xlim = c(400, 500))
grid()
```

the data looks well aligned, From this results we can conclude that retention time alignment works well on this data.

We in addition evaluate the signal for the example m/z range.

```{r}
par(mfrow = c(2, 1))
plot(a_2, col = paste0(col_sample_ftms, 80))
grid()
plot(a_adj, col = paste0(col_sample_ftms, 80))
grid()
```

the data seems to be better aligned. Finally, we zoom into a retention time window from 1300 to 1500 seconds.

```{r}
par(mfrow = c(2, 1))
plot(a_2, col = paste0(col_sample_ftms, 80), peakType = "none",
     xlim = c(1300, 1500))
plot(a_adj, col = paste0(col_sample_ftms, 80),
     xlim = c(1300, 1500), peakType = "none")
```

We can thus conclude that the settings for the retention time alignment worked on the present data set. We continue the preprocessing with the final correspondence analysis. We adapt now the settings, in particular the bw parameter, that can be much stricter because of the properly aligned data set. Again, we test the settings on the extracted ion signal.

```{r}
#' Configure settings
pdp <- PeakDensityParam(sampleGroups = sampleData(ftms)$sample_id,
                        bw = 2, minFraction = 0.5)

#' Evaluate settings on the full range
plotChromPeakDensity(a_adj, param = pdp, col = col_sample_ftms)
```

we zoom into a smaller region with multiple peaks

```{r}
a_adj_2 <- filterRt(a_adj, c(800, 1500))
plotChromPeakDensity(a_adj_2, param = pdp, col = col_sample_ftms)
```

```{r}
a_adj_2 <- filterRt(a_adj, c(1250, 1450))
plotChromPeakDensity(a_adj_2, param = pdp, col = col_sample_ftms)

```

now we group chromatographic peak

```{r}
ftms <- groupChromPeaks(ftms, param = pdp)
```

As a final step we next perform gap-filling to reduce the number of missing values in the data set.

```{r}
#' The number of missing values before gap filling
sum(is.na(featureValues(ftms)))
head(featureValues(ftms))
```

```{r}
#' Perform gap-filling
ftms <- fillChromPeaks(ftms, param = ChromPeakAreaParam())
```

```{r}
sum(is.na(featureValues(ftms)))
head(featureValues(ftms))
```

```{r}
#' Save the result object
save(ftms, file = "C:/Users/amentag/Desktop/Rcodes/RDynLib_ftms/data/ftms_final2.RData")
```

```{r}
load(file = "C:/Users/amentag/Desktop/Rcodes/RDynLib_ftms/data/ftms_final2.RData")
```

```{r}
nrow(featureDefinitions(ftms))
```

```{r}
msLevel(spectra(ftms)) |>
  table()
```


We next save the raw and adjusted BPC to a image file.

```{r}

#' Define the file path where to save the image
dr <- "images/ftms_glob/"
dir.create(dr, showWarnings = FALSE, recursive = TRUE)
png(file.path(dr, "ftms-bpc-aligned.png"), width = 14, height = 10,
    units = "cm", res = 600, pointsize = 4)

par(mfrow = c(2, 1))
plot(bpc0, col = paste0(col_sample_ftms, 80), main = "Raw data")
grid()
plot(bpc_adj, col = paste0(col_sample_ftms, 80),
     main = "Retention time aligned")
grid()

dev.off()
```
Now we will create the feature matrix of this data.

```{r}
corrected_features <- featureDefinitions(ftms)
#here we add a new column called featue in this format "MxTy" with:
#x: mzmed and y: rtmed of each feature

final_feature_matrix <- data.frame(
 `rt(min)` = corrected_features$rtmed / 60, 
  mzmed = corrected_features$mzmed,         
  feature_name = paste0(                     
    "M", round(corrected_features$mzmed, 0), 
    "T", round(corrected_features$rtmed, 0)
  )
)


```

```{r}





```

Now let's print the head of the resulting feature matrix

```{r}
head(final_feature_matrix)
```

```{r}
nrow(final_feature_matrix)
```




# Session information

The R version and package versions used for this analysis were:

```{r}
sessionInfo()
```
