---
title: "Selecting one representative tree per feature"
format:
  html:
    toc: true
    self-contained: true
author: "Ahlam Mentag, Johannes Rainer"
editor:
  markdown:
    wrap: 72
---

# Introduction

This Quarto document presents the analysis of FTMS flaxseed data for
selecting one representative tree per feature. Preprocessing and feature
definition was performed in [Preprocessing steps](ftms_xcms_short.qmd).

# Data import

Loading all required libraries and importing the results from the data
preprocessing.

```{r}
#| message: false
library(dplyr)
library(tidyr)
library(xcms)
library(MsExperiment)
library(Spectra)
library(readxl)
library(IRanges)
library(S4Vectors)
register(SerialParam())
pth <- getwd()
pth
```

First we load the `XcmsExperiment` ftms object with the preprocessing
results.

```{r}
load(file.path(pth, "data", "ftms.RData"))
class(ftms)
```

```{r}
ftms
library(Spectra)

sp <- spectra(ftms)

```

# Associate MS2 spectra to features

Next, we use `xcms::featureSpectra()` to extract all MS2 spectra for
each feature from `ftms` object.

> The `xcms::featureSpectra()` function returns a `Spectra` object.
>
> When using `msLevel = 2`, the resulting `Spectra` object contains all
> MS2 spectra whose retention time and precursor m/z fall within the
> ranges of any chromatographic peak of a feature.

```{r}
ms2 <- featureSpectra(dropFilledChromPeaks(ftms), msLevel = 2, skipFilled = TRUE)

msLevel(ms2) |>
  table()
```

it returns `r length(ms2)` MS level 2 spectra. The total number of
features for which an MS2 spectrum was identified is:

```{r}
length(unique(ms2$feature_id))
length(ms2$feature_id)
```

Here we print the initial number of MS levels.

```{r}
s <- spectra(ftms)
table(msLevel(s))
```

In total, there were `r table(msLevel(s))[2]` MS2 spectra. This means
that
`r round((table(msLevel(s))[2]-length(ms2$feature_id))*100/table(msLevel(s))[2], 1)`
% of the MS2 spectra could not be assigned to any chromatographic peak.

The percentage of features with at least one MS2:

```{r}
# Total number of features in ftms
n_total_features <- nrow(featureDefinitions(ftms))

# Features with ≥1 MS2 (those returned by featureSpectra)
n_features_with_ms2 <- length(unique(ms2$feature_id))

# Percentage
perc_features_with_ms2 <- (n_features_with_ms2 / n_total_features) * 100
perc_features_with_ms2

```

The total number of features with at least one MS2 is
`r n_features_with_ms2`, which corresponds to
`r format(perc_features_with_ms2, digits = 3)`%.

A summary on the number of MS2 spectra assigned to a feature is shown
below:

```{r}
table(ms2$feature_id) |>
    quantile()
```

The median number of MS2 spectra per feature (across all data files) is
hence `r median(table(ms2$feature_id))`.

the percentage of MS2 spectra that are assigned to a feature

```{r}
# Total number of MS2 spectra in the dataset
n_total_ms2 <- sum(msLevel(s) == 2)
sp <- spectraData(ms2)
sp$groups <- paste0(sp$dataOrigin, "_", sp$scanIndex)
# Number of MS2 spectra assigned to features
n_assigned_ms2 <- length(unique(sp$groups))

# Percentage assigned
perc_ms2_assigned <- (n_assigned_ms2 / n_total_ms2) * 100
perc_ms2_assigned

```

The total number of assigned MS2 spectra is `r n_assigned_ms2`, which
corresponds to a percentage of
`r format(perc_ms2_assigned, digits = 3)`.

Note that an MS2 spectrum could also be assigned to more than one
feature. The total number of MS2 spectra assigned to more than one
feature is:

```{r}
ms2_id <- paste0(ms2$dataOrigin, "_", ms2$acquisitionNum)
ms2_table <- table(ms2_id)
ms2_mult <- names(ms2_table)[ms2_table > 1]

```

Of the `r length(unique(ms2_id))` MS2 spectra, `r length(ms2_mult)` are
assigned to more than one feature.

# Extract all MS2-MSn spectra associated to features

We next define a function to extract the full MSn tree for features.
This function takes an `XcmsExperiment` object as input, extracts the
MS2 spectra using `xcms::featureSpectra()` function, and then retrieves
all MS3 (and subsequently MS4 spectra) for all these MS2 spectra using
the *precScanNum* spectra variable. The `ftms_all_levels` function
returns a combined `Spectra` object containing all MS2, MS3, and MS4
spectra, with each spectrum linked to a feature via a propagated
`feature_id` from the originally matched MS2 spectrum.

```{r}
ftms_all_levels <- function(ftms, ms2) {

  #ms2 <- featureSpectra(ftms, msLevel = 2)
  cat("Number of MS2 spectra found:", length(ms2), "\n")

  res <- lapply(unique(ms2$dataOrigin), function(origin) {
    ms2_subset <- filterDataOrigin(ms2, origin)

    ms3_filtered <- filterDataOrigin(filterMsLevel(spectra(ftms), 3), origin)
    ## To support n:m matches
    m <- findMatches(ms2_subset$acquisitionNum, ms3_filtered$precScanNum)
    cat("Number of MS3 matched to MS2:", length(m), "\n")
    ms3_filtered <- ms3_filtered[to(m)]
    ms3_filtered$feature_id <- ms2_subset$feature_id[from(m)]

    ms4_filtered <- filterDataOrigin(filterMsLevel(spectra(ftms), 4), origin)
    m <- findMatches(ms3_filtered$acquisitionNum, ms4_filtered$precScanNum)
    cat("Number of MS4 matched to MS3:", length(m), "\n")
    ms4_filtered <- ms4_filtered[to(m)]
    ms4_filtered$feature_id <- ms3_filtered$feature_id[from(m)]

    # Combine spectra for this origin
    c(ms2_subset, ms3_filtered, ms4_filtered)
  })
    do.call(c, res)
}

ftms_msn_tree <- ftms_all_levels(ftms,ms2)

table(msLevel(ftms_msn_tree))
```

The number of spectra from different MS levels is shown below:

```{r}
table(msLevel(ftms_msn_tree))
```

# Grouping MSn spectra into fragmentation trees

We now group all spectra that belong to the same fragmentation tree and record the precursor m/z of the originating MS2 spectrum for each spectrum within that tree.


The function `assign_msntree_id()` will:

-   Take a `Spectra` object `x` containing MS2–MSn spectra\
-   Assign tree IDs to each spectrum in `x`, based on precursor
    relationships within each sample (`dataOrigin`)\
-   Ensure that MS2 spectra always start a new tree, and that MS3/MS4
    spectra inherit the tree ID from their parent\
-   Propagate the MS2-level precursor m/z to all downstream spectra in
    the same tree (MS3, MS4, …), with `NA` for MS2 spectra\
-   Return the updated `Spectra` object with two new spectra variables:

    -   **`MSntreeID`** — integer ID of the fragmentation tree each
        spectrum belongs to\
    -   **`precursorMz.MS2`** — MS2-level precursor m/z propagated to
        MS3/MS4 (and beyond), `NA` for MS2
        
!!!!!!!!!!!!!TODO: fix function, precursorMz.MS2 is wrong; needs to be == precursorMz if msLevel == 3
```{r}
assign_msntree_id <- function(x) {
  ## Order x by MS level: ensures we first assign an ID to the MS2.
  o <- order(msLevel(x))
  x <- x[o]
  scan_index <- scanIndex(x)
  prec_scan_num <- precScanNum(x)
  ms_level <- msLevel(x)
  data_origin <- dataOrigin(x)
  
  ## store precursorMz values
  precursor_mz <- precursorMz(x)

  tree_id <- rep(NA_integer_, length(scan_index))
  ## initialize output vector for MS2-level precursor m/z
  precursor_mz_MS2 <- rep(NA_real_, length(scan_index))
  
  global_counter <- 0

  for (origin in unique(data_origin)) {
    idx <- which(data_origin == origin)
    origin_scan_index <- scan_index[idx]
    origin_prec_scan_num <- prec_scan_num[idx]
    origin_ms_level <- ms_level[idx]
    ## per-origin precursorMz vector
    origin_precursor_mz <- precursor_mz[idx]
    origin_tree_id <- rep(NA_integer_, length(idx))
    ## per-origin precursorMz.MS2 vector
    origin_precursor_mz_MS2 <- rep(NA_real_, length(idx))

    scan_idx_map <- setNames(seq_along(origin_scan_index), origin_scan_index)

    for (i in seq_along(idx)) {
      level <- origin_ms_level[i]
      if (level == 2) {
        ## If MS2 hasn't been assigned yet
        if (is.na(origin_tree_id[i])) {
          global_counter <- global_counter + 1
          origin_tree_id[i] <- global_counter
        }
        ## MS2 spectra have no MS2-level precursor
        origin_precursor_mz_MS2[i] <- NA_real_

      } else {
        ## Assign ID of the related precursor ID
        parent_scan <- origin_prec_scan_num[i]
        parent_pos <- scan_idx_map[as.character(parent_scan)]
        if (length(parent_pos) == 1 && origin_ms_level[parent_pos] == (level - 1)) {
          origin_tree_id[i] <- origin_tree_id[parent_pos]

          ## assign precursorMz.MS2 depending on depth
          if (level == 3) {
            ## MS3: parent is MS2; use parent’s precursorMz
            origin_precursor_mz_MS2[i] <- origin_precursor_mz[parent_pos]
          } else if (level == 4) {
            ## MS4: grandparent is MS2; fetch its precursorMz
            grandparent_scan <- origin_prec_scan_num[parent_pos]
            grandparent_pos <- scan_idx_map[as.character(grandparent_scan)]
            if (length(grandparent_pos) == 1 && origin_ms_level[grandparent_pos] == 2)
              origin_precursor_mz_MS2[i] <- origin_precursor_mz[grandparent_pos]
            else
              origin_precursor_mz_MS2[i] <- NA_real_
          } else {
            ## Optional: handle MS5+ recursively (rare)
            ancestor_pos <- parent_pos
            while (ancestor_pos > 0 &&
                   origin_ms_level[ancestor_pos] > 2) {
              parent_scan <- origin_prec_scan_num[ancestor_pos]
              ancestor_pos <- scan_idx_map[as.character(parent_scan)]
            }
            if (!is.na(ancestor_pos) && origin_ms_level[ancestor_pos] == 2)
              origin_precursor_mz_MS2[i] <- origin_precursor_mz[ancestor_pos]
            else
              origin_precursor_mz_MS2[i] <- NA_real_
          }

        } else {
          stop("MS", level, " with ", length(parent_pos), " precursors")
        }
      }
    }
    tree_id[idx] <- origin_tree_id
    ## store per-origin results
    precursor_mz_MS2[idx] <- origin_precursor_mz_MS2
  }

  ## Return in original order
  tree_id <- tree_id[order(o)]
  ## reorder new variable to match
  precursor_mz_MS2 <- precursor_mz_MS2[order(o)]
  
  x <- x[order(o)]

  ## add both to Spectra object
  x$MSntreeID <- tree_id
  x$precursorMz.MS2 <- precursor_mz_MS2

  return(x)
}


ftms_msn_tree <- assign_msntree_id(ftms_msn_tree)

spectraVariables(ftms_msn_tree)

```

```{r}
length(unique(ftms_msn_tree$MSntreeID))
```

There are in total `r length(unique(ftms_msn_tree$MSntreeID))` spectral
(fragmentation) trees for in total
`r length(unique(ftms_msn_tree$feature_id))` features. The distribution
of the lengths of the fragmentation trees is

```{r}
ftms_msn_tree$MSntreeID |>
    table() |>
    quantile()

table(table(ftms_msn_tree$MSntreeID))
```

Fragmentation trees have maximally 4 spectra. This is in accordance with
the setup of the ion trap instrument: The two most intense first order
product ions were refragmented (if their intensity was above a
threshold), resulting in maximally two MS3 spectra, and the most intense
second order product ion of the most intense first order product ion was
also refragmented (if it's intensity was above a threshold), resulting
in maximally one MS4 spectrum. So, a fragmentation tree can maximally
consist of one MS2 spectrum + two MS3 spectra + one MS4 spectrum = four
spectra.

```{r}
## we stored the tree ids and the msLevels to a dataframe
df <- spectraData(ftms_msn_tree, c("MSntreeID", "msLevel")) |>
    as.data.frame()

## here we take count the last level of each tree
library(dplyr)
max_level_per_tree <- df %>%
  group_by(MSntreeID) %>%
  summarise(max_level = max(msLevel), .groups = "drop")


print(t <- table(max_level_per_tree$max_level))

```

The number of trees that go up to MS2 level is : `r t[1]`. The number of
trees that go up to MS3 level is : `r t[2]`. The number of trees that go
up to MS4 level is : `r t[3]`.

# Select single best MSn spectral tree per feature

Next we select one most representative tree per feature based on (1) the
intensity of the precursor, (2) the length of the tree, (3) the total
number of peaks in all spectra of the tree, and (4) in rare cases of
several trees with maximal number of peaks: break ties based on
precursor intensity (max)

## (1) Filter out weak precursor trees (lowest precursor intensity)

Keep only trees (`vMSntreeID`v) whose `msLevel == 2` spectrum precursor
intensity is in the top 75%.

```{r}
sd <- as.data.frame(spectraData(ftms_msn_tree))
ms2 <- subset(sd, msLevel == 2)
threshold <- quantile(ms2$precursorIntensity, 0.25, na.rm = TRUE)
strong_MSntrees <- unique(ms2$MSntreeID[ms2$precursorIntensity > threshold])
sd_filtered <- subset(sd, MSntreeID %in% strong_MSntrees)
```

## (2) Among 75% remaining MSn trees, select the longest tree per feature

```{r}
# Get max msLevel per tree
tree_length <- aggregate(msLevel ~ MSntreeID + feature_id, data = sd_filtered, max)

# Get max depth per feature
max_len <- aggregate(msLevel ~ feature_id, data = tree_length, max)

# Keep only trees with the maximum length for their feature
longest_trees <- merge(tree_length, max_len, by = c("feature_id", "msLevel"))

```

## (3) select the tree with highest total number of peaks in all levels

total number of peaks: sum of peaksCount per tree_id in spectraData

```{r}
tree_peaks <- aggregate(peaksCount ~ MSntreeID, data = sd_filtered, sum)

candidate_trees <- merge(longest_trees, tree_peaks, by = "MSntreeID")
max_peaks <- aggregate(peaksCount ~ feature_id, data = candidate_trees, max)
candidate_trees <- merge(candidate_trees, max_peaks, by.x = c("feature_id", "peaksCount"), by.y = c("feature_id", "peaksCount"))

```

## (4) break ties by precursor intensity (highest)

```{r}
# Compute maximal precursor intensity of MS2 spectra per tree
prec_max <- aggregate(precursorIntensity ~ MSntreeID, data = ms2, max)
candidate_trees <- merge(candidate_trees, prec_max, by = "MSntreeID")

# select the tree with highest precursor intensity per feature
candidate_trees <- candidate_trees[order(candidate_trees$feature_id, -candidate_trees$precursorIntensity), ]
best_tree <- candidate_trees[!duplicated(candidate_trees$feature_id), c("feature_id", "MSntreeID")]
```

## (5) filter Spectra object

```{r}
best_MSntreeIDs <- best_tree$MSntreeID
ftms_msn_tree_best <- ftms_msn_tree[ which(spectraData(ftms_msn_tree)$MSntreeID %in% best_MSntreeIDs) ]
```

## (6) (Optional) evaluate based on known spectra

```{r}
idx <- which(round(ftms_msn_tree_best$precursorMz) == 1333)
idx
peaksData(ftms_msn_tree_best)[idx][[1]]


idx <- which(round(ftms_msn_tree_best$precursorMz) == 1171)
idx
peaksData(ftms_msn_tree_best)[idx][[1]]
peaksData(ftms_msn_tree_best)[idx][[2]]
```

This looks like a really good selection !

# Add additional MS3 and MS4 spectra, with different precursorMz, to the selected trees

Here, the purpose is to enrich the selected “best” MSn trees (`ftms_msn_tree_best`) with additional MS3-MS4 spectra from other MS2 and/or MS3 precursors belonging to the same feature, to build a more complete merged fragmentation tree.
Add, from the "strong" spectral trees with MS2 precursor intensity in top 75 %, MS3 spectra with different MS2 precursorMz, and MS4 spectra with different MS3 or MS4 precursorMz, to the selected `ftms_msn_tree_best`.
Create new integer variable: `ftms_msn_tree_best$MergedMSntreeID`.
All spectra belonging to the same merged tree (`ftms_msn_tree_best` + added MS3-MS4) will share the same `MergedMSntreeID`.
Each feature will have 1 merged tree ID.

## MS3-MS4 spectra from other MS2 precursor
Add within each MSntree of `ftms_msn_tree_best`, MS3 spectra, from the 75% "strong" trees `sd_filtered`, belonging to same feature but having different `round(precurosrMz)` :
Select longest tree (meaning MS3 with depending MS4 in the same MSntree_id)
from these, select highest total peakCounts
if ties, select highest precursor Intensity of MS3 spectrum

```{r}
#####CODE ALMOST WORKS!!!!!!!!!!!!!!!!!!!!!
# Extract spectraData from ftms_msn_tree_best
sd_best <- as.data.frame(spectraData(ftms_msn_tree_best))

sd_filtered$precMzRound <- round(sd_filtered$precursorMz)
sd_best$precMzRound <- round(sd_best$precursorMz)

# Extract MS4 spectra from sd_filtered
# ms3_acq: acquisition number of the parent MS3 spectrum
# ms4_acq: acquisition number of the MS4 spectrum
# peaksCount.MS4: peaksCOunt of MS4 spectrum
ms4_spectra <- sd_filtered %>%
  filter(msLevel == 4) %>%
  select(MSntreeID, precScanNum, acquisitionNum, peaksCount) %>%
  dplyr::rename(ms3_acq = precScanNum, ms4_acq = acquisitionNum, peaksCount.MS4 = peaksCount)

#testfunctie3 works! : it selects the candidate MS3 with highest depth -> most total peaks -> highest precursorIntensity
   testfunctie3 <- function(x) {
  x %>%
    {
      if (any(!is.na(.$ms4_acq))) {
        dplyr::filter(., !is.na(ms4_acq))
        
      } else {
        .
      }
    }%>%
    dplyr::mutate(pc = rowSums(cbind(peaksCount, peaksCount.MS4), na.rm = TRUE)) %>%
     # sort descending by pc, then precursorIntensity
    arrange(desc(pc), desc(precursorIntensity)) %>%
    # keep only the top row
    slice_head(n = 1)
   }
  
   # testfunctie4: appends also, to the selected additional MS3, the depending MS4 if any
  testfunctie4 <- function(x) {
  # Filter MS4 if any exist, compute pc, sort, keep top MS3
  top_ms3 <- x %>%
    {
      if (any(!is.na(.$ms4_acq))) {
        dplyr::filter(., !is.na(ms4_acq))
      } else {
        .
      }
    } %>%
    dplyr::mutate(pc = rowSums(cbind(peaksCount, peaksCount.MS4), na.rm = TRUE)) %>%
    dplyr::arrange(desc(pc), desc(precursorIntensity)) %>%
    dplyr::slice_head(n = 1)

  # Add corresponding MS4 spectra, if ms4_acq exists
  if (!all(is.na(top_ms3$ms4_acq))) {
    ms4_row <- sd_filtered %>%
  filter(msLevel == 4, acquisitionNum %in% top_ms3$ms4_acq, feature_id == top_ms3$feature_id)
combined <- bind_rows(top_ms3, ms4_row)
    
    
  } else {
    combined <- top_ms3
  }

  combined
} 
# Initialize empty list to store additional MS3
all_additional_ms3 <- list()

for (f in unique(sd_best$feature_id)) {
  
  # Best MSn tree for this feature
  best_tree <- sd_best %>% filter(feature_id == f)
  best_precMz <- unique(best_tree$precMzRound[best_tree$msLevel == 3])
  
  # Candidate MS3 spectra with different precursor m/z
  ms3_candidates <- sd_filtered %>%
    filter(feature_id == f,
           msLevel == 3,
           !(round(precursorMz) %in% best_precMz))
  
  # Only continue if there are candidates
  if (nrow(ms3_candidates) != 0) {
    print(paste(f,"has", nrow(ms3_candidates), "additional MS3 candidates.", sep=" " ))

    # Join MS3 candidates to their dependent MS4 (if exists)
    ms3_candidates <- ms3_candidates %>%
      left_join(ms4_spectra, by = c("MSntreeID" = "MSntreeID", "acquisitionNum" = "ms3_acq"))
    
  
    # Group by precMzRound and select best MS3

    ms3.p.precmz <- split(ms3_candidates, ms3_candidates$precMzRound)
    additional.ms3 <- lapply(ms3.p.precmz, testfunctie4)
    all_additional_ms3[[f]] <- do.call(rbind, additional.ms3)
  }
}

# Bind all features into a single data frame
all_additional_ms3 <- do.call(rbind, all_additional_ms3)
row.names(all_additional_ms3) <- NULL



# merge the additional MS3 spectra to the selected best trees
common_cols <- intersect(names(sd_best), names(all_additional_ms3))
merged <- bind_rows(
  sd_best[, common_cols],
  all_additional_ms3[, common_cols]
)  

# create new unique merged tree ID 
merged <- merged %>%
  mutate(MergedMSntreeID = as.integer(factor(feature_id)))


mergeKey <- paste(merged$dataOrigin, merged$scanIndex, sep = "_")

original_key <- paste(spectraData(ftms_msn_tree)$dataOrigin,
                      spectraData(ftms_msn_tree)$scanIndex,
                      sep = "_")

merged_spectra <- ftms_msn_tree[original_key %in% mergeKey]
merged <- S4Vectors::DataFrame(merged)
spectraData(merged_spectra) <- merged


```
TODO: solve BUG:
some lines are duplicated:from MSntreeID 1559, MS4 with precMzRound 545, is appended again
> merged[merged$feature_id=="FT4198",]
DataFrame with 8 rows and 46 columns
    msLevel     rtime acquisitionNum scanIndex    dataStorage     dataOrigin centroided  smoothed  polarity
  <integer> <numeric>      <integer> <integer>    <character>    <character>  <logical> <logical> <integer>
1         2   2066.95           3568      3568 D:\\RDynLib... D:\\RDynLib...       TRUE        NA         0
2         3   2064.87           3569      3569 D:\\RDynLib... D:\\RDynLib...       TRUE        NA         0
3         3   2065.19           3570      3570 D:\\RDynLib... D:\\RDynLib...       TRUE        NA         0
4         4   2065.51           3571      3571 D:\\RDynLib... D:\\RDynLib...       TRUE        NA         0
5         4   2064.52           3571      3571 D:\\RDynLib... D:\\RDynLib...       TRUE        NA         0
6         4   2065.51           3571      3571 D:\\RDynLib... D:\\RDynLib...       TRUE        NA         0
7         3   2062.21           3566      3566 D:\\RDynLib... D:\\RDynLib...       TRUE        NA         0
8         3   2067.24           3507      3507 D:\\RDynLib... D:\\RDynLib...       TRUE        NA         0
  precScanNum precursorMz precursorIntensity precursorCharge collisionEnergy isolationWindowLowerMz
    <integer>   <numeric>          <numeric>       <integer>       <numeric>              <numeric>
1        3567     677.134         126397.703               2              35                     NA
2        3568     707.100            883.993               0              35                     NA
3        3568     647.160            807.771               0              35                     NA
4        3569     545.130            321.628               0              35                     NA
5        3569     484.480            116.030               0              35                     NA
6        3569     545.130            321.628               0              35                     NA
7        3565     596.070            553.066               0              35                     NA
8        3506     729.090            535.596               0              35                     NA
  isolationWindowTargetMz isolationWindowUpperMz peaksCount totIonCurrent basePeakMZ basePeakIntensity
                <numeric>              <numeric>  <integer>     <numeric>  <numeric>         <numeric>
1                      NA                     NA         33    5758.27579    707.095         883.99316
2                      NA                     NA          6     475.90352    545.132         321.62772
3                      NA                     NA          4     463.92892    484.929         314.61465
4                      NA                     NA          3      73.31370    445.117          33.29948
5                      NA                     NA          1       8.67285    301.050           8.67285
6                      NA                     NA          3      73.31370    445.117          33.29948
7                      NA                     NA         11     489.00798    485.227          95.01157
8                      NA                     NA          4     124.09554    629.146          62.68373
  ionisationEnergy     lowMZ    highMZ mergedScan mergedResultScanNum mergedResultStartScanNum
         <numeric> <numeric> <numeric>  <integer>           <integer>                <integer>
1                0   295.291  1015.097         NA                  NA                       NA
2                0   299.205   563.556         NA                  NA                       NA
3                0   305.069   484.929         NA                  NA                       NA
4                0   254.943   445.117         NA                  NA                       NA
5                0   301.050   301.050         NA                  NA                       NA
6                0   254.943   445.117         NA                  NA                       NA
7                0   216.861   869.306         NA                  NA                       NA
8                0   427.205   711.211         NA                  NA                       NA
  mergedResultEndScanNum injectionTime filterString    spectrumId ionMobilityDriftTime scanWindowLowerLimit
               <integer>     <numeric>  <character>   <character>            <numeric>            <numeric>
1                     NA             0           NA controller...                   NA              295.291
2                     NA             0           NA controller...                   NA              299.205
3                     NA             0           NA controller...                   NA              305.069
4                     NA             0           NA controller...                   NA              254.943
5                     NA             0           NA controller...                   NA              301.050
6                     NA             0           NA controller...                   NA              254.943
7                     NA             0           NA controller...                   NA              216.861
8                     NA             0           NA controller...                   NA              427.205
  scanWindowUpperLimit electronBeamEnergy chrom_peak_rt chrom_peak_mz chrom_peak_id feature_rtmed
             <numeric>          <numeric>     <numeric>     <numeric>   <character>     <numeric>
1             1015.097                 NA       2070.51       677.134       CP57620       2070.93
2              563.556                 NA            NA            NA            NA            NA
3              484.929                 NA            NA            NA            NA            NA
4              445.117                 NA            NA            NA            NA            NA
5              301.050                 NA            NA            NA            NA            NA
6              445.117                 NA            NA            NA            NA            NA
7              869.306                 NA            NA            NA            NA            NA
8              711.211                 NA            NA            NA            NA            NA
  feature_mzmed  feature_id rtime_adjusted MSntreeID precursorMz.MS2 precMzRound MergedMSntreeID
      <numeric> <character>      <numeric> <numeric>       <numeric>   <numeric>       <integer>
1       677.133      FT4198             NA      1559              NA         677             298
2            NA      FT4198        2067.22      1559         677.134         707             298
3            NA      FT4198        2067.54      1559         677.134         647             298
4            NA      FT4198        2067.86      1559         677.134         545             298
5            NA      FT4198        2068.19      1390         677.132         484             298
6            NA      FT4198        2067.86      1559         677.134         545             298
7            NA      FT4198        2065.87      1389         677.132         596             298
8            NA      FT4198        2067.51      1189         677.130         729             298
## MS4 spectra from other `round(`precursor_mz_MS2)` 

## MS4 spectra from same `round(`precursor_mz_MS2)`, but different `round(precursorMz)`.


after selecting one representative tree per feature we have now: MS2 :
`r ta[1]`, MS3 : `r ta[2]`, and MS4 : `r ta[3]`.

Finally we verify if the number of unique features corresponds to the
number of unique MSntreeID to ensure that we have one tree per feature.

```{r}
length(unique(unlist(ftms_one_tree$feature_id)))

```

```{r}
length(unique(ftms_one_tree$MSntreeID))
```

we have `r length(unique(ftms_one_tree$MSntreeID))` unique feature and
unique MSntreeID, That's mean that we have for each feature one
representative tree.

Now we save the resulting ftms spectra object to a local directory.

```{r}
save(ftms_one_tree, file = file.path(pth, "data", "ftms_one_tree.RData"))
```

```{r}
load("data/ftms_one_tree.RData")
```

```{r}
tst <- spectraData(ftms_one_tree)[spectraData(ftms_one_tree)$feature_id == "FT5291",]
tst$dataOrigin
```

```{r}
peaksData(ftms_one_tree)[spectraData(ftms_one_tree)$feature_id == "FT5291" & spectraData(ftms_one_tree)$msLevel == 2,][[1,]]
```

```{r}
peaksData(ftms_one_tree)[spectraData(ftms_one_tree)$scanIndex == "FT5341" &
                  spectraData(ftms_one_tree)$msLevel == 2,][[1,]]
```

```{r}
spectraData(ftms_one_tree)
```

```{r}
spectraData(ftms_one_tree)[spectraData(ftms_one_tree)$feature_id == "FT0066", ]

```

# Session information

```{r}
sessionInfo()
```
