---
title: "Semi-automated library generation; FTMS LC-MSn data"
format: html
editor_options:
  chunk_output_type: inline
editor:
  markdown:
    wrap: 72
---

# Introduction

This document contains all code for data import, peak detection and refinement, retention time alignment and creation of a feature matrix, as optimized in FTMS_preprocessing.qmd. Code for visualization, and verification has been removed for fast execution.

# Data import

We first load all required R packages for the analysis in this document.

```{r}
#| message: false
library(MsExperiment)
library(Spectra)
library(xcms)
library(readxl)
library(pander)
library(pheatmap)
register(SerialParam())
```

Next we read the LC-MSn data from the mzML data files.

The sample/file information is retrieved from an Excel file
*ftms_data_files.xlsx* that must meet the following minimal
requirements:

– A column named *data_file* containing the filenames of the *.mzML*
files. – A column named *sample_name* with the corresponding sample
names. The sample name for blank control files should be `"blank"`. – A
column named *sample_id*.

All files should be placed locally within the working directory
containing also this file (i.e., *ftms-preprocessing.qmd*). The
directory hierarchy is expected to be the following:

– *\<my.workdir\>/data/mzML*: directory containing all *.mzML* files. –
*\<my.workdir\>/data/sample_data/*: directory containing the Excel file
with the sample and file information.

In the following code, *\<my.workdir\>* is the working directory:

```{r}
pth <- getwd()
pth
```

now we import our dataset

```{r}
sample_data_ftms <- read_xlsx(
    file.path(pth, "data/sample_data/ftms_data_files.xlsx")) |>
    as.data.frame()
MZML_PATH <- file.path(pth,"data","mzML")
ftms <- readMsExperiment(file.path(MZML_PATH, sample_data_ftms$data_file),
                         sampleData = sample_data_ftms)
ftms
```


We next clean the data set filtering to a retention time range of
10-3500 seconds.

```{r}
ftms <- filterSpectra(ftms, filterRt, c(10, 3500))
msLevel(spectra(ftms)) |>
 table()

```

This data contains 4 MS levels but with empty spectra so we removed them
using `filterEmptySpectra()` function.

```{r}
ftms <- filterSpectra(ftms, filterEmptySpectra)
msLevel(spectra(ftms)) |>
  table()
```

# Data preprocessing

## Peak detection on full dataset

We next perform the chromatographic peak detection, using the centWave
algorithm, on the full `MsExperiment`object, containing data of ten mzML
files. With `integrate = 2` we use an alternative algorithm to correctly
identify the boundaries of the identified chromatographic peaks.
Parameter `chunkSize` is used to control the number of files from which
the data should be loaded into memory at a time.

```{r}
ftms <- findChromPeaks(
    ftms, CentWaveParam(peakwidth = c(20, 80), integrate = 2, snthresh = 6,
                        ppm = 50, prefilter = c(3, 500), mzdiff = 0.001),
    chunkSize = 2)
```

> The `xcms::findChromPeaks()` function transforms the
> `MsExperiment` object (originally created using
> `MsExperiment::readMsExperiment()` with mzML files and sample
> metadata) into an `XcmsExperiment` object.
>
> `XcmsExperiment` is an extension of `MsExperiment`, allowing to
> include chromatographic peak information, alignment and grouping, all
> generated by core `xcms` functions.

With this setting we identified `r nrow(chromPeaks(ftms))` peaks in the
full data set. The number of peaks per sample are:

```{r}
pks <- chromPeaks(ftms)
nrow(pks)
table(pks[, "sample"])
```

The distribution of the peak widths in retention time and in m/z
dimensions are:

```{r}
quantile(unname(pks[, "rtmax"] - pks[, "rtmin"]))
quantile(unname(pks[, "mzmax"] - pks[, "mzmin"]))
```

We next perform the **chromatographic peak refinement** to reduce the
number of potential centWave-specific peak detection artifacts. We
choose settings that depend on the observed peak widths above (i.e., about half
of the observed median widths).

```{r}
mnpp <- MergeNeighboringPeaksParam(expandRt = 10, expandMz = 0.001,
                                   minProp = 0.66)
ftms <- refineChromPeaks(ftms, mnpp)
```

Here we add *CleanPeaksParam* to remove chromatographic peaks with a retention
time range larger than the provided maximal acceptable width (maxPeakwidth)
based on (75th percentile of peak widths) * 3.

```{r}
rt_widths <- unname(pks[, "rtmax"] - pks[, "rtmin"])
# Compute 75th percentile and multiply by 3
peakwidth <- quantile(rt_widths, 0.75) * 3
cpp <- CleanPeaksParam(maxPeakwidth = peakwidth)

ftms <- refineChromPeaks(ftms, cpp)
```

We evaluate the number of peaks and the observed peak widths also after
refinement.

```{r}
pks <- chromPeaks(ftms)
nrow(pks)
table(pks[, "sample"])
quantile(unname(pks[,"rtmax"] - pks[,"rtmin"]))
quantile(unname(pks[,"mzmax"] - pks[,"mzmin"]))
```

At last we calculate chromatographic peak quality metrics using the
`chromPeakSummary()` function with the `BetaDistributionParam()`. This resulting
extra peak shape descriptors, including *beta_cor* can then be used to filter
low quality peaks. *beta_cor* is the correlation coefficient between the
extracted ion chromatogram (EIC) peak and an idealized Gaussian peak fitted to
it. We can use this peak shape descriptor to filter out junk peaks. Here, we
first look at the distribution of beta_cor:

```{r}
beta_metrics <- chromPeakSummary(ftms, param = BetaDistributionParam())
colnames(beta_metrics)
summary(beta_metrics[, "beta_cor"])

hist(beta_metrics[, "beta_cor"], breaks = 50,
     main = "Distribution of beta_cor", xlab = "beta_cor")
```

`beta_cor` ranges from -1.0 to 1.0 and has median
`r median(beta_metrics[, "beta_cor"], na.rm = TRUE)`.
Most peaks thus have a `beta-cor`close to 1. Negative beta_cor
values indicate anti-correlation of observed signal with Gaussian-like
model. These peaks seem to have a valley instead of an apex, maybe multiple
shoulders or just baseline fluctuation. Peaks with **negative beta_cor must be
filtered out**. But we could first refine the peaks and evaluate the beta_cor
values again after refinement before filtering.

```{r}
#' silently plotting and saving examples for good, intermediate and bad quality
set.seed(123)
#' Selecting 12 random peaks with a negative beta_cor
cid <- rownames(beta_metrics)[which(beta_metrics[, "beta_cor"] < -0.5)] |>
    sample(size = 12)
eics <- chromPeakChromatograms(ftms, peaks = cid)
#' Saving the EICs to a file
dr <- file.path("images", "chrom_peak_evaluation")
dir.create(dr, recursive = TRUE, showWarnings = FALSE)
png(filename = file.path(dr, "chrom_peaks_negative_beta_cor.png"),
    width = 15, height = 15, units = "cm", res = 600, pointsize = 6)
plot(eics)
dev.off()

#' Selecting 12 random peaks with beta_cor around 0
cid <- rownames(beta_metrics)[which(beta_metrics[, "beta_cor"] > -0.2 &
                                    beta_metrics[, "beta_cor"] < 0.2)] |>
    sample(size = 12)
eics <- chromPeakChromatograms(ftms, peaks = cid)
#' Saving the EICs to a file
png(filename = file.path(dr, "chrom_peaks_0_beta_cor.png"),
    width = 15, height = 15, units = "cm", res = 600, pointsize = 6)
plot(eics)
dev.off()

#' Selecting 12 random peaks with beta_cor around 1
cid <- rownames(beta_metrics)[which(beta_metrics[, "beta_cor"] > 0.9)] |>
    sample(size = 12)
eics <- chromPeakChromatograms(ftms, peaks = cid)
#' Saving the EICs to a file
png(filename = file.path(dr, "chrom_peaks_1_beta_cor.png"),
    width = 15, height = 15, units = "cm", res = 600, pointsize = 6)
plot(eics)
dev.off()

#' Selecting 12 peaks with a NA beta cor
cid <- rownames(beta_metrics)[is.na(beta_metrics[, "beta_cor"])] |>
    sample(size = 12)
eics <- chromPeakChromatograms(ftms, peaks = cid)
#' Saving the EICs to a file
png(filename = file.path(dr, "chrom_peaks_NA_beta_cor.png"),
    width = 15, height = 15, units = "cm", res = 600, pointsize = 6)
plot(eics)
dev.off()

```

Now we want to filter out junk peaks based on `beta_cor < 0`. This will remove
`r length(which(beta_metrics[, "beta_cor"] < 0))` peaks with a negative beta
correlation and `r sum(is.na(beta_metrics[, "beta_cor"]))` peaks with a missing
beta correlation value of the in total
`r nrow(beta_metrics)` chromatographic peaks.

```{r}
#' Keep only chrom peaks with a positive correlation, removing also those
#' with a missing value.
ftms <- filterChromPeaks(ftms, keep = which(beta_metrics[, "beta_cor"] > 0))
```

We evaluate the number of peaks and the observed peak widths.

```{r}
pks <- chromPeaks(ftms)
nrow(pks)
table(pks[, "sample"])
quantile(unname(pks[,"rtmax"] - pks[,"rtmin"]))
quantile(unname(pks[,"mzmax"] - pks[,"mzmin"]))

```

## Retention time alignment

We next perform an initial correspondence analysis that is needed for
the subsequent retention time alignment. We evaluate the settings on the
example m/z range from above.

```{r}
#' Configure settings
pdp <- PeakDensityParam(sampleGroups = sampleData(ftms)$sample_id,
                        bw = 6, minFraction = 0.3, binSize = 0.1,
                        ppm = 10)
# group peaks
ftms <- groupChromPeaks(ftms, param = pdp)

#We next perform retention time alignment:
pgp <- PeakGroupsParam(
    minFraction = 0.9,
    extraPeaks = 100,
    smooth = "loess",
    span = 0.4,
    family = "gaussian"
)
ftms <- adjustRtime(ftms, param = pgp)

```

Below we plot the results from the retention time alignment, i.e., the
difference between the original and adjusted retention times (y-axis)
along the retention time axis (x-axis).

```{r}
plotAdjustedRtime(ftms)
grid()
```

We can see from the plot above that there is one sample that has been
largely adjusted, so it goes up to 15 seconds.


We continue the preprocessing with the final correspondence analysis.
We adapt now the settings, in particular
the bw parameter, that can be much stricter because of the properly
aligned data set. Again, we test the settings on the extracted ion
signal.

```{r}
#' Configure settings
pdp <- PeakDensityParam(sampleGroups = sampleData(ftms)$sample_id,
                        bw = 4, minFraction = 0.3, binSize = 0.1,
                        ppm = 10)

```


now we group chromatographic peaks

```{r}
ftms <- groupChromPeaks(ftms, param = pdp)
```

As a final step we next perform gap-filling to reduce the number of
missing values in the data set.

```{r}
#' The number of missing values before gap filling
sum(is.na(featureValues(ftms)))
head(featureValues(ftms))
```

```{r}
#' Perform gap-filling
ftms <- fillChromPeaks(ftms, param = ChromPeakAreaParam())
```

```{r}
sum(is.na(featureValues(ftms)))
head(featureValues(ftms))
```

```{r}
#' Save the result object

save(ftms, file = file.path(pth, "data", "ftms_fill_filtergauss.RData"))

```


```{r}
tol <- 1e-6
featureDefinitions(ftms)[
  !is.na(featureDefinitions(ftms)$mzmed) &
  featureDefinitions(ftms)$mzmed > 137.023 &
  featureDefinitions(ftms)$mzmed < 137.025, ]

```



```{r}
nrow(featureDefinitions(ftms))
```

```{r}
msLevel(spectra(ftms)) |>
  table()
```


## Feature matrix

Now we will create the feature matrix of this data.

```{r}
corrected_features <- featureDefinitions(ftms)
#here we add a new column called featue in this format "MxTy" with:
#x: mzmed and y: rtmed of each feature

final_feature_matrix <- data.frame(
 `rt(min)` = corrected_features$rtmed / 60,
  mzmed = corrected_features$mzmed,
  feature_name = paste0(
    "M", round(corrected_features$mzmed, 0),
    "T", round(corrected_features$rtmed, 0)
  )
)


```

Now let's print the head of the resulting feature matrix

```{r}
head(final_feature_matrix)
```

```{r}
nrow(final_feature_matrix)
```

# Session information

The R version and package versions used for this analysis were:

```{r}
sessionInfo()
```
