---
title: "Selecting one representative tree per feature"
format:
  html:
    toc: true
    self-contained: true
author: "Ahlam Mentag, Johannes Rainer"
editor: 
  markdown: 
    wrap: 72
---

# Introduction

This Quarto document presents the analysis of FTMS flaxseed data for
selecting one representative tree per feature. Preprocessing and feature
definition was performed in
[ftms_preprocessing.qmd](ftms_preprocessing.qmd).

# Data import

Loading all required libraries and importing the results from the data
preprocessing.

```{r}
#| message: false
library(dplyr)
library(tidyr)
library(xcms)
library(MsExperiment)
library(Spectra)
library(readxl)
library(IRanges)   
library(S4Vectors)  
register(SerialParam())
pth <- getwd()
pth
```

First we load the `XcmsExperiment` ftms object

```{r}

load(file.path(pth, "data", "ftms_sh_minfr.RData"))
class(ftms)
```

# Explore MS2 spectra associated to features

Next, we use `xcms::featureSpectra()` to extract all MS2 spectra for
each feature from `ftms` object.

> The `xcms::featureSpectra()` function returns a `Spectra` object.
>
> When using `msLevel = 2`, the resulting `Spectra` object contains all
> MS2 spectra whose retention time and precursor m/z fall within the
> ranges of any chromatographic peak of a feature.

```{r}
ms2 <- featureSpectra(ftms, msLevel = 2)

msLevel(ms2) |>
  table()
```
```{r}
spectraData(ms2)
```


it returns 4329 MS level 2 spectra. The total number of features for
which an MS2 spectrum was identified is:

```{r}
length(unique(ms2$feature_id))
length(ms2$feature_id)
```

Here we print the initial number of MSLevels.

```{r}
s <- spectra(ftms)
table(msLevel(s))
```

In total, there were 20804 MS2 spectra. This means that 79% of the MS2
spectra ((20804-4329)\*100/20804) could not be assigned to any
chromatographic peak.

The percentage of features with at least one MS2:
```{r}
# Total number of features in ftms
n_total_features <- nrow(featureDefinitions(ftms))

# Features with ≥1 MS2 (those returned by featureSpectra)
n_features_with_ms2 <- length(ms2)

# Percentage
perc_features_with_ms2 <- (n_features_with_ms2 / n_total_features) * 100
perc_features_with_ms2



```
The percentage of features with at least one MS2 is 47%

the percentage of MS2 spectra that are assigned to a feature

```{r}
# Total number of MS2 spectra in the dataset
n_total_ms2 <- sum(msLevel(s) == 2)

# Number of MS2 spectra assigned to features
n_assigned_ms2 <- sum(lengths(ms2))

# Percentage assigned
perc_ms2_assigned <- (n_assigned_ms2 / n_total_ms2) * 100
perc_ms2_assigned

```
the percentage of MS2 spectra that are assigned to a feature is 23%

Now, we will keep only one MS2 spectrum for that if more than one MS2 spectrum 
is linked to the same feature. First, we select the duplicated MS2 spectra by 
identifying those with the same dataOrigin and acquisitionNum but different
feature_id to detect duplicates assigned to multiple features. From these 
duplicated MS2 spectra, we then check first(caseA) if a feature has unique 
MS2 spectra elsewhere, its assignment in the duplicated group is removed and 
the others are flagged as ambiguous. Then (case B), if no feature has unique 
spectra, we keep the feature whose retention time is closest or whose
chromatographic intensity is strongest (top 10%), and mark the rest 
as ambiguous.
Thus, each MS2 spectrum is finally linked to only one feature, while alternative
links are marked as ambiguous.

```{r}
select_unique_MS2 <- function(ms2, ftms) {
  
  sp <- spectraData(ms2)
  
  # Create a unique identifier for each unique groupe of (dataOrigin, scanIndex)
  sp$groups <- paste0(sp$dataOrigin, "_", sp$scanIndex)
  
  # Create indices for duplicated groups 
  group_idx <- split(seq_len(nrow(sp)), sp$groups)
  
  # Initialization
  final_feature <- vector("list", nrow(sp))  # For the chosen features
  amb_f <- vector("list", nrow(sp))          # For ambiguous features
  
  # Table of occurences for each feature in a dataOrigine
  feature_table <- xtabs(~ feature_id + dataOrigin, data = sp)
  
  # Loop on each group
  for (gname in names(group_idx)) {
    # line indices of each group
    gidx <- group_idx[[gname]]  
    # all features that were assigned to the current group
    feature_ids <- unique(unlist(sp$feature_id[gidx]))  
    
    # if a scanIndex was assigned to a unique feature, we don't do anything
    if (length(feature_ids) == 1) {
      final_feature[[gidx[1]]] <- feature_ids
      amb_f[[gidx[1]]] <- character(0)
      next  # go to the next group...
    }
    
    #rt of the first group
    ms2_rt <- rtime(ms2)[gidx[1]]
    
    # Identify sample name based on the path in dataStorage
    fname <- basename(sp$dataStorage[gidx[1]])
    sample_col <- sub("R[0-9]+\\.mzML$", "", fname)
    
    # rtmed and intensity of the feature
    rt_features <- featureDefinitions(ftms)[feature_ids, "rtmed"]
    int_features <- featureDefinitions(ftms)[feature_ids, sample_col]
    
    # Initialisation
    keep_feat <- character(0) #features to keep
    amb_feat <- feature_ids #ambiguous features
    
    # Loop on each feature in the group
    for (i in seq_along(feature_ids)) {
      i_id <- feature_ids[i]
      
      # Case A 
      # Verify if a feature was assigned at least once to a unique MS2 so it's
      #an unambigious feature
      unamb_f <- any(feature_table[i_id, ] == 1)
      if (unamb_f) {
        # If yes we remove it's assignation to this scanIndex and 
        #we marque other features as ambiguous
        next  # go to the next feature...
      }
      
      # Case B
      # Calculate the RT ratio 
      num <- abs(ms2_rt - rt_features[i])
      ratios_rt <- sapply(seq_along(feature_ids)[-i], function(j) {
        denom <- abs(ms2_rt - rt_features[j])
        if (denom == 0) return(Inf)
        num / denom
      })
      
      # Calculate the intensity ratio 
      ratios_int <- sapply(seq_along(feature_ids)[-i], function(j) {
        denom <- int_features[j]
        if (is.na(denom) || denom <= 0) return(NA_real_)
        int_features[i] / denom
      })
      ratios_int <- ratios_int[!is.na(ratios_int)]
      
      # Condition RT
      keep_rt <- which(ratios_rt < 1)  # keep features with ratio < 1
      if (length(keep_rt) > 0) {
        thr <- quantile(ratios_rt[keep_rt], 0.1)  # top 10% of the smallest rt
        if (any(ratios_rt[keep_rt] <= thr)) keep_feat <- c(keep_feat, i_id)
      }
      
      #Intensity condition
      keep_int <- which(ratios_int > 1)  # keep features with int ratio> 1
      if (length(keep_int) > 0) {
        thr <- quantile(ratios_int[keep_int], 0.9)  # top 10% of the highest rt
        if (any(ratios_int[keep_int] >= thr)) keep_feat <- c(keep_feat, i_id)
      }
    }
    
    # Upate the features to keep and the ambiguous ones
    final_feature[[gidx[1]]] <- unique(keep_feat)
    amb_f[[gidx[1]]] <- setdiff(feature_ids, keep_feat)
  }
  
  # Create a new spectra object with the updated features_id and ambiguous column
  ms2_res <- ms2
  spectraData(ms2_res)$feature_id <- CharacterList(final_feature)
  spectraData(ms2_res)$ambiguous <- CharacterList(amb_f)
  
  # Remove duplicates
  keep <- sapply(final_feature, length) > 0
  ms2_res <- ms2_res[keep]
  
  return(ms2_res)
}





ms2 <- select_unique_MS2(ms2, ftms)
spectraData(ms2)
```
at least we have 3660 MS2 spectra

Verification
```{r}
sp <- spectraData(ms2)

# Extract indices where ambiguous is not empty
idx <- which(lengths(sp$ambiguous) > 0)

# Extract those rows
ambiguous_spectra <- sp[idx, ]
ambiguous_spectra

```
there is 133 line with alt_feature_id different to NA from 3907

In the below code we explore, for each
sample (dataOrigin), if it still some MS2 spectra assigned to more than one
feature. 
```{r}
scanIndex.per.ori <- split(ms2$scanIndex, ms2$dataOrigin)
lapply(scanIndex.per.ori, function(x) any(duplicated(x)))
lapply(scanIndex.per.ori, function(x) x[duplicated(x)])
```

No MS2 spectra are indeed assigned to multiple features. We have a
closer look at sample s1r3:

```{r}
ms2.per.ori <- split(ms2, ms2$dataOrigin)
s1r3 <- ms2.per.ori[[3]]                     
spdat  <-spectraData(s1r3)
spdat$scanIndex[duplicated(spdat$scanIndex)]
```

```{r}
spdat[spdat$scanIndex == 381, ]
```

In sample s1r3, spectra are not assigned to a unique feature.

```{r}
spdat[spdat$scanIndex == 1814,]
spdat[spdat$scanIndex == 1540,]
spdat[spdat$scanIndex == 381,]
spdat_si381 <- spdat[spdat$scanIndex == 381,]
spdat_si381$rtime - spdat_si381$feature_rtmed
```



# Extract all MS2-MSn spectra associated to features

We next define a function to extract the full MSn tree for features.
This function takes an `XcmsExperiment` object as input, extracts the
MS2 spectra using `xcms::featureSpectra()` function, and then retrieves
all MS3 (and subsequently MS4 spectra) for all these MS2 spectra using
the *precScanNum* spectra variable. The `ftms_all_levels` function
returns a combined `Spectra` object containing all MS2, MS3, and MS4
spectra, with each spectrum linked to a feature via a propagated
`feature_id` from the originally matched MS2 spectrum.

```{r}
ftms_all_levels <- function(ftms, ms2) {

  #ms2 <- featureSpectra(ftms, msLevel = 2)
  cat("Number of MS2 spectra found:", length(ms2), "\n")

  res <- lapply(unique(ms2$dataOrigin), function(origin) {
    ms2_subset <- filterDataOrigin(ms2, origin)

    ms3_filtered <- filterDataOrigin(filterMsLevel(spectra(ftms), 3), origin)
    ## To support n:m matches
    m <- findMatches(ms2_subset$acquisitionNum, ms3_filtered$precScanNum)
    cat("Number of MS3 matched to MS2:", length(m), "\n")
    ms3_filtered <- ms3_filtered[to(m)]
    ms3_filtered$feature_id <- ms2_subset$feature_id[from(m)]

    ms4_filtered <- filterDataOrigin(filterMsLevel(spectra(ftms), 4), origin)
    m <- findMatches(ms3_filtered$acquisitionNum, ms4_filtered$precScanNum)
    cat("Number of MS4 matched to MS3:", length(m), "\n")
    ms4_filtered <- ms4_filtered[to(m)]
    ms4_filtered$feature_id <- ms3_filtered$feature_id[from(m)]

    # Combine spectra for this origin
    c(ms2_subset, ms3_filtered, ms4_filtered)
  })
    do.call(c, res)
}

ftms_msn_tree <- ftms_all_levels(ftms,ms2)

table(msLevel(ftms_msn_tree))
```




The number of spectra from different MS levels is shown below:

```{r}
table(msLevel(ftms_msn_tree))
```


# Grouping MSn spectra into fragmentation trees

We next need to group all MSn spectra from one fragmentation tree
together.

The function `assign_msntree_id()` will:

-   Take a `Spectra` object `x` containing MS2–MSn spectra\
-   Assign tree IDs to each spectrum in `x`, based on precursor
    relationships within each sample (`dataOrigin`)\
-   Ensure that MS2 spectra always start a new tree, and that MS3/MS4
    spectra inherit the tree ID from their parent
-   Return a vector of integers, the same length as the number of
    spectra in `x`, where each integer represents the fragmentation tree
    ID the spectrum belongs to

The output of the `assign_msntree_id()` function is then added to the
`Spectra` object as a new spectra variable *MSntreeID*.

```{r}

assign_msntree_id <- function(x) {
  ## Order x by MS level: ensures we first assign an ID to the MS2.
  o <- order(msLevel(x))
  x <- x[o]
  scan_index <- scanIndex(x)
  prec_scan_num <- precScanNum(x)
  ms_level <- msLevel(x)
  data_origin <- dataOrigin(x)

  tree_id <- rep(NA_integer_, length(scan_index))
  global_counter <- 0

  for (origin in unique(data_origin)) {
    idx <- which(data_origin == origin)
    origin_scan_index <- scan_index[idx]
    origin_prec_scan_num <- prec_scan_num[idx]
    origin_ms_level <- ms_level[idx]
    origin_tree_id <- rep(NA_integer_, length(idx))

    scan_idx_map <- setNames(seq_along(origin_scan_index), origin_scan_index)

    for (i in seq_along(idx)) {
      level <- origin_ms_level[i]
      if (level == 2) {
        ## If MS2 hasn't been assigned yet
        if (is.na(origin_tree_id[i])) {
          global_counter <- global_counter + 1
          origin_tree_id[i] <- global_counter
        }
      } else {
        ## Assign ID of the related precursor ID
        parent_scan <- origin_prec_scan_num[i]
        parent_pos <- scan_idx_map[as.character(parent_scan)]
        if (length(parent_pos) == 1 && origin_ms_level[parent_pos] == (level - 1))
          origin_tree_id[i] <- origin_tree_id[parent_pos]
        else
          stop("MS", level, " with ", length(parent_pos), " precursors")
      }
    }
    tree_id[idx] <- origin_tree_id
  }
  ## Return in original order
  return(tree_id[order(o)])
}

msntree_ids <- assign_msntree_id(ftms_msn_tree)

ftms_msn_tree$MSntreeID <- msntree_ids
spectraVariables(ftms_msn_tree)

```

The total number of MSn trees is 3660:

```{r}
length(unique(ftms_msn_tree$MSntreeID))
```

for in total `r length(unique(ftms_msn_tree$feature_id))` features. The
distribution of the lengths of the fragmentation trees is

```{r}
ftms_msn_tree$MSntreeID |>
    table() |>
    quantile()
```

There is apparently (at least) one tree with 4 spectra.
Extracting this particular MSn tree.

```{r}
max_index <- ftms_msn_tree$MSntreeID |>
    table() |>
    as.integer() |>
    which.max()

max_tree <- ftms_msn_tree[ftms_msn_tree$MSntreeID == max_index]
max_tree$rtime
max_tree$msLevel
max_tree$precScanNum
max_tree$scanIndex
max_tree$acquisitionNum
max_tree$dataOrigin
max_tree$feature_id
max_tree$MSntreeID
```

The MSntree with maximum number of spectra corresponds to the tree,
within dataOrigin S1R1, depending on MS2 with ScanIndex 2182 which was
assigned to one FeatureID: FT2705 Within dataOrigin S1R1,
scanIndex 2183 (MS3), 2184 (MS3), and 2185 (MS4) are assigned to a single
feature: FT2705.




```{r}
## we stored the tree ids and the msLevels to a dataframe
df <- spectraData(ftms_msn_tree, c("MSntreeID", "msLevel")) |>
    as.data.frame()

## here we take count the last level of each tree
library(dplyr)
max_level_per_tree <- df %>%
  group_by(MSntreeID) %>%
  summarise(max_level = max(msLevel), .groups = "drop")


table(max_level_per_tree$max_level)

```

The number of trees that go up to MS2 level is : 1803 The number of trees
that go up to MS3 level is : 405 The number of trees that go up to MS4
level is : 1452

Next we select one most representative tree per feature based on the
longest one and if we have many, we select one tree with the highest MS2
precursor intensity

```{r}

filtered_trees_spectra <- function(db) {

  # Extract only the needed metadata 
  sps_data <- as.data.frame(
    spectraData(db)[, c("feature_id", "MSntreeID", "msLevel", "precursorIntensity")]
  )

  # Split by feature_id
  feat_block <- split(sps_data, f = sps_data$feature_id)

  # Selection function applied to each feature block
  selection <- function(feature_block) {
    
    tbl <- table(feature_block$MSntreeID)
    max_number <- max(tbl)
    FTmax <- names(tbl)[tbl == max_number]

    # Keep only candidates with the most frequent MSntreeID
    feat_block_sel1 <- feature_block[feature_block$MSntreeID %in% as.numeric(FTmax), ]

    # Keep only MS2
    feat_block_sel1_MS2 <- feat_block_sel1[feat_block_sel1$msLevel == 2, ]
    if (nrow(feat_block_sel1_MS2) == 0) return(NULL)

    # Select the one with the highest precursor intensity
    index <- which.max(feat_block_sel1_MS2$precursorIntensity)
    selected_FT <- feat_block_sel1_MS2$MSntreeID[index]

    # Return all spectra belonging to that chosen MSntreeID
    feature_block[feature_block$MSntreeID == selected_FT, ]
  }

  # Apply selection across all feature blocks
  selected_trees_list <- lapply(feat_block, selection)
  selected_trees_df <- do.call(rbind, selected_trees_list)

 
  keep_idx <- match(selected_trees_df$MSntreeID, db$MSntreeID, nomatch = 0)
  selected_sps <- db[keep_idx]

  return(selected_sps)
}


ftms_one_tree <- filtered_trees_spectra(ftms_msn_tree)

msLevel(ftms_one_tree) |> table()

```

after selecting one representative tree per feature we have now: MS2 :




Finally we verify if the number of unique features corresponds to the
number of unique MSntreeID to ensure that we have one tree per feature.

```{r}
length(unique(ftms_one_tree$feature_id))
```

```{r}
length(unique(ftms_one_tree$MSntreeID))
```

we have *** unique feature and unique MSntreeID, That's mean that we
have for each feature one representative tree.

Now we save the resulting ftms spectra object to a local directory.

```{r}
save(ftms_one_tree, file = file.path(pth, "data", "unique_tree_ftms.RData"))
```

# Session information

```{r}
sessionInfo()
```
