---
title: "Create an assembled MS/MS data from Ftms MSn data"
format:
  html:
    toc: true
    self-contained: true
author: "Ahlam Mentag, Johannes Rainer"
editor:
  markdown:
    wrap: 72
---

# Introduction

In this quarto document we create an assembled MS/MS data from ftms
object that contains one tree per feature created in the
[ftms_filtering.qmd](https://github.com/rebdau/RDynLib_ftms/blob/ahlam/ftms_filtering.qmd)
file.

# Data import

Loading all required libraries and importing the results from the ftms
filtered object.

```{r}
#| message: false
library(dplyr)
library(tidyr)
library(xcms)
library(Spectra)

register(SerialParam())
pth <- getwd()
pth
```

First we load the `Spectra` ftms_one_tree object with one representative
tree per feature.

```{r}
load(file.path(pth, "data", "ftms_one_merged_tree.RData"))
class(ftms_one_merged_tree)
```

```{r}
library(Spectra)

# Extract spectraData
sd <- spectraData(ftms_one_merged_tree)

# Count unique features per MS level
table(sd$msLevel)  # total spectra per MS level

# Count unique feature IDs per MS level
unique_features_ms <- tapply(sd$feature_id, sd$msLevel, function(x) length(unique(x)))
unique_features_ms

```

## Preprocessing: round mz and remove duplicates

Next we round `mz` values to the first integer using `round_perl()`
function and keep only the most intense peak if rounding results in
duplicated mz.

```{r}
round_perl <- function(number) {
  floor(number + 0.5)
}

peaks_round_mz <- function(x, spectrumMsLevel, ...) {
    x[, 1L] <- round_perl(x[, 1L])
    df <- as.data.frame(x)
  colnames(df) <- c("mz", "intensity")
  # Keep only highest intensity for duplicated mz
  df <- df %>%
    group_by(mz) %>%
    slice_max(order_by = intensity, n = 1, with_ties = FALSE) %>%
    arrange(mz)
  # Return as matrix
  as.matrix(df)

}

ftms_one_merged_tree <- addProcessing(ftms_one_merged_tree, peaks_round_mz)


```

## Combine spectra from the same tree

Create a function that takes a list of spectra (matrices), sums
intensities for identical m/z values across spectra, and returns a
merged spectrum.

```{r}
#'Combine spectra across a tree
#'sum intensities if the same mz appears in different spectra
combine_spectra <- function(spectra_list) {
  do.call(rbind, spectra_list) %>%
    as.data.frame() %>%
    group_by(mz) %>%
    summarise(intensity = sum(intensity), .groups = "drop") %>%
    arrange(mz) %>%
    as.matrix()
}
```

## Create assembled spectra object

final function to create a new assembled spectra object

```{r}

create_assembled_spectra <- function(ftms) {
  #'ftms must be a spectra object
  #'Converting the backend to MsBackendMemory 
  #'so we could modify the peaksData
  ftms <- setBackend(ftms, MsBackendMemory())
  meta <- spectraData(ftms)
  pd_list <- peaksData(ftms)
  tree_ids <- unique(meta$MSntreeID)
  
  assembled_list <- vector("list", length(tree_ids))
  k <- 0
  
  for (tree_id in tree_ids) {
    idx <- which(meta$MSntreeID == tree_id)
    ms_levels <- meta$msLevel[idx]
    
    #'Use already preprocessed peaks 
    spectra_list <- pd_list[idx]
    
    # Combine all spectra by summing intensities for matching m/z
    merged_spec <- combine_spectra(spectra_list)
    
    # Use metadata from the first MS2 spectrum
      ms2_row <- idx[ms_levels == 2][1]
      ms2_meta <- meta[ms2_row, , drop = FALSE]
      
    # create new assembled spectrum
      df <- ms2_meta
      #'Mark it as assembled
      df$spectrum.type <- "assembled"
      df$MSntreeID <- tree_id
      #copy the same mz and intensity (already cleaned)
      df$mz <- list(merged_spec[, "mz"])
      df$intensity <- list(merged_spec[, "intensity"])
      # Remove rows where msLevel is NA
      df <- df[!is.na(df$msLevel), , drop = FALSE]
      k <- k + 1
      #'Convert the modified metadata+peaks into a new 
      #'Spectra object and append it to assembled_list
      assembled_list[[k]] <- Spectra(df)
    }
    
  
    
  
  if (length(assembled_list) == 0) {
    warning("No assembled spectra found!")
    return(NULL)
  }
  
  do.call(c, assembled_list)
}


```

Calling the main function

```{r}
ftms_assembled <- create_assembled_spectra(ftms_one_merged_tree)

df <- spectraData(ftms_assembled)
unique(df$msLevel)

```

```{r}
df <- spectraData(ftms_assembled)
df[!is.na(df$msLevel), ]
```

```{r}
peaksData(ftms_one_merged_tree)[spectraData(ftms_one_merged_tree)$MSntreeID == 1788]
```

```{r}
peaksData(ftms_assembled)[spectraData(ftms_assembled)$MSntreeID == 1788]
```

```{r}
save(ftms_assembled, file = file.path(pth, "data", "ftms_assembled.RData"))
```
