---
title: "Flax Qtof negative to Sql format"
format:
  html:
    toc: true
    self-contained: true
author: "Ahlam Mentag, Johannes Rainer"
---

## Shiny Documents

This Quarto document represents the folowing steps to create a Sql database for
FTMS Flax data

```{r}
library(Spectra)
library(xcms)
library(MsExperiment)
library(Spectra)
library(CompoundDb)
library(RSQLite)
library(S4Vectors) 
library(tidyr)
library(dplyr)
register(SerialParam())
setwd("C:/Users/amentag/Desktop/Rcodes/RDynLib_ftms")
pth <- getwd()
pth
```


```{r}
load("data/ftms_assembled.RData")
load("data/ftms_one_merged_tree.RData")
```

```{r}
spectraData(ftms_one_merged_tree)$spectrum.type <- "not_assembled"
```

```{r}
df <- spectraData(ftms_assembled)
df[is.na(df$msLevel), ]
```


```{r}

library(Spectra)

ftms_one_merged_tree <- setBackend(ftms_one_merged_tree,  MsBackendMemory())
ftms_assembled <- setBackend(ftms_assembled,  MsBackendMemory())

# Concatenate
ftms_all <- concatenateSpectra(ftms_one_merged_tree, ftms_assembled)


peaksData(ftms_all)[[1,]]            

spectraData(ftms_all)

```

```{r}
con <- dbConnect(RSQLite::SQLite(), "data/ftms_flax_neg.sqlite")
```

```{r}
#new_exp_id <- max(dbReadTable(con, "experiment")$expid, na.rm = TRUE) + 1
dbExecute(con, "
  CREATE TABLE IF NOT EXISTS experiment (
    expid INTEGER PRIMARY KEY AUTOINCREMENT,  -- auto-incrementing ID
    date TEXT,
    user TEXT,
    machine TEXT,
    column TEXT,
    mstype TEXT,
    buffera TEXT,
    bufferb TEXT,
    gradient_time TEXT,
    source TEXT,
    mode TEXT,
    species TEXT,
    tissue TEXT,
    ce TEXT,
    meta TEXT
  );
")

```

```{r}
dbExecute(con, "
  INSERT INTO experiment 
    (date, user, machine, column, mstype, buffera, bufferb, gradient_time, source, mode, species, tissue, ce, meta)
  VALUES 
    ('2025-10-28', 'Ahlam', 'FTMS', NULL, 'FTMS', NULL, NULL, NULL, NULL, 'neg', NULL, 'seed', NULL, NULL);
")

```

```{r}
dbGetQuery(con, "SELECT * FROM experiment;")

```

Create table ms_compound
```{r}
dbExecute(con, "
  CREATE TABLE IF NOT EXISTS ms_compound (
    compound_id TEXT PRIMARY KEY,
    expid INTEGER,
    nodename TEXT,
    retention_time REAL,
    mass_measured REAL,
    name TEXT,
    formula TEXT,
    exactmass REAL,
    ppm_deviation REAL,
    subsid INTEGER,
    conversion TEXT,
    wavelen REAL,
    smiles TEXT,
    isotope_ratio REAL,
    drift_time REAL,
    composition TEXT,
    inchi TEXT,
    inchikey TEXT,
    FOREIGN KEY (expid) REFERENCES experiment(expid)
  );
")


```

```{r}
sp_data <- spectraData(ftms_all)
head(sp_data[, c("feature_id", "feature_rtmed", "feature_mzmed")])

```

```{r}
library(dplyr)

# Convert Spectra DFrame to a normal data.frame
sp_df <- as.data.frame(sp_data)

# Keep unique features with valid mz and rt
cmp <- sp_df %>%
  select(feature_id, feature_rtmed, feature_mzmed) %>%
  filter(!is.na(feature_id),
         !is.na(feature_rtmed),
         !is.na(feature_mzmed)) %>%
  distinct(feature_id, .keep_all = TRUE)

n <- nrow(cmp)


last_id <- dbGetQuery(con, "
    SELECT MAX(CAST(compound_id AS INTEGER)) AS max_id
    FROM ms_compound;
")$max_id

start_id <- if (is.na(last_id)) 1 else last_id + 1
last_expid <- tail(dbReadTable(con, "experiment")$expid, 1)

# Build final dataframe
ms_compound_df <- data.frame(
  compound_id     = as.character(seq(start_id, length.out = n)),
  expid           = rep(last_expid, n),
  nodename        = cmp$feature_id,
  retention_time  = cmp$feature_rtmed,
  mass_measured   = cmp$feature_mzmed,
  name = NA_character_,
  formula = NA_character_,
  exactmass = NA_real_,
  ppm_deviation = NA_real_,
  subsid = NA_integer_,
  conversion = NA_character_,
  wavelen = NA_real_,
  smiles = NA_character_,
  isotope_ratio = NA_real_,
  drift_time = NA_real_,
  composition = NA_character_,
  inchi = NA_character_,
  inchikey = NA_character_,
  stringsAsFactors = FALSE
)



```

```{r}
#Insert ms_compound table 
dbWriteTable(con, "ms_compound", ms_compound_df, append = TRUE, row.names = FALSE)

```

```{r}
#Verification
dbGetQuery(con, "SELECT * FROM ms_compound LIMIT 10;")
```

Verify if there is duplicated nodename
```{r}
dbGetQuery(
  con,
  "
  SELECT compound_id, retention_time, nodename
  FROM ms_compound
  WHERE expid = 1
    AND nodename IN (
        SELECT nodename
        FROM ms_compound
        WHERE expid = 1
        GROUP BY nodename
        HAVING COUNT(*) > 1
    )
  ORDER BY nodename
  "
)
```
```{r}
ft_compounds <- dbGetQuery(con, "SELECT compound_id, retention_time FROM ms_compound WHERE expid = 1")

# Count NA and non-NA
ft_na_count <- sum(is.na(ft_compounds$retention_time))
ft_not_na_count <- sum(!is.na(ft_compounds$retention_time))

cat("FTMS compounds:\n")
cat("  NA retention times:", ft_na_count, "\n")
cat("  Non-NA retention times:", ft_not_na_count, "\n\n")
```
all the retention times are not null


Create msms_spectrum table
```{r}
# Drop old table
dbExecute(con, "DROP TABLE IF EXISTS msms_spectrum;")

# Force all numeric columns in msms_spectrum to REAL
dbExecute(con, "
CREATE TABLE msms_spectrum (
  spectrum_id INTEGER PRIMARY KEY AUTOINCREMENT,
  compound_id TEXT,
  ms_level INTEGER,
  polarity INTEGER,
  spectrum_type TEXT,
  precursor_mz REAL,
  precursorIntensity REAL,
  precursorCharge INTEGER,
  collision_energy TEXT,
  isolationWindowLowerMz REAL,
  isolationWindowTargetMz REAL,
  isolationWindowUpperMz REAL,
  peaks_count INTEGER,
  totIonCurrent REAL,
  basePeakMZ REAL,
  basePeakIntensity REAL,
  ionisationEnergy REAL,
  lowMZ REAL,
  highMZ REAL,
  mergedScan INTEGER,
  mergedResultScanNum INTEGER,
  mergedResultStartScanNum INTEGER,
  mergedResultEndScanNum INTEGER,
  injectionTime REAL,
  filterString TEXT,
  spectrumId INTEGER,
  ionMobilityDriftTime REAL,
  scanWindowLowerLimit REAL,
  scanWindowUpperLimit REAL,
  electronBeamEnergy REAL,
  originalPrecursorMz REAL,
  precursorPurity REAL,
  chromPeakRT REAL,
  chromPeakMz REAL,
  chromPeakId TEXT,
  rtime REAL,
  scanIndex INTEGER,
  dataStorage TEXT,
  centroided INTEGER,
  smoothed INTEGER,
  instrument TEXT,
  splash TEXT,
  instrument_type TEXT,
  acquisitionNum INTEGER,
  precScanNum INTEGER,
  predicted REAL,
  dataOrigin TEXT,
  original_id TEXT,
  FOREIGN KEY (compound_id) REFERENCES ms_compound(compound_id)
);
")



```


```{r}
sp <- spectraData(ftms_all)

getcol <- function(x) if (x %in% names(sp)) sp[[x]] else NA

msms_df <- data.frame(
  compound_id = NA,
  ms_level = getcol("msLevel"),
  polarity = getcol("polarity"),
  spectrum_type = getcol("spectrum.type"),
  precursor_mz = getcol("precursorMz"),
  precursorIntensity = getcol("precursorIntensity"),
  precursorCharge = getcol("precursorCharge"),  
  collision_energy = getcol("collisionEnergy"),
  isolationWindowLowerMz = getcol("isolationWindowLowerMz"),   
  isolationWindowTargetMz = getcol("isolationWindowTargetMz"),
  isolationWindowUpperMz = getcol("isolationWindowUpperMz"),
  peaks_count = getcol("peaksCount"),
  totIonCurrent = getcol("totIonCurrent"),
  basePeakMZ = getcol("basePeakMZ"),
  basePeakIntensity = getcol("basePeakIntensity"),
  ionisationEnergy = getcol("ionisationEnergy"),
  lowMZ = getcol("lowMZ"),
  highMZ = getcol("highMZ"),
  mergedScan = getcol("mergedScan"),
  mergedResultScanNum = getcol("mergedResultScanNum"),
  mergedResultStartScanNum = getcol("mergedResultStartScanNum"),
  mergedResultEndScanNum = getcol("mergedResultEndScanNum"),
  injectionTime = getcol("injectionTime"),
  filterString = getcol("filterString"),
  spectrumId = getcol("spectrumId"),
  ionMobilityDriftTime = getcol("ionMobilityDriftTime"),
  scanWindowLowerLimit = getcol("scanWindowLowerLimit"),
  scanWindowUpperLimit = getcol("scanWindowUpperLimit"),
  electronBeamEnergy = getcol("electronBeamEnergy"),
  originalPrecursorMz = getcol("originalPrecursorMz"),   
  precursorPurity = getcol("precursorPurity"),
  chromPeakRT = getcol("chromPeakRT"),
  chromPeakMz = getcol("chromPeakMz"),
  chromPeakId = getcol("chromPeakId"),
  rtime = getcol("rtime"),
  scanIndex = getcol("scanIndex"),
  dataStorage = getcol("dataStorage"),
  centroided = getcol("centroided"),
  smoothed = getcol("smoothed"),
  instrument = getcol("instrument"),
  instrument_type = getcol("instrument_type"),
  acquisitionNum = getcol("acquisitionNum"),
  precScanNum = getcol("precScanNum"),
  predicted = getcol("predicted"),
  dataOrigin = getcol("dataOrigin"),
  original_id = getcol("original_id"),
  stringsAsFactors = FALSE
)
msms_df$compound_id <- ms_compound_df$compound_id[match(sp$feature_id, ms_compound_df$nodename)]



```


```{r}
dbWriteTable(con, "msms_spectrum", msms_df, append = TRUE, row.names = FALSE)

```

```{r}
dbGetQuery(con, "SELECT * FROM msms_spectrum LIMIT 10;")
```

```{r}
dbGetQuery(con, "
  SELECT *
  FROM msms_spectrum
  WHERE compound_id IN (
    SELECT compound_id
    FROM msms_spectrum
    GROUP BY compound_id
    HAVING COUNT(*) > 1
  )
  ORDER BY compound_id;
")

```


Create msms_spectrum_peak
```{r}
dbExecute(con, "
  CREATE TABLE IF NOT EXISTS msms_spectrum_peak (
    peak_id INTEGER PRIMARY KEY AUTOINCREMENT,
    spectrum_id INTEGER,
    mz REAL,
    intensity REAL,
    FOREIGN KEY (spectrum_id) REFERENCES msms_spectrum(spectrum_id)
  );
")


```


```{r}
pd <- peaksData(ftms_all)

# Flatten peaks list into a data frame
peaks_df <- do.call(
  rbind,
  lapply(seq_along(pd), function(i) {
    if (!is.null(pd[[i]]) && nrow(pd[[i]]) > 0) {
      data.frame(
        spectrum_id = i,  # corresponds to row order in msms_spectrum
        mz = pd[[i]][, 1],
        intensity = pd[[i]][, 2]
      )
    }
  })
)

head(peaks_df)

```


```{r}
dbWriteTable(con, "msms_spectrum_peak", peaks_df, append = TRUE, row.names = FALSE)

```


```{r}
dbGetQuery(con, "SELECT * FROM msms_spectrum_peak LIMIT 10;")
```

Create Synonym table
```{r}
dbExecute(con, "
  CREATE TABLE IF NOT EXISTS Synonym (
    id INTEGER PRIMARY KEY
  )
")

```

Create featue_matrix table

```{r}
dbExecute(con, "
  CREATE TABLE IF NOT EXISTS feature_matrix (
    nodename TEXT,
    dataOrigin TEXT,
    Value REAL,
    expid INTEGER,
    FOREIGN KEY (expid) REFERENCES experiment(expid)
  );
")

```

Create the content of the feature matrix
```{r}
load("data/ftms.RData")
feature_Values <- featureValues(ftms)
head(feature_Values)
```

```{r}
matrix_to_long <- function(feature_values) {
  feature_values %>%
    as.data.frame() %>%
    tibble::rownames_to_column("feature_id") %>%
    pivot_longer(
      cols = -feature_id,
      names_to = "dataOrigin",
      values_to = "Value"
    )
}

# Here we call the function
long_format <- matrix_to_long(feature_Values)
head(long_format, 10)
nrow(long_format)
```

```{r}
# Get expid from experiment table
exp_id <- dbGetQuery(con, "SELECT expid FROM experiment LIMIT 1;")$expid

# Add expid column
long_format$expid <- exp_id


```

```{r}
names(long_format)[names(long_format) == "feature_id"] <- "nodename"

dbWriteTable(con, "feature_matrix", long_format, append = TRUE, row.names = FALSE)

```

```{r}
dbGetQuery(con, "Select * from feature_matrix LIMIT 10")
```

Create metadata table
```{r}
 md <- make_metadata(source = "Flax FTMS neg", url = NA_character_,
                    source_version = "1.0.0", source_date = "2025-10-28")
dbWriteTable(con, name = "metadata", md, row.names = FALSE)
```

Here we print all the con tables
```{r}

dbListTables(con)
```

Convert the data to compDb object
```{r}
cdb <- CompDb(con)
cdb
```

Convert the compoundDb to a spectra object
```{r}
sp <- Spectra(cdb)
```

```{r}
spectraData(sp, columns = c("msLevel", "compound_id", "nodename", "rtime", "retention_time", "spectrum_type"))
```

```{r}
spectraVariables(sp)
```

