---
title: "Selecting one representative tree per feature"
format:
  html:
    toc: true
    self-contained: true
author: "Ahlam Mentag, Johannes Rainer"
editor:
  markdown:
    wrap: 72
---

# Introduction

This Quarto document presents the analysis of FTMS flaxseed data for
selecting one representative tree per feature. Preprocessing and feature
definition was performed in
[ftms_preprocessing.qmd](ftms_preprocessing.qmd).

# Data import

Loading all required libraries and importing the results from the data
preprocessing.

```{r}
#| message: false
library(dplyr)
library(tidyr)
library(xcms)
library(MsExperiment)
library(Spectra)
library(readxl)
library(IRanges)
library(S4Vectors)
register(SerialParam())
pth <- getwd()
pth
```

First we load the `XcmsExperiment` ftms object with the preprocessing
results.

```{r}
load(file.path(pth, "data", "ftms_fill.RData"))
class(ftms)
```

# Explore MS2 spectra associated to features

Next, we use `xcms::featureSpectra()` to extract all MS2 spectra for
each feature from `ftms` object.

> The `xcms::featureSpectra()` function returns a `Spectra` object.
>
> When using `msLevel = 2`, the resulting `Spectra` object contains all
> MS2 spectra whose retention time and precursor m/z fall within the
> ranges of any chromatographic peak of a feature.

```{r}
ms2 <- featureSpectra(dropFilledChromPeaks(ftms), msLevel = 2, skipFilled = TRUE)

msLevel(ms2) |>
  table()
```

```{r}
spectraData(ms2)
```

it returns `r length(ms2)` MS level 2 spectra. The total number of features for
which an MS2 spectrum was identified is:

```{r}
length(unique(ms2$feature_id))
length(ms2$feature_id)
```

Here we print the initial number of MS levels.

```{r}
s <- spectra(ftms)
table(msLevel(s))
```



In total, there were 20904 MS2 spectra. This means that 91.10% of the MS2
spectra ((20904-1860)\*100/20904) could not be assigned to any
chromatographic peak.

The percentage of features with at least one MS2:

```{r}
# Total number of features in ftms
n_total_features <- nrow(featureDefinitions(ftms))

# Features with ≥1 MS2 (those returned by featureSpectra)
n_features_with_ms2 <- length(unique(ms2$feature_id))

# Percentage
perc_features_with_ms2 <- (n_features_with_ms2 / n_total_features) * 100
perc_features_with_ms2

```

The total number of features with at least one MS2 is `r n_features_with_ms2`,
which corresponds to `r format(perc_features_with_ms2, digits = 3)`%.

A summary on the number of MS2 spectra assigned to a feature is shown
below:

```{r}
table(ms2$feature_id) |>
    quantile()
```

The median number of MS2 spectra per feature (across all data files) is
hence `r median(table(ms2$feature_id))`.

the percentage of MS2 spectra that are assigned to a feature

```{r}
# Total number of MS2 spectra in the dataset
n_total_ms2 <- sum(msLevel(s) == 2)
sp <- spectraData(ms2)
sp$groups <- paste0(sp$dataOrigin, "_", sp$scanIndex)
# Number of MS2 spectra assigned to features
n_assigned_ms2 <- length(unique(sp$groups))

# Percentage assigned
perc_ms2_assigned <- (n_assigned_ms2 / n_total_ms2) * 100
perc_ms2_assigned

```

The total number of assigned MS2 spectra is `r n_assigned_ms2`, which
corresponds to a percentage of `r format(perc_ms2_assigned, digits = 3)`.

Note that an MS2 spectrum could also be assigned to more than one
feature. The total number of MS2 spectra assigned to more than one
feature is:

```{r}
ms2_id <- paste0(ms2$dataOrigin, "_", ms2$acquisitionNum)
ms2_table <- table(ms2_id)
ms2_mult <- names(ms2_table)[ms2_table > 1]

```

Of the `r length(unique(ms2_id))` MS2 spectra, `r length(ms2_mult)` are assigned
to more than one feature. 

# Extract all MS2-MSn spectra associated to features

We next define a function to extract the full MSn tree for features.
This function takes an `XcmsExperiment` object as input, extracts the
MS2 spectra using `xcms::featureSpectra()` function, and then retrieves
all MS3 (and subsequently MS4 spectra) for all these MS2 spectra using
the *precScanNum* spectra variable. The `ftms_all_levels` function
returns a combined `Spectra` object containing all MS2, MS3, and MS4
spectra, with each spectrum linked to a feature via a propagated
`feature_id` from the originally matched MS2 spectrum.

```{r}
ftms_all_levels <- function(ftms, ms2) {

  #ms2 <- featureSpectra(ftms, msLevel = 2)
  cat("Number of MS2 spectra found:", length(ms2), "\n")

  res <- lapply(unique(ms2$dataOrigin), function(origin) {
    ms2_subset <- filterDataOrigin(ms2, origin)

    ms3_filtered <- filterDataOrigin(filterMsLevel(spectra(ftms), 3), origin)
    ## To support n:m matches
    m <- findMatches(ms2_subset$acquisitionNum, ms3_filtered$precScanNum)
    cat("Number of MS3 matched to MS2:", length(m), "\n")
    ms3_filtered <- ms3_filtered[to(m)]
    ms3_filtered$feature_id <- ms2_subset$feature_id[from(m)]

    ms4_filtered <- filterDataOrigin(filterMsLevel(spectra(ftms), 4), origin)
    m <- findMatches(ms3_filtered$acquisitionNum, ms4_filtered$precScanNum)
    cat("Number of MS4 matched to MS3:", length(m), "\n")
    ms4_filtered <- ms4_filtered[to(m)]
    ms4_filtered$feature_id <- ms3_filtered$feature_id[from(m)]

    # Combine spectra for this origin
    c(ms2_subset, ms3_filtered, ms4_filtered)
  })
    do.call(c, res)
}

ftms_msn_tree <- ftms_all_levels(ftms,ms2)

table(msLevel(ftms_msn_tree))
```

The number of spectra from different MS levels is shown below:

```{r}
table(msLevel(ftms_msn_tree))
```

# Grouping MSn spectra into fragmentation trees

We next need to group all MSn spectra from one fragmentation tree
together.

The function `assign_msntree_id()` will:

-   Take a `Spectra` object `x` containing MS2–MSn spectra\
-   Assign tree IDs to each spectrum in `x`, based on precursor
    relationships within each sample (`dataOrigin`)\
-   Ensure that MS2 spectra always start a new tree, and that MS3/MS4
    spectra inherit the tree ID from their parent
-   Return a vector of integers, the same length as the number of
    spectra in `x`, where each integer represents the fragmentation tree
    ID the spectrum belongs to

The output of the `assign_msntree_id()` function is then added to the
`Spectra` object as a new spectra variable *MSntreeID*.

```{r}

assign_msntree_id <- function(x) {
  ## Order x by MS level: ensures we first assign an ID to the MS2.
  o <- order(msLevel(x))
  x <- x[o]
  scan_index <- scanIndex(x)
  prec_scan_num <- precScanNum(x)
  ms_level <- msLevel(x)
  data_origin <- dataOrigin(x)

  tree_id <- rep(NA_integer_, length(scan_index))
  global_counter <- 0

  for (origin in unique(data_origin)) {
    idx <- which(data_origin == origin)
    origin_scan_index <- scan_index[idx]
    origin_prec_scan_num <- prec_scan_num[idx]
    origin_ms_level <- ms_level[idx]
    origin_tree_id <- rep(NA_integer_, length(idx))

    scan_idx_map <- setNames(seq_along(origin_scan_index), origin_scan_index)

    for (i in seq_along(idx)) {
      level <- origin_ms_level[i]
      if (level == 2) {
        ## If MS2 hasn't been assigned yet
        if (is.na(origin_tree_id[i])) {
          global_counter <- global_counter + 1
          origin_tree_id[i] <- global_counter
        }
      } else {
        ## Assign ID of the related precursor ID
        parent_scan <- origin_prec_scan_num[i]
        parent_pos <- scan_idx_map[as.character(parent_scan)]
        if (length(parent_pos) == 1 && origin_ms_level[parent_pos] == (level - 1))
          origin_tree_id[i] <- origin_tree_id[parent_pos]
        else
          stop("MS", level, " with ", length(parent_pos), " precursors")
      }
    }
    tree_id[idx] <- origin_tree_id
  }
  ## Return in original order
  return(tree_id[order(o)])
}

msntree_ids <- assign_msntree_id(ftms_msn_tree)

ftms_msn_tree$MSntreeID <- msntree_ids
spectraVariables(ftms_msn_tree)

```


```{r}
length(unique(ftms_msn_tree$MSntreeID))
```

for in total `r length(unique(ftms_msn_tree$feature_id))` features. The
distribution of the lengths of the fragmentation trees is

```{r}
ftms_msn_tree$MSntreeID |>
    table() |>
    quantile()
```

There is apparently (at least) one tree with 4 spectra. Extracting this
particular MSn tree.

```{r}
max_index <- ftms_msn_tree$MSntreeID |>
    table() |>
    as.integer() |>
    which.max()

max_tree <- ftms_msn_tree[ftms_msn_tree$MSntreeID == max_index]
max_tree$rtime
max_tree$msLevel
max_tree$precScanNum
max_tree$scanIndex
max_tree$acquisitionNum
max_tree$dataOrigin
max_tree$feature_id
max_tree$MSntreeID
```

The MSntree with maximum number of spectra corresponds to the tree,
within dataOrigin S1R1, depending on MS2 with ScanIndex 2182 which was
assigned to one FeatureID: FT1254 Within dataOrigin S1R1, scanIndex 2183
(MS3), 2184 (MS3), and 2185 (MS4) are assigned to a single feature:
FT1254.

```{r}
## we stored the tree ids and the msLevels to a dataframe
df <- spectraData(ftms_msn_tree, c("MSntreeID", "msLevel")) |>
    as.data.frame()

## here we take count the last level of each tree
library(dplyr)
max_level_per_tree <- df %>%
  group_by(MSntreeID) %>%
  summarise(max_level = max(msLevel), .groups = "drop")


table(max_level_per_tree$max_level)

```

The number of trees that go up to MS2 level is : 503 The number of
trees that go up to MS3 level is : 377 The number of trees that go up to
MS4 level is : 980

Next we select one most representative tree per feature based on the
longest one and if we have many, we select one tree with the highest MS2
precursor intensity

```{r}

unique_tree_feature <- function(db) {
  # Store MS2 indices
  ms2_idx <- which(msLevel(db) == 2)

  # Table of linked MS2 features only
  feat_ms2 <- data.frame(
    idx = ms2_idx,
    feature_id = sapply(db$feature_id[ms2_idx], function(x) x[1]),
    MSntreeID = db$MSntreeID[ms2_idx],
    precursorIntensity = db$precursorIntensity[ms2_idx]
  )

  # for each feature choose the longest tree and the one with the highest
  #precursorIntensity()  in case of tie
  selected_tree_ids <- tapply(seq_len(nrow(feat_ms2)), feat_ms2$feature_id, function(i) {
    #the longest tree
    tbl <- table(feat_ms2$MSntreeID[i])
    FTmax <- as.numeric(names(tbl)[tbl == max(tbl)])
    #the highest precursorIntensity()
    sub_idx <- i[feat_ms2$MSntreeID[i] %in% FTmax]
    sub_idx[which.max(feat_ms2$precursorIntensity[sub_idx])]
  })

  tree_ids_keep <- feat_ms2$MSntreeID[unlist(selected_tree_ids)]
  db[which(db$MSntreeID %in% tree_ids_keep)]
}

ftms_one_tree <- unique_tree_feature(ftms_msn_tree)

msLevel(ftms_one_tree) |> table()
```

after selecting one representative tree per feature we have now: MS2 :
520 and MS3 : 550 and MS4 : 265

Finally we verify if the number of unique features corresponds to the
number of unique MSntreeID to ensure that we have one tree per feature.

```{r}
length(unique(unlist(ftms_one_tree$feature_id)))

```

```{r}
length(unique(ftms_one_tree$MSntreeID))
```

we have `r length(unique(ftms_one_tree$MSntreeID))` unique feature and unique MSntreeID, That's mean that we
have for each feature one representative tree.

Now we save the resulting ftms spectra object to a local directory.

```{r}
save(ftms_one_tree, file = file.path(pth, "data", "ftms_unique_tree.RData"))
```

# Session information

```{r}
sessionInfo()
```
