---
title: "Selecting one representative tree per feature"
format:
  html:
    toc: true
    self-contained: true
author: "Ahlam Mentag, Johannes Rainer"
editor: 
  markdown: 
    wrap: 72
---

# Introduction

This Quarto document presents the analysis of FTMS flaxseed data for
selecting one representative tree per feature. Preprocessing and feature
definition was performed in
[ftms_preprocessing.qmd](ftms_preprocessing.qmd).

# Data import

Loading all required libraries and importing the results from the data
preprocessing.

```{r}
#| message: false
library(dplyr)
library(tidyr)
library(xcms)
library(MsExperiment)
library(Spectra)
library(readxl)
register(SerialParam())
pth <- getwd()
pth
```

First we load the `XcmsExperiment` ftms object

```{r}

load(file.path(pth, "data", "ftms_sh_minfr.RData"))
class(ftms)
```

# Explore MS2 spectra associated to features

Next, we use `xcms::featureSpectra()` to extract all MS2 spectra for
each feature from `ftms` object.

> The `xcms::featureSpectra()` function returns a `Spectra` object.
>
> When using `msLevel = 2`, the resulting `Spectra` object contains all
> MS2 spectra whose retention time and precursor m/z fall within the
> ranges of any chromatographic peak of a feature.

```{r}
ms2 <- featureSpectra(ftms, msLevel = 2)

msLevel(ms2) |>
  table()
```
```{r}
spectraData(ms2)
```


it returns 4329 MS level 2 spectra. The total number of features for
which an MS2 spectrum was identified is:

```{r}
length(unique(ms2$feature_id))
length(ms2$feature_id)
```

Here we print the initial number of MSLevels.

```{r}
s <- spectra(ftms)
table(msLevel(s))
```

In total, there were 20804 MS2 spectra. This means that 79% of the MS2
spectra ((20804-4329)\*100/20804) could not be assigned to any
chromatographic peak.

The percentage of features with at least one MS2:
```{r}
# Total number of features in ftms
n_total_features <- nrow(featureDefinitions(ftms))

# Features with ≥1 MS2 (those returned by featureSpectra)
n_features_with_ms2 <- length(ms2)

# Percentage
perc_features_with_ms2 <- (n_features_with_ms2 / n_total_features) * 100
perc_features_with_ms2



```
The percentage of features with at least one MS2 is 47%

the percentage of MS2 spectra that are assigned to a feature

```{r}
# Total number of MS2 spectra in the dataset
n_total_ms2 <- sum(msLevel(s) == 2)

# Number of MS2 spectra assigned to features
n_assigned_ms2 <- sum(lengths(ms2))

# Percentage assigned
perc_ms2_assigned <- (n_assigned_ms2 / n_total_ms2) * 100
perc_ms2_assigned

```
the percentage of MS2 spectra that are assigned to a feature is 23%

Now, we will keep only one MS2 spectrum for that if more than one MS2 spectrum 
is linked to the same feature. First, we select the duplicated MS2 spectra by 
identifying those with the same dataOrigin and acquisitionNum but different
feature_id. From these duplicated MS2 spectra, we then check whether an MS2 
spectrum is associated with the same feature across multiple samples.
In this case, we keep the feature with the maximum occurrences.
Otherwise, if the MS2 spectrum is linked to each feature only once, 
we keep the one with a retention time closest to that of the feature.

```{r}

select_unique_MS2 <- function(ms2, ftms){
  sp <- spectraData(ms2)
  
  # Group by dataOrigin and acquisitionNum
  sp$groupID <- paste0(sp$dataOrigin, "_", sp$acquisitionNum)
  groups <- split(1:nrow(sp), sp$groupID)
  
  keep_indices <- c()
  occurrence <- logical(length=nrow(sp))
  alt_feature <- rep(NA_character_, nrow(sp))
  
  # Calculate the number of occurrences for the feature_id and dataOrigin couple 
  # (note: acquisitionNum is not considered here) and store it in a table 
  feature_table <- xtabs(~ feature_id + dataOrigin, data = sp)
  
  for (idx in groups) {
    # store feature indexes of the duplicated MS2 spectra
    feature_ids <- sp$feature_id[idx]
    
    # Check if feature_ids are different
    if(length(unique(feature_ids)) > 1){
      
      # number of samples (dataOrigin) where the feature occurs exactly once
      cross_occurrences <- sapply(unique(feature_ids), function(f){
        sum(feature_table[f, ] == 1)  
      })
      
      if(any(cross_occurrences > 0)){
        # Choose the feature_id that is unambiguous in other samples
        # (i.e. the feature that occurs exactly once in more samples)
        chosen_feature <- names(cross_occurrences)[which.max(cross_occurrences)]
        chosen <- idx[sp$feature_id[idx] == chosen_feature][1]
      } else {
        # Otherwise choose by closest RT
        rt_diff <- abs(rtime(ms2)[idx] - featureDefinitions(ftms)[feature_ids, "rtmed"])
        chosen <- idx[which.min(rt_diff)]
      }
      
      keep_indices <- c(keep_indices, chosen)
      occurrence[chosen] <- TRUE
      
      # Store the removed feature_ids
      alt_ids <- setdiff(feature_ids, sp$feature_id[chosen])
      alt_feature[chosen] <- paste(alt_ids, collapse=";")
      
    } else {
      keep_indices <- c(keep_indices, idx)
    }
  }
  
  ms2_unique <- ms2[keep_indices]
  spectraData(ms2_unique)$alt_feature_id <- alt_feature[keep_indices]
  
  return(ms2_unique)
}



ms2 <- select_unique_MS2(ms2, ftms)
spectraData(ms2)

```



Verification
```{r}
sp <- spectraData(ms2)

# Extract indices where alt_feature_id is different to NA
idx <- which(!is.na(sp$alt_feature_id))

#Extract those lines
ambiguous_spectra <- sp[idx, ]
ambiguous_spectra
```
there is 400 line with alt_feature_id different to NA from 3907

In the below code we explore, for each
sample (dataOrigin), if it still some MS2 spectra assigned to more than one
feature. 
```{r}
scanIndex.per.ori <- split(ms2$scanIndex, ms2$dataOrigin)
lapply(scanIndex.per.ori, function(x) any(duplicated(x)))
lapply(scanIndex.per.ori, function(x) x[duplicated(x)])
```

No MS2 spectra are indeed assigned to multiple features. We have a
closer look at sample s1r3:

```{r}
ms2.per.ori <- split(ms2, ms2$dataOrigin)
s1r3 <- ms2.per.ori[[3]]                     
spdat  <-spectraData(s1r3)
spdat$scanIndex[duplicated(spdat$scanIndex)]
```

```{r}
spdat[spdat$scanIndex == 381, ]
```

In sample s1r3, spectra are not assigned to a unique feature.

```{r}
spdat[spdat$scanIndex == 1814,]
spdat[spdat$scanIndex == 1540,]
spdat[spdat$scanIndex == 381,]
spdat_si381 <- spdat[spdat$scanIndex == 381,]
spdat_si381$rtime - spdat_si381$feature_rtmed
```



# Extract all MS2-MSn spectra associated to features

We next define a function to extract the full MSn tree for features.
This function takes an `XcmsExperiment` object as input, extracts the
MS2 spectra using `xcms::featureSpectra()` function, and then retrieves
all MS3 (and subsequently MS4 spectra) for all these MS2 spectra using
the *precScanNum* spectra variable. The `ftms_all_levels` function
returns a combined `Spectra` object containing all MS2, MS3, and MS4
spectra, with each spectrum linked to a feature via a propagated
`feature_id` from the originally matched MS2 spectrum.

```{r}
ftms_all_levels <- function(ftms, ms2) {

  #ms2 <- featureSpectra(ftms, msLevel = 2)
  cat("Number of MS2 spectra found:", length(ms2), "\n")

  res <- lapply(unique(ms2$dataOrigin), function(origin) {
    ms2_subset <- filterDataOrigin(ms2, origin)

    ms3_filtered <- filterDataOrigin(filterMsLevel(spectra(ftms), 3), origin)
    ## To support n:m matches
    m <- findMatches(ms2_subset$acquisitionNum, ms3_filtered$precScanNum)
    cat("Number of MS3 matched to MS2:", length(m), "\n")
    ms3_filtered <- ms3_filtered[to(m)]
    ms3_filtered$feature_id <- ms2_subset$feature_id[from(m)]

    ms4_filtered <- filterDataOrigin(filterMsLevel(spectra(ftms), 4), origin)
    m <- findMatches(ms3_filtered$acquisitionNum, ms4_filtered$precScanNum)
    cat("Number of MS4 matched to MS3:", length(m), "\n")
    ms4_filtered <- ms4_filtered[to(m)]
    ms4_filtered$feature_id <- ms3_filtered$feature_id[from(m)]

    # Combine spectra for this origin
    c(ms2_subset, ms3_filtered, ms4_filtered)
  })
    do.call(c, res)
}

ftms_msn_tree <- ftms_all_levels(ftms,ms2)

table(msLevel(ftms_msn_tree))
```

Checking the MSn data for one feature - and evaluate if the relationship
between the MSn spectra is correct.

```{r}
ft <- "FT1287"

a <- ftms_msn_tree[ftms_msn_tree$feature_id == ft]
a

#' split by dataOrigin
al <- split(a, a$dataOrigin)

al[[1]]

spectraData(al[[1]], columns = c("msLevel", "acquisitionNum", "precScanNum"))
#' This looks correct. Testing the next

spectraData(al[[2]], columns = c("msLevel", "acquisitionNum", "precScanNum"))
#' This too.

ft <- "FT6908"
a <- ftms_msn_tree[ftms_msn_tree$feature_id == ft]
a

#' split by dataOrigin
al <- split(a, a$dataOrigin)

spectraData(al[[1]], columns = c("dataOrigin", "msLevel", "acquisitionNum", "precScanNum", "precursorMz")) |>
    as.data.frame()


```


The number of spectra from different MS levels is shown below:

```{r}
table(msLevel(ftms_msn_tree))
```

We check MSn trees propagated from MS2 spectra that were assigned to
multiple chromatographic features.

```{r}
al <- split(ftms_msn_tree, ftms_msn_tree$dataOrigin)
S1R3 <- al[[3]]
spdat  <-spectraData(S1R3)
spdat[spdat$scanIndex %in% 1815:1817,]
spdat[spdat$feature_id %in% c("FT3146","FT3149"),]
spdat[spdat$feature_id %in% c("FT0268","FT0270"),]
```
There is no feature assigned to multiple spectra

# Grouping MSn spectra into fragmentation trees

We next need to group all MSn spectra from one fragmentation tree
together.

The function `assign_msntree_id()` will:

-   Take a `Spectra` object `x` containing MS2–MSn spectra\
-   Assign tree IDs to each spectrum in `x`, based on precursor
    relationships within each sample (`dataOrigin`)\
-   Ensure that MS2 spectra always start a new tree, and that MS3/MS4
    spectra inherit the tree ID from their parent
-   Return a vector of integers, the same length as the number of
    spectra in `x`, where each integer represents the fragmentation tree
    ID the spectrum belongs to

The output of the `assign_msntree_id()` function is then added to the
`Spectra` object as a new spectra variable *MSntreeID*.

```{r}

assign_msntree_id <- function(x) {
  ## Order x by MS level: ensures we first assign an ID to the MS2.
  o <- order(msLevel(x))
  x <- x[o]
  scan_index <- scanIndex(x)
  prec_scan_num <- precScanNum(x)
  ms_level <- msLevel(x)
  data_origin <- dataOrigin(x)

  tree_id <- rep(NA_integer_, length(scan_index))
  global_counter <- 0

  for (origin in unique(data_origin)) {
    idx <- which(data_origin == origin)
    origin_scan_index <- scan_index[idx]
    origin_prec_scan_num <- prec_scan_num[idx]
    origin_ms_level <- ms_level[idx]
    origin_tree_id <- rep(NA_integer_, length(idx))

    scan_idx_map <- setNames(seq_along(origin_scan_index), origin_scan_index)

    for (i in seq_along(idx)) {
      level <- origin_ms_level[i]
      if (level == 2) {
        ## If MS2 hasn't been assigned yet
        if (is.na(origin_tree_id[i])) {
          global_counter <- global_counter + 1
          origin_tree_id[i] <- global_counter
        }
      } else {
        ## Assign ID of the related precursor ID
        parent_scan <- origin_prec_scan_num[i]
        parent_pos <- scan_idx_map[as.character(parent_scan)]
        if (length(parent_pos) == 1 && origin_ms_level[parent_pos] == (level - 1))
          origin_tree_id[i] <- origin_tree_id[parent_pos]
        else
          stop("MS", level, " with ", length(parent_pos), " precursors")
      }
    }
    tree_id[idx] <- origin_tree_id
  }
  ## Return in original order
  return(tree_id[order(o)])
}

msntree_ids <- assign_msntree_id(ftms_msn_tree)

ftms_msn_tree$MSntreeID <- msntree_ids
spectraVariables(ftms_msn_tree)

```

The total number of MSn trees is 3907:

```{r}
length(unique(ftms_msn_tree$MSntreeID))
```

for in total `r length(unique(ftms_msn_tree$feature_id))` features. The
distribution of the lengths of the fragmentation trees is

```{r}
ftms_msn_tree$MSntreeID |>
    table() |>
    quantile()
```

There is apparently (at least) one tree with 4 spectra. Extracting this
particular MSn tree.

We will check MSntreeID of spectra that were assigned to multiple
chromatographic features.

```{r}
max_index <- ftms_msn_tree$MSntreeID |>
    table() |>
    as.integer() |>
    which.max()

max_tree <- ftms_msn_tree[ftms_msn_tree$MSntreeID == max_index]
max_tree$rtime
max_tree$msLevel
max_tree$precScanNum
max_tree$scanIndex
max_tree$acquisitionNum
max_tree$dataOrigin
max_tree$feature_id
max_tree$MSntreeID
```

The MSntree with maximum number of spectra corresponds to the tree,
within dataOrigin S1R1, depending on MS2 with ScanIndex 102 which was
assigned to one FeatureID: FT3099 Within dataOrigin S1R1,
scanIndex 103 (MS3), 104 (MS3), and 105 (MS4) are assigned to a single
feature: FT3099.




```{r}
## we stored the tree ids and the msLevels to a dataframe
df <- spectraData(ftms_msn_tree, c("MSntreeID", "msLevel")) |>
    as.data.frame()

## here we take count the last level of each tree
library(dplyr)
max_level_per_tree <- df %>%
  group_by(MSntreeID) %>%
  summarise(max_level = max(msLevel), .groups = "drop")


table(max_level_per_tree$max_level)

```

The number of trees that go up to MS2 level is : 1943 The number of trees
that go up to MS3 level is : 427 The number of trees that go up to MS4
level is : 1537

Next we select one most representative tree per feature based on the
longest one and if we have many, we select one tree with the highest MS2
precursor intensity

```{r}

filtered_trees_spectra <- function(db) {

  # Extract metadata from the Spectra object
  sps_data <- as.data.frame(spectraData(db))

  # Split by feature_id
  feat_block <- split(sps_data, f = sps_data$feature_id)

  # Selection function applied to each feature block
  selection <- function(feature_block) {
    #Select the most frequent MSnTreeID
    max_number <- max(table(feature_block$MSntreeID))
    FTmax <- names(table(feature_block$MSntreeID))[table(feature_block$MSntreeID) == max_number]
    feat_block_sel1 <- feature_block[feature_block$MSntreeID %in% as.numeric(FTmax), ]
    feat_block_sel1_MS2 <- feat_block_sel1[feat_block_sel1$msLevel == 2, ]

    if (nrow(feat_block_sel1_MS2) == 0) return(NULL)
    #Select the one with the highest MS2 precursor Intensity
    index <- order(feat_block_sel1_MS2$precursorIntensity, decreasing = TRUE)[1]
    selected_FT <- feat_block_sel1_MS2$MSntreeID[index]
    feature_block[feature_block$MSntreeID == selected_FT, ]
  }

  # Apply selection across all feature blocks
  selected_trees_list <- lapply(feat_block, selection)
  selected_trees_df <- do.call(rbind, selected_trees_list)

  # Subset the original spectra object
  selected_sps <- db[db$MSntreeID %in% selected_trees_df$MSntreeID]

  cat("Spectra object filtré avec succès!\n")
  return(selected_sps)
}

ftms_one_tree <- filtered_trees_spectra(ftms_msn_tree)

msLevel(ftms_one_tree) |> table()

```

after selecting one representative tree per feature we have now: MS2 :
712, MS3 : 675, MS4 : 321

Finally we verify if the number of unique features corresponds to the
number of unique MSntreeID to ensure that we have one tree per feature.

```{r}
length(unique(ftms_one_tree$feature_id))
```

```{r}
length(unique(ftms_one_tree$MSntreeID))
```

we have 712 unique feature and unique MSntreeID, That's mean that we
have for each feature one representative tree.

Now we save the resulting ftms spectra object to a local directory.

```{r}
save(ftms_one_tree, file = file.path(pth, "data", "unique_tree_ftms.RData"))
```

# Session information

```{r}
sessionInfo()
```
