---
title: "Selecting one representative tree per feature"
format:
  html:
    toc: true
    self-contained: true
author: "Ahlam Mentag, Johannes Rainer"
editor: 
  markdown: 
    wrap: 72
---

# Introduction

This Quarto document presents the analysis of FTMS flaxseed data for
selecting one representative tree per feature. Preprocessing and feature
definition was performed in
[ftms_preprocessing.qmd](ftms_preprocessing.qmd).

# Data import

Loading all required libraries and importing the results from the data
preprocessing.

```{r}
#| message: false
library(dplyr)
library(tidyr)
library(xcms)
library(MsExperiment)
library(Spectra)
library(readxl)
register(SerialParam())
pth <- getwd()
pth
```

First we load the `XcmsExperiment` ftms object

```{r}

load(file.path(pth, "data", "ftms_sh.RData"))
class(ftms)
```

# Explore MS2 spectra associated to features

Next, we use `xcms::featureSpectra()` to extract all MS2 spectra for
each feature from `ftms` object.

> The `xcms::featureSpectra()` function returns a `Spectra` object.
>
> When using `msLevel = 2`, the resulting `Spectra` object contains all
> MS2 spectra whose retention time and precursor m/z fall within the
> ranges of any chromatographic peak of a feature.

```{r}
ms2 <- featureSpectra(ftms, msLevel = 2)

msLevel(ms2) |>
  table()
```
```{r}
spectraData(ms2)
```


it returns 2755 MS level 2 spectra. The total number of features for
which an MS2 spectrum was identified is:

```{r}
length(unique(ms2$feature_id))
length(ms2$feature_id)
```

Here we print the initial number of MSLevels.

```{r}
s <- spectra(ftms)
table(msLevel(s))
```

In total, there were 20804 MS2 spectra. This means that 86% of the MS2
spectra ((20804-1679)\*100/20804) could not be assigned to any
chromatographic peak.

The percentage of features with at least one MS2:
```{r}
# Total number of features in ftms
n_total_features <- nrow(featureDefinitions(ftms))

# Features with ≥1 MS2 (those returned by featureSpectra)
n_features_with_ms2 <- length(ms2)

# Percentage
perc_features_with_ms2 <- (n_features_with_ms2 / n_total_features) * 100
perc_features_with_ms2



```
The percentage of features with at least one MS2 is 49%

the percentage of MS2 spectra that are assigned to a feature

```{r}
# Total number of MS2 spectra in the dataset
n_total_ms2 <- sum(msLevel(s) == 2)

# Number of MS2 spectra assigned to features
n_assigned_ms2 <- sum(lengths(ms2))

# Percentage assigned
perc_ms2_assigned <- (n_assigned_ms2 / n_total_ms2) * 100
perc_ms2_assigned

```
the percentage of MS2 spectra that are assigned to a feature is 15%

Now we will keep just one MS2 the closest in retention time if there is more
than one MS2 spectra linked to the same feature.
First we select the duplicated MS2 spectra for that we select those with the
same dataOrigin, same acquisitionNum, and different feature_id. and from those 
duplicated MS2 spectra we keep just one with a retention time close to the
retention time of the feature

```{r}

select_unique_MS2 <- function(ms2, ftms){
  sp <- spectraData(ms2)
  
  #Create indices for the duplicated groupes (i.e: dataOrigine + acquisitionNum)
  sp$groupID <- paste0(sp$dataOrigine, "_", sp$acquisitionNum)
  
  #Group Indices by groupID
  groups <- split(1:nrow(sp), sp$groupID)
  
  #List to store indices that we will keep
  keep_indices <- unlist(
    lapply(groups, function(idx){
      #keep just the indices that have different feature_id
      if(length(unique(sp$feature_id[idx])) > 1){
        #Calclate the difference between the RT of each MS2 spectra with
        #the RT of the feature that is assigned to 
        rt_diff <- abs(rtime(ms2)[idx] - featureDefinitions(ftms)[sp$feature_id[idx], "rtmed"])
        
        #keep the index with the smallest RT difference
        idx[which.min(rt_diff)]
      }else{
        idx
      }
      })
  )
  
  ms2_unique <- ms2[keep_indices]
  
  return(ms2_unique)
}

ms2 <- select_unique_MS2(ms2, ftms)
spectraData(ms2)

```


In the below code we explore, for each
sample (dataOrigin), if it still some MS2 spectra assigned to more than one
feature. 
```{r}
scanIndex.per.ori <- split(ms2$scanIndex, ms2$dataOrigin)
lapply(scanIndex.per.ori, function(x) any(duplicated(x)))
lapply(scanIndex.per.ori, function(x) x[duplicated(x)])
```

No MS2 spectra are indeed assigned to multiple features. We have a
closer look at sample s1r3:

```{r}
ms2.per.ori <- split(ms2, ms2$dataOrigin)
s1r3 <- ms2.per.ori[[3]]                     
spdat  <-spectraData(s1r3)
spdat$scanIndex[duplicated(spdat$scanIndex)]
```

```{r}
spdat[spdat$scanIndex == 381, ]
```

In sample s1r3, spectra are not assigned to a unique feature.

```{r}
spdat[spdat$scanIndex == 1814,]
spdat[spdat$scanIndex == 1540,]
spdat[spdat$scanIndex == 381,]
spdat_si381 <- spdat[spdat$scanIndex == 381,]
spdat_si381$rtime - spdat_si381$feature_rtmed
```



# Extract all MS2-MSn spectra associated to features

We next define a function to extract the full MSn tree for features.
This function takes an `XcmsExperiment` object as input, extracts the
MS2 spectra using `xcms::featureSpectra()` function, and then retrieves
all MS3 (and subsequently MS4 spectra) for all these MS2 spectra using
the *precScanNum* spectra variable. The `ftms_all_levels` function
returns a combined `Spectra` object containing all MS2, MS3, and MS4
spectra, with each spectrum linked to a feature via a propagated
`feature_id` from the originally matched MS2 spectrum.

```{r}
ftms_all_levels <- function(ftms, ms2) {

  #ms2 <- featureSpectra(ftms, msLevel = 2)
  cat("Number of MS2 spectra found:", length(ms2), "\n")

  res <- lapply(unique(ms2$dataOrigin), function(origin) {
    ms2_subset <- filterDataOrigin(ms2, origin)

    ms3_filtered <- filterDataOrigin(filterMsLevel(spectra(ftms), 3), origin)
    ## To support n:m matches
    m <- findMatches(ms2_subset$acquisitionNum, ms3_filtered$precScanNum)
    cat("Number of MS3 matched to MS2:", length(m), "\n")
    ms3_filtered <- ms3_filtered[to(m)]
    ms3_filtered$feature_id <- ms2_subset$feature_id[from(m)]

    ms4_filtered <- filterDataOrigin(filterMsLevel(spectra(ftms), 4), origin)
    m <- findMatches(ms3_filtered$acquisitionNum, ms4_filtered$precScanNum)
    cat("Number of MS4 matched to MS3:", length(m), "\n")
    ms4_filtered <- ms4_filtered[to(m)]
    ms4_filtered$feature_id <- ms3_filtered$feature_id[from(m)]

    # Combine spectra for this origin
    c(ms2_subset, ms3_filtered, ms4_filtered)
  })
    do.call(c, res)
}

ftms_msn_tree <- ftms_all_levels(ftms,ms2)

table(msLevel(ftms_msn_tree))
```

Checking the MSn data for one feature - and evaluate if the relationship
between the MSn spectra is correct.

```{r}
ft <- "FT4282"

a <- ftms_msn_tree[ftms_msn_tree$feature_id == ft]
a

#' split by dataOrigin
al <- split(a, a$dataOrigin)

al[[1]]

spectraData(al[[1]], columns = c("msLevel", "acquisitionNum", "precScanNum"))
#' This looks correct. Testing the next

spectraData(al[[2]], columns = c("msLevel", "acquisitionNum", "precScanNum"))
#' This too.

ft <- "FT0556"
a <- ftms_msn_tree[ftms_msn_tree$feature_id == ft]
a

#' split by dataOrigin
al <- split(a, a$dataOrigin)

spectraData(al[[1]], columns = c("dataOrigin", "msLevel", "acquisitionNum", "precScanNum", "precursorMz")) |>
    as.data.frame()


```

The above result shows that `filterPrecursorScan()` extracts the full
MSn tree: it includes spectra of levels MS2 or higher

-   whose acquisition number itself matches the given Precursor scan
    number
-   AND whose precScanNum matches the given Precursor scan number
-   AND higher level spectra that depend on it
-   AND lower level spectra that gave rise to it

The number of spectra from different MS levels is shown below:

```{r}
table(msLevel(ftms_msn_tree))
```

We check MSn trees propagated from MS2 spectra that were assigned to
multiple chromatographic features.

```{r}
al <- split(ftms_msn_tree, ftms_msn_tree$dataOrigin)
S1R3 <- al[[3]]
spdat  <-spectraData(S1R3)
spdat[spdat$scanIndex %in% 1815:1817,]
spdat[spdat$feature_id %in% c("FT3146","FT3149"),]
spdat[spdat$feature_id %in% c("FT0268","FT0270"),]
```
There is no feature assigned to multiple spectra

# Grouping MSn spectra into fragmentation trees

We next need to group all MSn spectra from one fragmentation tree
together.

The function `assign_msntree_id()` will:

-   Take a `Spectra` object `x` containing MS2–MSn spectra\
-   Assign tree IDs to each spectrum in `x`, based on precursor
    relationships within each sample (`dataOrigin`)\
-   Ensure that MS2 spectra always start a new tree, and that MS3/MS4
    spectra inherit the tree ID from their parent
-   Return a vector of integers, the same length as the number of
    spectra in `x`, where each integer represents the fragmentation tree
    ID the spectrum belongs to

The output of the `assign_msntree_id()` function is then added to the
`Spectra` object as a new spectra variable *MSntreeID*.

```{r}

assign_msntree_id_per_feature <- function(x) {
  o <- order(msLevel(x))
  x <- x[o]

  scan_index   <- scanIndex(x)
  prec_scan_num <- precScanNum(x)
  ms_level     <- msLevel(x)
  data_origin  <- dataOrigin(x)
  feature_id   <- x$feature_id

  tree_id <- rep(NA_integer_, length(scan_index))

  # Iterate over features
  for (feat in unique(feature_id)) {
    feat_idx <- which(feature_id == feat)
    feat_scan_index   <- scan_index[feat_idx]
    feat_prec_scan_num <- prec_scan_num[feat_idx]
    feat_ms_level     <- ms_level[feat_idx]
    feat_origin       <- data_origin[feat_idx]

    # Counter per feature
    feat_counter <- 0
    feat_tree_id <- rep(NA_integer_, length(feat_idx))

    # Loop per dataOrigin (optional if you have multiple files)
    for (origin in unique(feat_origin)) {
      idx <- which(feat_origin == origin)

      origin_scan_index   <- feat_scan_index[idx]
      origin_prec_scan_num <- feat_prec_scan_num[idx]
      origin_ms_level     <- feat_ms_level[idx]

      origin_tree_id <- rep(NA_integer_, length(idx))
      scan_idx_map <- setNames(seq_along(origin_scan_index), origin_scan_index)
      ms2_tree_map <- list()

      for (i in seq_along(idx)) {
        level <- origin_ms_level[i]

        if (level == 2) {
          this_scan <- origin_scan_index[i]
          if (!is.null(ms2_tree_map[[as.character(this_scan)]])) {
            origin_tree_id[i] <- ms2_tree_map[[as.character(this_scan)]]
          } else {
            feat_counter <- feat_counter + 1
            origin_tree_id[i] <- feat_counter
            ms2_tree_map[[as.character(this_scan)]] <- feat_counter
          }
        } else {
          parent_scan <- origin_prec_scan_num[i]
          parent_pos  <- scan_idx_map[as.character(parent_scan)]
          if (length(parent_pos) == 1 && origin_ms_level[parent_pos] == (level - 1)) {
            origin_tree_id[i] <- origin_tree_id[parent_pos]
          } else {
            stop("MS", level, " with ", length(parent_pos), " precursors")
          }
        }
      }

      feat_tree_id[idx] <- origin_tree_id
    }

    # Make the IDs globally unique by prefixing with feature_id
    tree_id[feat_idx] <- paste0(feat, "_", feat_tree_id)
  }

  return(tree_id[order(o)])
}


msntree_ids <- assign_msntree_id(ftms_msn_tree)

ftms_msn_tree$MSntreeID <- msntree_ids
spectraVariables(ftms_msn_tree)

```

The total number of MSn trees is 2157:

```{r}
length(unique(ftms_msn_tree$MSntreeID))
```

for in total `r length(unique(ftms_msn_tree$feature_id))` features. The
distribution of the lengths of the fragmentation trees is

```{r}
ftms_msn_tree$MSntreeID |>
    table() |>
    quantile()
```

There is apparently (at least) one tree with 4 spectra. Extracting this
particular MSn tree.

We will check MSntreeID of spectra that were assigned to multiple
chromatographic features.

```{r}
max_index <- ftms_msn_tree$MSntreeID |>
    table() |>
    as.integer() |>
    which.max()

max_tree <- ftms_msn_tree[ftms_msn_tree$MSntreeID == max_index]
max_tree$rtime
max_tree$msLevel
max_tree$precScanNum
max_tree$scanIndex
max_tree$acquisitionNum
max_tree$dataOrigin
max_tree$feature_id
max_tree$MSntreeID
```

The MSntree with maximum number of spectra corresponds to the tree,
within dataOrigin S1R5, depending on MS2 with ScanIndex 103 which was
assigned to one FeatureID: FT1717 Within dataOrigin S1R5,
scanIndex 104 (MS3), 105 (MS3), and 106 (MS4) are assigned to a single
feature: FT1717.




```{r}
## we stored the tree ids and the msLevels to a dataframe
df <- spectraData(ftms_msn_tree, c("MSntreeID", "msLevel")) |>
    as.data.frame()

## here we take count the last level of each tree
library(dplyr)
max_level_per_tree <- df %>%
  group_by(MSntreeID) %>%
  summarise(max_level = max(msLevel), .groups = "drop")


table(max_level_per_tree$max_level)

```

The number of trees that go up to MS2 level is : 727 The number of trees
that go up to MS3 level is : 297 The number of trees that go up to MS4
level is : 1133

Next we select one most representative tree per feature based on the
longest one and if we have many, we select one tree with the highest MS2
precursor intensity

```{r}

filtered_trees_spectra <- function(db) {

  # Extract metadata from the Spectra object
  sps_data <- as.data.frame(spectraData(db))

  # Split by feature_id
  feat_block <- split(sps_data, f = sps_data$feature_id)

  # Selection function applied to each feature block
  selection <- function(feature_block) {
    #Select the most frequent MSnTreeID
    max_number <- max(table(feature_block$MSntreeID))
    FTmax <- names(table(feature_block$MSntreeID))[table(feature_block$MSntreeID) == max_number]
    feat_block_sel1 <- feature_block[feature_block$MSntreeID %in% as.numeric(FTmax), ]
    feat_block_sel1_MS2 <- feat_block_sel1[feat_block_sel1$msLevel == 2, ]

    if (nrow(feat_block_sel1_MS2) == 0) return(NULL)
    #Select the one with the highest MS2 precursor Intensity
    index <- order(feat_block_sel1_MS2$precursorIntensity, decreasing = TRUE)[1]
    selected_FT <- feat_block_sel1_MS2$MSntreeID[index]
    feature_block[feature_block$MSntreeID == selected_FT, ]
  }

  # Apply selection across all feature blocks
  selected_trees_list <- lapply(feat_block, selection)
  selected_trees_df <- do.call(rbind, selected_trees_list)

  # Subset the original spectra object
  selected_sps <- db[db$MSntreeID %in% selected_trees_df$MSntreeID]

  cat("Spectra object filtré avec succès!\n")
  return(selected_sps)
}

ftms_one_tree <- filtered_trees_spectra(ftms_msn_tree)

msLevel(ftms_one_tree) |> table()

```

after selecting one representative tree per feature we have now: MS2 :
519, MS3 : 573, MS4 : 285

Finally we verify if the number of unique features corresponds to the
number of unique MSntreeID to ensure that we have one tree per feature.

```{r}
length(unique(ftms_one_tree$feature_id))
```

```{r}
length(unique(ftms_one_tree$MSntreeID))
```

we have 519 unique feature and unique MSntreeID, That's mean that we
have for each feature one representative tree.

Now we save the resulting ftms spectra object to a local directory.

```{r}
save(ftms_one_tree, file = file.path(pth, "data", "ftms_one_tree_sh.RData"))
```

# Session information

```{r}
sessionInfo()
```
