---
title: "Selecting one representative tree per feature"
format:
  html:
    toc: true
    self-contained: true
author: "Ahlam Mentag, Johannes Rainer"
editor: 
  markdown: 
    wrap: 72
---

# Introduction

This Quarto document presents the analysis of FTMS flaxseed data for
selecting one representative tree per feature. Preprocessing and feature
definition was performed in
[ftms_preprocessing.qmd](ftms_preprocessing.qmd).

# Data import

Loading all required libraries and importing the results from the data
preprocessing.

```{r}
#| message: false
library(dplyr)
library(tidyr)
library(xcms)
library(MsExperiment)
library(Spectra)
library(readxl)
register(SerialParam())
pth <- getwd()
pth
```

First we load the `XcmsExperiment` ftms object

```{r}

load(file.path(pth, "data", "ftms.RData"))
class(ftms)
```
# Explore MS2 spectra associated to features

Next, we use `xcms::featureSpectra()` to extract all MS2 spectra for
each feature from `ftms` object.

> The `xcms::featureSpectra()` function returns a `Spectra` object.
>
> When using `msLevel = 2`, the resulting `Spectra` object contains all
> MS2 spectra whose retention time and precursor m/z fall within the
> ranges of any chromatographic peak of a feature.

```{r}
ms2 <- featureSpectra(ftms, msLevel = 2)

msLevel(ms2) |>
  table()
```

it returns 1679 MS level 2 spectra. The total number of features for
which an MS2 spectrum was identified is:

```{r}
length(unique(ms2$feature_id))
```

```{r}
scanIndex.per.ori <- split(ms2$scanIndex, ms2$dataOrigin)
lapply(scanIndex.per.ori, function(x) any(duplicated(x)))
lapply(scanIndex.per.ori, function(x) x[duplicated(x)])
ms2.per.ori <- split(ms2, ms2$dataOrigin)
s1r3 <- ms2.per.ori[[3]]                     
spdat  <-spectraData(s1r3)
spdat[spdat$scanIndex %in% 1815:1817,]
names(spdat)
spdat$scanIndex[duplicated(spdat$scanIndex)]
spdat[spdat$scanIndex == 1814,]#ScanIndex 1814 is assigned to two FeatureIDs: FT3146 and FT3149
spdat[spdat$scanIndex == 1540,]#ScanIndex 1540 is assigned to two FeatureIDs: FT0870 and FT0871
spdat[spdat$scanIndex == 381,]#ScanIndex 381 is assigned to two FeatureIDs: FT0268 and FT0270 and their retention time is 30 seconds different
spdat_si381 <- spdat[spdat$scanIndex == 381,]
spdat_si381$rtime - spdat_si381$feature_rtmed#rt of spectrum is closer to rtmed of feature FT0268
#but the precursor could be a mixture of both features


```


```{r}
fdefs <- featureDefinitions(ftms)
feature_ids <- rownames(fdefs) 
#FT3146
fid <- "FT0268"
finfo <- fdefs[fid, ]
eic1 <- chromatogram(
  ftms,
  mz = c(finfo$mzmin, finfo$mzmax),
  rt = c(150, 250)
)

#FT3149
fid <- "FT0270"
finfo <- fdefs[fid, ]
eic2 <- chromatogram(
  ftms,
  mz = c(finfo$mzmin, finfo$mzmax),
  rt = c(150, 250)
)

par(mfrow = c(2, 1))
plot(eic1)
plot(eic2)

```
Here we print the initial number of MSLevels.

```{r}
s <- spectra(ftms)
table(msLevel(s))
```

In total, there were 20804 MS2 spectra. This means that 92% of the MS2
spectra ((20804-1679)\*100/20804) could not be assigned to any
chromatographic peak.

# Extract all MS2-MSn spectra associated to features

We next define a function to extract the full MSn tree for features.
This function takes an `XcmsExperiment` object as input, extracts the
MS2 spectra using `xcms::featureSpectra()` function, and then retrieves
all MS3 (and subsequently MS4 spectra) for all these MS2 spectra using
the *precScanNum* spectra variable. The `ftms_all_levels` function
returns a combined `Spectra` object containing all MS2, MS3, and MS4
spectra, with each spectrum linked to a feature via a propagated
`feature_id` from the originally matched MS2 spectrum.

```{r}
ftms_all_levels <- function(ftms) {

  ms2 <- featureSpectra(ftms, msLevel = 2)
  cat("Number of MS2 spectra found:", length(ms2), "\n")

  res <- lapply(unique(ms2$dataOrigin), function(origin) {
    ms2_subset <- filterDataOrigin(ms2, origin)

    ms3_filtered <- filterDataOrigin(filterMsLevel(spectra(ftms), 3), origin)
    ## To support n:m matches
    m <- findMatches(ms2_subset$acquisitionNum, ms3_filtered$precScanNum)
    cat("Number of MS3 matched to MS2:", length(m), "\n")
    ms3_filtered <- ms3_filtered[to(m)]
    ms3_filtered$feature_id <- ms2_subset$feature_id[from(m)]

    ms4_filtered <- filterDataOrigin(filterMsLevel(spectra(ftms), 4), origin)
    m <- findMatches(ms3_filtered$acquisitionNum, ms4_filtered$precScanNum)
    cat("Number of MS4 matched to MS3:", length(m), "\n")
    ms4_filtered <- ms4_filtered[to(m)]
    ms4_filtered$feature_id <- ms3_filtered$feature_id[from(m)]

    # Combine spectra for this origin
    c(ms2_subset, ms3_filtered, ms4_filtered)
  })
    do.call(c, res)
}

ftms_msn_tree <- ftms_all_levels(ftms)

table(msLevel(ftms_msn_tree))
```

Checking the MSn data for one feature - and evaluate if the relationship
between the MSn spectra is correct.

```{r}
ft <- "FT4282"

a <- ftms_msn_tree[ftms_msn_tree$feature_id == ft]
a

#' split by dataOrigin
al <- split(a, a$dataOrigin)

al[[1]]

spectraData(al[[1]], columns = c("msLevel", "acquisitionNum", "precScanNum"))
#' This looks correct. Testing the next

spectraData(al[[2]], columns = c("msLevel", "acquisitionNum", "precScanNum"))
#' This too.

ft <- "FT0649"
a <- ftms_msn_tree[ftms_msn_tree$feature_id == ft]
a

#' split by dataOrigin
al <- split(a, a$dataOrigin)

spectraData(al[[1]], columns = c("dataOrigin", "msLevel", "acquisitionNum", "precScanNum", "precursorMz")) |>
    as.data.frame()

#' Looks correct. To feature FT0649, in sample S1R1, there are 4 MS2 spectra linked, which correctly have all the same precursorMz = 285. The four MS2 spectra have acquisitionNum 3851, 3870, 3905, and 3919.
#' Let's see what filterPrecursorScan does

filterPrecursorScan(al[[1]], 3851)
filterPrecursorScan(al[[1]], 3870)
filterPrecursorScan(al[[1]], 3905)
filterPrecursorScan(al[[1]], 3919)

spectraData(filterPrecursorScan(al[[1]], 3851), columns = c("msLevel", "acquisitionNum", "precScanNum", "precursorMz"))|>as.data.frame()
spectraData(filterPrecursorScan(al[[1]], 3852), columns = c("msLevel", "acquisitionNum", "precScanNum", "precursorMz"))|>as.data.frame()
spectraData(filterPrecursorScan(al[[1]], 3854), columns = c("msLevel", "acquisitionNum", "precScanNum", "precursorMz"))|>as.data.frame()
#' Seems to be OK
```

The above result shows that `filterPrecursorScan()` extracts the full
MSn tree: it includes spectra of levels MS2 or higher

-   whose acquisition number itself matches the given Precursor scan
    number
-   AND whose precScanNum matches the given Precursor scan number
-   AND higher level spectra that depend on it
-   AND lower level spectra that gave rise to it

The number of spectra from different MS levels is shown below:

```{r}
table(msLevel(ftms_msn_tree))
```

We next need to group all MSn spectra from one fragmentation tree
together.

The function `assign_msntree_id()` will:

-   Take a `Spectra` object `x` containing MS2–MSn spectra\
-   Assign tree IDs to each spectrum in `x`, based on precursor
    relationships within each sample (`dataOrigin`)\
-   Ensure that MS2 spectra always start a new tree, and that MS3/MS4
    spectra inherit the tree ID from their parent
-   Return a vector of integers, the same length as the number of
    spectra in `x`, where each integer represents the fragmentation tree
    ID the spectrum belongs to

The output of the `assign_msntree_id()` function is then added to the
`Spectra` object as a new spectra variable *MSntreeID*.

```{r}

assign_msntree_id <- function(x) {
  ## Order x by MS level: ensures we first assign an ID to the MS2.
  o <- order(msLevel(x))
  x <- x[o]
  scan_index <- scanIndex(x)
  prec_scan_num <- precScanNum(x)
  ms_level <- msLevel(x)
  data_origin <- dataOrigin(x)

  tree_id <- rep(NA_integer_, length(scan_index))
  global_counter <- 0

  for (origin in unique(data_origin)) {
    idx <- which(data_origin == origin)
    origin_scan_index <- scan_index[idx]
    origin_prec_scan_num <- prec_scan_num[idx]
    origin_ms_level <- ms_level[idx]
    origin_tree_id <- rep(NA_integer_, length(idx))

    scan_idx_map <- setNames(seq_along(origin_scan_index), origin_scan_index)

    for (i in seq_along(idx)) {
      level <- origin_ms_level[i]
      if (level == 2) {
        ## If MS2 hasn't been assigned yet
        if (is.na(origin_tree_id[i])) {
          global_counter <- global_counter + 1
          origin_tree_id[i] <- global_counter
        }
      } else {
        ## Assign ID of the related precursor ID
        parent_scan <- origin_prec_scan_num[i]
        parent_pos <- scan_idx_map[as.character(parent_scan)]
        if (length(parent_pos) == 1 && origin_ms_level[parent_pos] == (level - 1))
          origin_tree_id[i] <- origin_tree_id[parent_pos]
        else
          stop("MS", level, " with ", length(parent_pos), " precursors")
      }
    }
    tree_id[idx] <- origin_tree_id
  }
  ## Return in original order
  return(tree_id[order(o)])
}

msntree_ids <- assign_msntree_id(ftms_msn_tree)

ftms_msn_tree$MSntreeID <- msntree_ids
spectraVariables(ftms_msn_tree)

```

The total number of MSn trees:

```{r}
length(unique(ftms_msn_tree$MSntreeID))
```

for in total `r length(unique(ftms_msn_tree$feature_id))` features. The
distribution of the lengths of the fragmentation trees is

```{r}
ftms_msn_tree$MSntreeID |>
    table() |>
    quantile()
```

There is apparently (at least) one tree with 7 spectra. Extracting this
particular MSn tree.

```{r}
max_index <- ftms_msn_tree$MSntreeID |>
    table() |>
    as.integer() |>
    which.max()

max_tree <- ftms_msn_tree[ftms_msn_tree$MSntreeID == max_index]
max_tree$rtime
max_tree$msLevel
max_tree$precScanNum
max_tree$scanIndex
max_tree$acquisitionNum
max_tree$dataOrigin
max_tree$feature_id
```

Within dataOrigin S1R3, scanIndex 1815, 1816, and 1817 are assigned to
two different features: FT3146, and FT3149. Below, we have a closer look
at these two features.

```{r}
feat <- c("FT3146","FT3149")
feat_idx <- ms2sp$feature_id %in% feat
ms2sp[feat_idx,]
```

Features FT3146, and FT3149 have almost exactly the same mzmin and
mzmax, but 10 seconds difference in rtmed. We have a look at EICs for
these two features.

```{r}
fdefs <- featureDefinitions(ftms)
feature_ids <- rownames(fdefs) 
#FT3146
fid <- "FT3146"
finfo <- fdefs[fid, ]
eic1 <- chromatogram(
  ftms,
  mz = c(finfo$mzmin, finfo$mzmax),
  rt = c(1000, 1100)
)

#FT3149
fid <- "FT3149"
finfo <- fdefs[fid, ]
eic2 <- chromatogram(
  ftms,
  mz = c(finfo$mzmin, finfo$mzmax),
  rt = c(1000, 1100)
)

par(mfrow = c(2, 1))
plot(eic1)
plot(eic2)

```

The chromatographic peaks of these two features are extremely similar,
there is no real shift in retention time. It looks like we should merge
these these two features.

The code below merges two specific features in an XcmsExperiment object:
[QUESTION for JoRainer: is there an existing xcms function to merge two features?]
```{r}
# Manually merging two features in an XcmsExperiment object
# Extract feature definitions, chromPeaks, and Sample Data
fdefs <- featureDefinitions(ftms)
cp <- chromPeaks(ftms)
sdata <- sampleData(ftms)

# Extract the two feature IDs to merge
fid1 <- "FT3146"
fid2 <- "FT3149"

f1 <- fdefs[fid1, ]
f2 <- fdefs[fid2, ]

# Merge the peak indices
merged_peakidx <- unique(c(unlist(f1$peakidx), unlist(f2$peakidx)))

# Get sample indices for merged peaks
sample_indices <- cp[merged_peakidx, "sample"]

# Determine sample group membership
sample_group_membership <- sdata$sample_id[sample_indices]
sample_counts <- table(factor(sample_group_membership, levels = unique(sample_groups)))

# Construct merged feature row
merged_feature <- DataFrame(
  mzmed = mean(c(f1$mzmed, f2$mzmed)),
  mzmin = min(f1$mzmin, f2$mzmin),
  mzmax = max(f1$mzmax, f2$mzmax),
  rtmed = mean(c(f1$rtmed, f2$rtmed)),
  rtmin = min(f1$rtmin, f2$rtmin),
  rtmax = max(f1$rtmax, f2$rtmax),
  npeaks = as.numeric(length(unique(sample_names))),
  peakidx = I(list(merged_peakidx)),
  ms_level = unique(c(f1$ms_level, f2$ms_level))
)

# Add per-group sample columns
group_cols <- colnames(fdefs)[colnames(fdefs) %in% names(sample_counts)]
for (grp in group_cols) {
  merged_feature[[grp]] <- as.numeric(sample_counts[[grp]])
}
merged_feature <- merged_feature[, colnames(fdefs)]

# Replace original features with merged one
fdefs_new <- fdefs[!(rownames(fdefs) %in% c(fid1, fid2)), ]
fdefs_new <- rbind(fdefs_new, as.data.frame(merged_feature))
rownames(fdefs_new)[nrow(fdefs_new)] <- paste0(fid1, "_", fid2, "_merged")

# Update the XcmsExperiment object
featureDefinitions(ftms) <- fdefs_new


```

TODO: After merging two features: in spectraData of ftms_msn_tree (sampleOrigin S1R3), scanIndex 1815, 1816, 1817 are double. One should be removed.
```{r}
al <- split(ftms_msn_tree, ftms_msn_tree$dataOrigin)
S1R3 <- al[[3]]
spdat  <-spectraData(S1R3)
spdat[spdat$scanIndex %in% 1815:1817,]

```


TODO: create code to check if MSn_trees are assigned to more than one
feature

```{r}
## we stored the tree ids and the msLevels to a dataframe
df <- spectraData(ftms_msn_tree, c("MSntreeID", "msLevel")) |>
    as.data.frame()

## here we take count the last level of each tree
library(dplyr)
max_level_per_tree <- df %>%
  group_by(MSntreeID) %>%
  summarise(max_level = max(msLevel), .groups = "drop")


table(max_level_per_tree$max_level)

```

The number of trees that go up to MS2 level is : 415 The number of trees
that go up to MS3 level is : 338 The number of trees that go up to MS4
level is : 926

Next we select one most representative tree per feature based on the
longest one and if we have many, we select one tree with the highest MS2
precursor intensity

```{r}

filtered_trees_spectra <- function(db) {

  # Extract metadata from the Spectra object
  sps_data <- as.data.frame(spectraData(db))

  # Split by feature_id
  feat_block <- split(sps_data, f = sps_data$feature_id)

  # Selection function applied to each feature block
  selection <- function(feature_block) {
    #Select the most frequent MSnTreeID
    max_number <- max(table(feature_block$MSntreeID))
    FTmax <- names(table(feature_block$MSntreeID))[table(feature_block$MSntreeID) == max_number]
    feat_block_sel1 <- feature_block[feature_block$MSntreeID %in% as.numeric(FTmax), ]
    feat_block_sel1_MS2 <- feat_block_sel1[feat_block_sel1$msLevel == 2, ]

    if (nrow(feat_block_sel1_MS2) == 0) return(NULL)
    #Select the one with the highest MS2 precursor Intensity
    index <- order(feat_block_sel1_MS2$precursorIntensity, decreasing = TRUE)[1]
    selected_FT <- feat_block_sel1_MS2$MSntreeID[index]
    feature_block[feature_block$MSntreeID == selected_FT, ]
  }

  # Apply selection across all feature blocks
  selected_trees_list <- lapply(feat_block, selection)
  selected_trees_df <- do.call(rbind, selected_trees_list)

  # Subset the original spectra object
  selected_sps <- db[db$MSntreeID %in% selected_trees_df$MSntreeID]

  cat("Spectra object filtré avec succès!\n")
  return(selected_sps)
}

ftms_one_tree <- filtered_trees_spectra(ftms_msn_tree)

msLevel(ftms_one_tree) |> table()

```

after selecting one representative tree per feature we have now: MS2 :
437, MS3 : 523, MS4 : 262

Finally we verify if the number of unique features corresponds to the
number of unique MSntreeID to ensure that we have one tree per feature.

```{r}
length(unique(ftms_one_tree$feature_id))
```

```{r}
length(unique(ftms_one_tree$MSntreeID))
```

we have 437 unique feature and unique MSntreeID, That's mean that we
have for each feature one representative tree.

Now we save the resulting ftms spectra object to a local directory.

```{r}
save(ftms_one_tree, file = file.path(pth, "data", "ftms_one_tree.RData"))
```

# Session information

```{r}
sessionInfo()
```
