---
title: "Selecting one representative tree per feature"
format:
  html:
    toc: true
    self-contained: true
author: "Ahlam Mentag, Johannes Rainer"
editor:
  markdown:
    wrap: 72
---

# Introduction

This Quarto document presents the analysis of FTMS flaxseed data for
selecting one representative tree per feature. Preprocessing and feature
definition was performed in
[ftms_preprocessing.qmd](ftms_preprocessing.qmd).

# Data import

Loading all required libraries and importing the results from the data
preprocessing.

```{r}
#| message: false
library(dplyr)
library(tidyr)
library(xcms)
library(MsExperiment)
library(Spectra)
library(readxl)
library(IRanges)
library(S4Vectors)
register(SerialParam())
pth <- getwd()
pth
```

First we load the `XcmsExperiment` ftms object with the preprocessing results.

```{r}
load(file.path(pth, "data", "ftms_sh_minfr.RData"))
class(ftms)
```

# Explore MS2 spectra associated to features

Next, we use `xcms::featureSpectra()` to extract all MS2 spectra for
each feature from `ftms` object.

> The `xcms::featureSpectra()` function returns a `Spectra` object.
>
> When using `msLevel = 2`, the resulting `Spectra` object contains all
> MS2 spectra whose retention time and precursor m/z fall within the
> ranges of any chromatographic peak of a feature.

```{r}
ms2 <- featureSpectra(ftms, msLevel = 2)

msLevel(ms2) |>
  table()
```

```{r}
spectraData(ms2)
```


it returns 4409 MS level 2 spectra. The total number of features for
which an MS2 spectrum was identified is:

```{r}
length(unique(ms2$feature_id))
length(ms2$feature_id)
```

Here we print the initial number of MS levels.

```{r}
s <- spectra(ftms)
table(msLevel(s))
```

In total, there were 20804 MS2 spectra. This means that 79% of the MS2
spectra ((20804-4409)\*100/20804) could not be assigned to any
chromatographic peak.

The percentage of features with at least one MS2:

```{r}
# Total number of features in ftms
n_total_features <- nrow(featureDefinitions(ftms))

# Features with ≥1 MS2 (those returned by featureSpectra)
n_features_with_ms2 <- length(unique(ms2$feature_id))

# Percentage
perc_features_with_ms2 <- (n_features_with_ms2 / n_total_features) * 100
perc_features_with_ms2

```

The percentage of features with at least one MS2 is 8%

A summary on the number of MS2 spectra assigned to a feature is shown below:

```{r}
table(ms2$feature_id) |>
    quantile()
```

The median number of MS2 spectra per feature (across all data files) is hence 3.

the percentage of MS2 spectra that are assigned to a feature

```{r}
# Total number of MS2 spectra in the dataset
n_total_ms2 <- sum(msLevel(s) == 2)
sp <- spectraData(ms2)
sp$groups <- paste0(sp$dataOrigin, "_", sp$scanIndex)
# Number of MS2 spectra assigned to features
n_assigned_ms2 <- length(unique(sp$groups))



# Percentage assigned
perc_ms2_assigned <- (n_assigned_ms2 / n_total_ms2) * 100
perc_ms2_assigned

```

the percentage of MS2 spectra that are assigned to a feature is 19%

Note that an MS2 spectrum could also be assigned to more than one feature. The
total number of MS2 spectra assigned to more than one feature is:

```{r}
ms2_id <- paste0(ms2$dataOrigin, ms2$acquisitionNum)
ms2_mult <- names(table(ms2_id)[table(ms2_id) > 1])
length(ms2_mult)
table(table(ms2_id))
ms2_three <- names(table(ms2_id)[table(ms2_id) == 3])

```

Most MS2 spectra that are assigned to more than one feature, are assigned to two features (420 MS2 spectra). MS2 spectra are maximally assigned to 3 features (38 MS2 spectra)

Manually evaluate some of these cases:

```{r}
a <- ms2[ms2_id == ms2_mult[3L]]
print(spa <- spectraData(a))

a_chr <- featureChromatograms(ftms, features = a$feature_id)
par(mfrow = c(1, 2))
plot(a_chr[1])
abline(v = a$rtime, col = "red")
plot(a_chr[2])
abline(v = a$rtime, col = "red")

ms2_rt <- a$rtime
abs(ms2_rt - spa$feature_rtmed)
abs(ms2_rt[1] - spa$feature_rtmed[1])/abs(ms2_rt[2] - spa$feature_rtmed[2])
```

The two features to which MS2 spectrum "a" was assigned have exactly the same feature_mzmed = 285.181. The features are also very close in the retention time dimension, eventually
even not well separated. The MS2 spectrum (retention time highlighted in red) is
associated to a chromatographic peak in one sample that was eventually wrongly
assigned to the earlier feature.
In this case, the MS2 assignment to feature FT2029 should be removed. This can be done based on the relative retention time differences with respect to the two features.  

For most other examples, the feature seem to represent noise or at least no
clear chromatographic signal (see for example below).

```{r}
a <- ms2[ms2_id == ms2_mult[200L]]
print(spa <- spectraData(a))

a_chr <- featureChromatograms(ftms, features = a$feature_id)
par(mfrow = c(1, 2))
plot(a_chr[1])
abline(v = a$rtime, col = "red")
plot(a_chr[2])
abline(v = a$rtime, col = "red")
ms2_rt <- a$rtime
abs(ms2_rt - spa$feature_rtmed)
abs(ms2_rt[1] - spa$feature_rtmed[1])/abs(ms2_rt[2] - spa$feature_rtmed[2])
```

Here we have a look at MS2 that were assigned to three different features:
```{r}
a <- ms2[ms2_id == ms2_three[1L]]
print(spa <- spectraData(a))

a_chr <- featureChromatograms(ftms, features = a$feature_id)
par(mfrow = c(1, 3))
plot(a_chr[1])
abline(v = a$rtime, col = "red")
plot(a_chr[2])
abline(v = a$rtime, col = "red")
plot(a_chr[3])
abline(v = a$rtime, col = "red")
ms2_rt <- a$rtime

abs(ms2_rt - spa$feature_rtmed)
abs(ms2_rt[1] - spa$feature_rtmed[1])/abs(ms2_rt[2] - spa$feature_rtmed[2])
abs(ms2_rt[1] - spa$feature_rtmed[1])/abs(ms2_rt[3] - spa$feature_rtmed[3])
```
Also here, all three features have exactly the same mzmed. The first feature (FT0170) has the retention time closest to the MS2 RT, and has also the best peak shape. the two other features don't have a proper peak shape and look mainly like noise.

```{r}
a <- ms2[ms2_id == ms2_three[2L]]
print(spa <- spectraData(a))

a_chr <- featureChromatograms(ftms, features = a$feature_id)
par(mfrow = c(1, 3))
plot(a_chr[1])
abline(v = a$rtime, col = "red")
plot(a_chr[2])
abline(v = a$rtime, col = "red")
plot(a_chr[3])
abline(v = a$rtime, col = "red")
ms2_rt <- a$rtime

abs(ms2_rt - spa$feature_rtmed)
abs(ms2_rt[1] - spa$feature_rtmed[1])/abs(ms2_rt[2] - spa$feature_rtmed[2])
abs(ms2_rt[1] - spa$feature_rtmed[1])/abs(ms2_rt[3] - spa$feature_rtmed[3])
```
Also here, all three features have exactly the same mzmed. The first feature (FT2021) has the retention time closest to the MS2 RT, and clearly has also the best peak shape. The two other features don't have a proper peak shape, they look like multiple unseparated peaks.

**Question**: (maybe for the next meeting with Rebecca) should we maybe consider
keeping only *high quality* signals?  trying to remove these noisy
features/chromatographic peaks? 
==> we could try gradually increasing the sn ratio again

**TODO**: We should also have a look specifically at the MS2 spectra that have been assigned to features with different mzmed.

## Resolving multiple assignments of MS2 spectra to features

We next evaluate the cases above to reduce/remove the multiple assignments and
keep only a single MS2-feature association.
First, we select the duplicated MS2 spectra by
identifying those with the same *dataOrigin* and *acquisitionNum* but different
*feature_id* to detect duplicates assigned to multiple features. From these
duplicated MS2 spectra, we then check first (*case A*) if a feature has unique
MS2 spectra elsewhere. If that is the case any ambiguous assignment is removed
and only the unambiguous mappings are kept for the feature. Subsequently, the
assignments are checked again and if there are still duplicated assignments the
distance of the MS2 to the respective features is evaluated (*case B*): we keep
the top 10% MS2-feature assignments. All other assignments are dropped for the
respective feature.

```{r}
select_unique_MS2 <- function(ms2, ftms) {

  sp <- spectraData(ms2)

  # Create a unique identifier for each unique groupe of (dataOrigin, scanIndex)
  sp$groups <- paste0(sp$dataOrigin, "_", sp$scanIndex)

  # Create indices for duplicated groups
  group_idx <- split(seq_len(nrow(sp)), sp$groups)

  # define MS2 as ambiguous if they are duplicated in sp (i.e., are assigned
  # to >= 2 features. We use the unique ID for the MS2 spectra and their
  # indices in sp for that.
  sp$is_ambiguous <- FALSE
  sp$is_ambiguous[sp$groups %in% names(group_idx)[lengths(group_idx) > 1]] <- TRUE

  # Initialization
  final_feature <- vector("list", nrow(sp))  # For the chosen features
  amb_f <- vector("list", nrow(sp))          # For ambiguous features

  # Table of occurences for each feature in a dataOrigine
  # jo: I think it's not good to use this table. it can happen (and it's not
  # a problem) if we have 2 MS2 spectra for the same feature in the same
  # sample (dataOrigin). The problem is if the same MS2 is assigned to
  # two different features.
  # feature_table <- xtabs(~ feature_id + dataOrigin, data = sp)

  # Loop on each group
  for (gname in names(group_idx)) {
    # line indices of each group
    gidx <- group_idx[[gname]]
    # all features that were assigned to the current group
    feature_ids <- unique(unlist(sp$feature_id[gidx]))

    # if a scanIndex was assigned to a unique feature, we don't do anything
    if (length(feature_ids) == 1) {
      final_feature[[gidx[1]]] <- feature_ids
      amb_f[[gidx[1]]] <- character(0)
      next  # go to the next group...
    }

    #rt of the first group
    ms2_rt <- rtime(ms2)[gidx[1]]


    # rtmed and intensity of the feature
    rt_features <- featureDefinitions(ftms)[feature_ids, "rtmed"]

    # Initialisation
    keep_feat <- character(0) #features to keep
    amb_feat <- feature_ids #ambiguous features

    # Loop on each feature in the group
    for (i in seq_along(feature_ids)) {
      i_id <- feature_ids[i]

      # Case A
      # Verify if a feature was assigned at least once to a unique MS2 so there
      # is an(other) unambiguous MS2-feature assignment. We thus skip/drop
      # the current assignment and move to the next.
      # unamb_f <- any(feature_table[i_id, ] == 1)
      if (any(!sp$is_ambiguous[sp$feature_id == i_id])) {
        # If yes we remove it's assignation to this scanIndex and
        next  # go to the next feature...
      }

      # Case B
      # Calculate the RT ratio
      num <- abs(ms2_rt - rt_features[i])
      ratios_rt <- sapply(seq_along(feature_ids)[-i], function(j) {
        denom <- abs(ms2_rt - rt_features[j])
        if (denom == 0) return(Inf)
        num / denom
      })

      # Condition RT
      keep_rt <- which(ratios_rt < 1)  # keep assignment to features with ratio < 1
      if (length(keep_rt)) {
        thr <- quantile(ratios_rt[keep_rt], 0.1)  # top 10% of the smallest rt
        if (any(ratios_rt[keep_rt] <= thr)) keep_feat <- c(keep_feat, i_id)
      }
    }

    # Update the features to keep the final features and the ambiguous ones
    final_feature[[gidx[1]]] <- unique(keep_feat)
    amb_f[[gidx[1]]] <- setdiff(feature_ids, keep_feat)
  }

  # Create a new spectra object with the updated features_id and ambiguous column
  ms2$is_ambiguous <- sp$is_ambiguous
  ms2_res <- ms2[lengths(final_feature) > 0]
  ms2_res$feature_id <- unlist(final_feature, use.names = FALSE)
  ms2_res$ambiguous <- CharacterList(amb_f[lengths(final_feature) > 0])
  ms2_res
}

ms2 <- select_unique_MS2(ms2, ftms)
```

Removing duplicated assignment reduced the data set to `r length(ms2)`
spectra for `r length(unique(ms2$feature_id))` unique features.

### Validation

We next evaluate whether the assignment of single MS2 spectra to multiple
features has been resolved. First we extract the spectra that have been
declared as *ambiguous* (i.e., that were initially assigned to more than one
feature).

```{r}
sp <- spectraData(ms2[ms2$is_ambiguous], c("acquisitionNum", "dataOrigin",
                                           "feature_id", "ambiguous"))
as.data.frame(sp)
```

Next we evaluate if each MS2 spectrum is now assigned to a single feature (even
if it might be/was ambiguous).

```{r}
sp_id <- paste0(ms2$dataOrigin, ms2$acquisitionNum)
max(table(sp_id))
```

Each MS2 spectrum is thus assigned to a single feature.


# Extract all MS2-MSn spectra associated to features

We next define a function to extract the full MSn tree for features.
This function takes an `XcmsExperiment` object as input, extracts the
MS2 spectra using `xcms::featureSpectra()` function, and then retrieves
all MS3 (and subsequently MS4 spectra) for all these MS2 spectra using
the *precScanNum* spectra variable. The `ftms_all_levels` function
returns a combined `Spectra` object containing all MS2, MS3, and MS4
spectra, with each spectrum linked to a feature via a propagated
`feature_id` from the originally matched MS2 spectrum.

```{r}
ftms_all_levels <- function(ftms, ms2) {

  #ms2 <- featureSpectra(ftms, msLevel = 2)
  cat("Number of MS2 spectra found:", length(ms2), "\n")

  res <- lapply(unique(ms2$dataOrigin), function(origin) {
    ms2_subset <- filterDataOrigin(ms2, origin)

    ms3_filtered <- filterDataOrigin(filterMsLevel(spectra(ftms), 3), origin)
    ## To support n:m matches
    m <- findMatches(ms2_subset$acquisitionNum, ms3_filtered$precScanNum)
    cat("Number of MS3 matched to MS2:", length(m), "\n")
    ms3_filtered <- ms3_filtered[to(m)]
    ms3_filtered$feature_id <- ms2_subset$feature_id[from(m)]

    ms4_filtered <- filterDataOrigin(filterMsLevel(spectra(ftms), 4), origin)
    m <- findMatches(ms3_filtered$acquisitionNum, ms4_filtered$precScanNum)
    cat("Number of MS4 matched to MS3:", length(m), "\n")
    ms4_filtered <- ms4_filtered[to(m)]
    ms4_filtered$feature_id <- ms3_filtered$feature_id[from(m)]

    # Combine spectra for this origin
    c(ms2_subset, ms3_filtered, ms4_filtered)
  })
    do.call(c, res)
}

ftms_msn_tree <- ftms_all_levels(ftms,ms2)

table(msLevel(ftms_msn_tree))
```




The number of spectra from different MS levels is shown below:

```{r}
table(msLevel(ftms_msn_tree))
```


# Grouping MSn spectra into fragmentation trees

We next need to group all MSn spectra from one fragmentation tree
together.

The function `assign_msntree_id()` will:

-   Take a `Spectra` object `x` containing MS2–MSn spectra\
-   Assign tree IDs to each spectrum in `x`, based on precursor
    relationships within each sample (`dataOrigin`)\
-   Ensure that MS2 spectra always start a new tree, and that MS3/MS4
    spectra inherit the tree ID from their parent
-   Return a vector of integers, the same length as the number of
    spectra in `x`, where each integer represents the fragmentation tree
    ID the spectrum belongs to

The output of the `assign_msntree_id()` function is then added to the
`Spectra` object as a new spectra variable *MSntreeID*.

```{r}

assign_msntree_id <- function(x) {
  ## Order x by MS level: ensures we first assign an ID to the MS2.
  o <- order(msLevel(x))
  x <- x[o]
  scan_index <- scanIndex(x)
  prec_scan_num <- precScanNum(x)
  ms_level <- msLevel(x)
  data_origin <- dataOrigin(x)

  tree_id <- rep(NA_integer_, length(scan_index))
  global_counter <- 0

  for (origin in unique(data_origin)) {
    idx <- which(data_origin == origin)
    origin_scan_index <- scan_index[idx]
    origin_prec_scan_num <- prec_scan_num[idx]
    origin_ms_level <- ms_level[idx]
    origin_tree_id <- rep(NA_integer_, length(idx))

    scan_idx_map <- setNames(seq_along(origin_scan_index), origin_scan_index)

    for (i in seq_along(idx)) {
      level <- origin_ms_level[i]
      if (level == 2) {
        ## If MS2 hasn't been assigned yet
        if (is.na(origin_tree_id[i])) {
          global_counter <- global_counter + 1
          origin_tree_id[i] <- global_counter
        }
      } else {
        ## Assign ID of the related precursor ID
        parent_scan <- origin_prec_scan_num[i]
        parent_pos <- scan_idx_map[as.character(parent_scan)]
        if (length(parent_pos) == 1 && origin_ms_level[parent_pos] == (level - 1))
          origin_tree_id[i] <- origin_tree_id[parent_pos]
        else
          stop("MS", level, " with ", length(parent_pos), " precursors")
      }
    }
    tree_id[idx] <- origin_tree_id
  }
  ## Return in original order
  return(tree_id[order(o)])
}

msntree_ids <- assign_msntree_id(ftms_msn_tree)

ftms_msn_tree$MSntreeID <- msntree_ids
spectraVariables(ftms_msn_tree)

```

The total number of MSn trees is 3643:

```{r}
length(unique(ftms_msn_tree$MSntreeID))
```

for in total `r length(unique(ftms_msn_tree$feature_id))` features. The
distribution of the lengths of the fragmentation trees is

```{r}
ftms_msn_tree$MSntreeID |>
    table() |>
    quantile()
```

There is apparently (at least) one tree with 4 spectra.
Extracting this particular MSn tree.

```{r}
max_index <- ftms_msn_tree$MSntreeID |>
    table() |>
    as.integer() |>
    which.max()

max_tree <- ftms_msn_tree[ftms_msn_tree$MSntreeID == max_index]
max_tree$rtime
max_tree$msLevel
max_tree$precScanNum
max_tree$scanIndex
max_tree$acquisitionNum
max_tree$dataOrigin
max_tree$feature_id
max_tree$MSntreeID
```

The MSntree with maximum number of spectra corresponds to the tree,
within dataOrigin S1R1, depending on MS2 with ScanIndex 2182 which was
assigned to one FeatureID: FT2705 Within dataOrigin S1R1,
scanIndex 2183 (MS3), 2184 (MS3), and 2185 (MS4) are assigned to a single
feature: FT2705.




```{r}
## we stored the tree ids and the msLevels to a dataframe
df <- spectraData(ftms_msn_tree, c("MSntreeID", "msLevel")) |>
    as.data.frame()

## here we take count the last level of each tree
library(dplyr)
max_level_per_tree <- df %>%
  group_by(MSntreeID) %>%
  summarise(max_level = max(msLevel), .groups = "drop")


table(max_level_per_tree$max_level)

```

The number of trees that go up to MS2 level is : 1808 The number of trees
that go up to MS3 level is : 404 The number of trees that go up to MS4
level is : 1339

Next we select one most representative tree per feature based on the
longest one and if we have many, we select one tree with the highest MS2
precursor intensity

```{r}

unique_tree_feature <- function(db) {
  # Store MS2 indices
  ms2_idx <- which(msLevel(db) == 2)

  # Table of linked MS2 features only
  feat_ms2 <- data.frame(
    idx = ms2_idx,
    feature_id = sapply(db$feature_id[ms2_idx], function(x) x[1]),
    MSntreeID = db$MSntreeID[ms2_idx],
    precursorIntensity = db$precursorIntensity[ms2_idx]
  )

  # for each feature choose the longest tree and the one with the highest
  #precursorIntensity()  in case of tie
  selected_tree_ids <- tapply(seq_len(nrow(feat_ms2)), feat_ms2$feature_id, function(i) {
    #the longest tree
    tbl <- table(feat_ms2$MSntreeID[i])
    FTmax <- as.numeric(names(tbl)[tbl == max(tbl)])
    #the highest precursorIntensity()
    sub_idx <- i[feat_ms2$MSntreeID[i] %in% FTmax]
    sub_idx[which.max(feat_ms2$precursorIntensity[sub_idx])]
  })

  tree_ids_keep <- feat_ms2$MSntreeID[unlist(selected_tree_ids)]
  db[which(db$MSntreeID %in% tree_ids_keep)]
}

ftms_one_tree <- unique_tree_feature(ftms_msn_tree)

msLevel(ftms_one_tree) |> table()
```

after selecting one representative tree per feature we have now:
MS2 : 715 and MS3 : 615 and MS4 : 292


Finally we verify if the number of unique features corresponds to the
number of unique MSntreeID to ensure that we have one tree per feature.

```{r}
length(unique(unlist(ftms_one_tree$feature_id)))
```

```{r}
length(unique(ftms_one_tree$MSntreeID))
```

we have 715 unique feature and unique MSntreeID, That's mean that we
have for each feature one representative tree.

Now we save the resulting ftms spectra object to a local directory.

```{r}
save(ftms_one_tree, file = file.path(pth, "data", "unique_tree_ftms.RData"))
```

# Session information

```{r}
sessionInfo()
```
