---
title: "Untitled"
format: html
server: shiny
---

## Shiny Documents

This Quarto document is made interactive using Shiny. Interactive documents allow readers to modify parameters and see the results immediately. Learn more about Shiny interactive documents at <https://quarto.org/docs/interactive/shiny/>.

## Inputs and Outputs

You can embed Shiny inputs and outputs in your document. Outputs are automatically updated whenever inputs change. This demonstrates how a standard R plot can be made interactive:

```{r}
library(MsExperiment)
library(Spectra)
library(xcms)
library(readxl)
library(pander)
library(pheatmap)
register(SerialParam())
```

```{r}
setwd("C:/Users/amentag/Desktop/Rcodes/RDynLib_ftms")
pth <- getwd()
pth
```
```{r}
sample_data_ftms <- read_xlsx(
    file.path(pth, "data/sample_data/ftms_data_files.xlsx")) |>
    as.data.frame()
MZML_PATH <- file.path(pth,"data","mzML_cent1")
ftms <- readMsExperiment(file.path(MZML_PATH, sample_data_ftms$data_file),
                         sampleData = sample_data_ftms)
ftms
```


```{r}
s <- spectra(ftms)
```


```{r}
spectraData(s)[ spectraData(s)$scanIndex == 2742 &
                  spectraData(s)$dataOrigin == "C:\\Users\\amentag\\Desktop\\Rcodes\\RDynLib_ftms\\data\\mzML_cent1\\S9R5.mzML",]

```



```{r}
peaksData(s)[spectraData(s)$scanIndex == 2742 &
                  spectraData(s)$dataOrigin == "C:\\Users\\amentag\\Desktop\\Rcodes\\RDynLib_ftms\\data\\mzML_cent1\\S9R5.mzML",][[1,]]
```
```{r}
library(RColorBrewer)
# colors
col_sample_ftms_id <- brewer.pal(9, "Set1")[c(1, 2)]
names(col_sample_ftms_id) <- unique(sampleData(ftms)$sample_id)
# Assigne colors for each single name in the sample_id
names(col_sample_ftms_id) <- c("S1", "S9")
col_sample_ftms <- col_sample_ftms_id[sampleData(ftms)$sample_id]
```



```{r}
mzr <- c(1083.27, 1083.29)

a_kn <- chromatogram(ftms, mz = mzr, aggregationFun = "max")
plot(a_kn, col = paste0(col_sample_ftms, 80))
grid()
```





```{r}
library(Spectra)

mzr <- c(463.01, 463.15)

tmp <- filterMz(ftms, mzr) |>
        filterRt(rt = c(2760, 2766))|>
     filterMsLevel(1L)
plot(tmp[1])



```




```{r}
library(Spectra)
library(MsExperiment)
library(readxl)


MZXML_PATH <- file.path(pth, "data", "mzXML")
MZML_PATH  <- file.path(pth, "data", "mzML_cent1")

if (!dir.exists(MZML_PATH))
    dir.create(MZML_PATH, recursive = TRUE)

#Process and convert each mzXML file 
fls <- dir(MZXML_PATH, full.names = TRUE)

for (f in fls) {
    message("Processing: ", f)
  
    a <- Spectra(f)
    a <- setBackend(a, MsBackendMemory())
    
    ms1 <- filterMsLevel(a, 1L)
    msx <- filterMsLevel(a, msLevel(a) > 1L)
    
    fn <- file.path(MZML_PATH, sub("mzXML$", "mzML", basename(f)))
    #Centroid just MS1 spectra 
    ms1_cent <- Spectra::pickPeaks(
        ms1,
        halfWindowSize = 10L,
        k = 2L,
        threshold = 0.33
    )
    ms1_cent <- applyProcessing(ms1_cent)
    # Merge and sort by retention time
    merged <- c(ms1_cent, msx)
    merged <- merged[order(rtime(merged))]
    
    #Export to mzML
    
    export(
        merged,
        backend = MsBackendMzR(),
        file = fn,
        format = "mzML",
        copy = TRUE
    )
}

b <- Spectra(fn)
all(centroided(b))
all.equal(a$acquisitionNum, b$acquisitionNum)
all.equal(a$precScanNum, b$precScanNum)
all.equal(a$scanIndex, b$scanIndex)

```




```{r}
sample_data_ftms <- read_xlsx(file.path(pth,"data","sample_data","ftms_data_files.xlsx")) |>
    as.data.frame()

ftms <- readMsExperiment(file.path(MZML_PATH, sample_data_ftms$data_file),
                         sampleData = sample_data_ftms)
ftms
```


