---
title: "Untitled"
format: html
server: shiny
---

## Shiny Documents

This Quarto document is made interactive using Shiny. Interactive documents allow readers to modify parameters and see the results immediately. Learn more about Shiny interactive documents at <https://quarto.org/docs/interactive/shiny/>.

## Inputs and Outputs

You can embed Shiny inputs and outputs in your document. Outputs are automatically updated whenever inputs change. This demonstrates how a standard R plot can be made interactive:

```{r}
library(dplyr)
library(tidyr)
library(xcms)
library(MsExperiment)
library(Spectra)
library(readxl)
library(IRanges)
library(S4Vectors)
register(SerialParam())
setwd("C:/Users/amentag/Desktop/Rcodes/RDynLib_ftms")
pth <- getwd()
pth
```

```{r}
load(file.path(pth, "data", "ftms.RData"))
class(ftms)
```

```{r}
ftms
library(Spectra)

sp <- spectra(ftms)

sp_target <- spectraData(sp)[ spectraData(sp)$scanIndex == 2891 &
                  spectraData(sp)$dataOrigin == "C:\\Users\\amentag\\Desktop\\Rcodes\\RDynLib_ftms\\data\\mzML_cent1\\S9R4.mzML",] 
sp_target
```

```{r}
peaksData(sp)[spectraData(sp)$scanIndex == 2772 &
                  spectraData(sp)$dataOrigin == "C:\\Users\\amentag\\Desktop\\Rcodes\\RDynLib_ftms\\data\\mzML_cent1\\S1R4.mzML",][[1,]]
```

```{r}
ms2 <- featureSpectra(dropFilledChromPeaks(ftms), msLevel = 2, skipFilled = TRUE)

msLevel(ms2) |>
  table()
```

```{r}
ms2_S1R4 <- spectraData(ms2)[ spectraData(ms2)$scanIndex == 4883 &
                  spectraData(ms2)$dataOrigin == "C:\\Users\\amentag\\Desktop\\Rcodes\\RDynLib_ftms\\data\\mzML_cent1\\S1R5.mzML",]
ms2_S1R4
```

```{r}
peaksData(ms2)[spectraData(ms2)$scanIndex == 4883 &
                  spectraData(ms2)$dataOrigin == "C:\\Users\\amentag\\Desktop\\Rcodes\\RDynLib_ftms\\data\\mzML_cent1\\S1R5.mzML",][[1,]]

```

```{r}
FT5341 <- spectraData(ms2)[ spectraData(ms2)$feature_id == "FT5733",]

FT5341
```

```{r}
peaksData(ms2)[spectraData(ms2)$scanIndex == 4824 &
                  spectraData(ms2)$feature_id == "FT5733",][[1,]]
```

```{r}
peaksData(ms2)[spectraData(ms2)$scanIndex == 2713 &
                  spectraData(ms2)$feature_id == "FT5341",][[1,]]
```

```{r}
peaksData(ms2)[spectraData(ms2)$scanIndex == 2654 &
                  spectraData(ms2)$feature_id == "FT5341",][[1,]]
```

```{r}
peaksData(ms2)[spectraData(ms2)$scanIndex == 2701 &
                  spectraData(ms2)$feature_id == "FT5341",][[1,]]
```

```{r}
peaksData(ms2)[spectraData(ms2)$scanIndex == 2683 &
                  spectraData(ms2)$feature_id == "FT5229",][[1,]]
```

```{r}
load("C:\\Users\\amentag\\Desktop\\Rcodes\\RDynLib_ftms\\data\\ftms_one_tree_test.RData")
```

```{r}
tst <- spectraData(ftms_one_tree)[spectraData(ftms_one_tree)$feature_id == "FT5733",]
tst
```

```{r}
peaksData(ftms_one_tree)[spectraData(ftms_one_tree)$scanIndex == 2683 &
                  spectraData(ftms_one_tree)$feature_id == "FT5229",][[1,]]
```
